---
title: "Project 1, Part 5"
output: 
  rmdformats::robobook:
    css: "homework.css"
    highlight: pygments
link-citations: yes
---

**Due:** Monday, April 18, 12:00pm (noon) on Canvas

**Submission:** A template file is provided: [project_1_part_5_template.Rmd](https://sta112-s22.github.io/projects/project_1_part_5_template.Rmd)

Once you have completed the report, submit your 
knitted HTML file to Canvas.

# Overview

So far, you have used exploratory data analysis and linear regression with a single predictor to answer your research questions. The next step will be to add more variables to your model(s), to answer more in-depth questions. In Part 4, you described potential approaches for answering these research questions. In this part of the project, you will do the analysis and report your results. You do not need to follow your plan from Part 4 exactly -- when we actually do data analysis, our plans sometimes change!

This project must be completed independently. See "Collaboration" below for more details.

# Data

The [College Scorecard](https://collegescorecard.ed.gov/), compiled by the US Department of Education, is meant to be a tool for college applicants to compare different institutions. In this project, you will work with data from the scorecard to advise a high school senior on choosing a college. (We will use data that are several years old, so may be slightly out of date, but we will hope that the same general trends are still true today).

## Variables

The dataset provided for this project contains the following columns:

* `INSTNM`: school name
* `STABBR`: state in which the school is located
* `CONTROL`: whether the school is public or private
* `SATVRMID`: midpoint of SAT critical reading scores of students attending the school
* `SATMTMID`: midpoint of SAT math scores
* `SATWRMID`: midpoint of SAT writing scores
* `ACTCMMID`: midpoint of the ACT cumulative scores
* `ACTENMID`: midpoint of ACT English scores
* `ACTMTMID`: midpoint of ACT math scores
* `ACTWRMID`: midpoint of ACT writing scores
* `UGDS`: number of undergraduate students at the school
* `NPT4`: average cost to attend the school
* `PCTFLOAN`: fraction of undergraduates receiving a federal student loan
* `MD_EARN_WNE_P10`: median salary of students 10 years after graduation
* `GRAD_DEBT_MDN_SUPP`: median debt of graduated students
* `RPY_3YR_RT_SUPP`: proportion of students who are actively repaying their loans 3 years after graduation
* `RPY_3YR_70`: whether the proportion of students actively repaying their loans 3 years after graduation is > 70% (1 = yes, 0 = no)

## Downloading the data

The data for this project is stored in a CSV file, and can be imported into R with the `read_csv` function in the `readr` package. The following code imports the data and saves it as a data frame called `scorecard`:

```r
scorecard <- read_csv("https://sta112-s22.github.io/projects/scorecard.csv")
```

This code is included in the setup chunk of the R Markdown template for the project.

# Research questions

For this project, suppose you have been approached by a high school senior applying to college. They are looking through the college scorecard data, and they have some questions for you about the value of different schools. 

1. Do students who graduate from more expensive schools earn more money after graduation, after accounting for student debt, school type (public or private), and the test scores of enrolled students?

2. Is the relationship between school cost and graduate earnings the same for public and private schools, after accounting for student debt and the test scores of enrolled students?


# Content

As in Part 3, your written report for Part 5 will contain the following sections: Introduction, Data, Modeling, and Discussion. 

* The Introduction can be similar to your introduction from Part 3. The only difference is that you will now be working with multiple predictors in your regression models.
* The Data section will be similar to Part 3. The only difference may be some plots for one or two of the new variables you are including in your multiple regression model(s). Note that I ask you to consider an additional variable in the univariate EDA section.
* The Modeling section will be new, because you will be fitting regression model(s) with multiple predictors. Accordingly, you will have different models and diagnostics than in Part 3. However, the format will be similar to the format in Part 3.
* The Discussion can be similar to your discussion from Part 3, but should be updated to discuss your new models and results.

You can base your analysis in the Modeling section on the data analysis plan you submitted in Part 4. However, you do not need to follow this plan exactly, and you can make changes based on exploratory data analysis and assumption diagnostics. For example, you should transform your data if needed.

## Introduction

Write 1-2 paragraphs in which you:

* Motivate the research questions
* Describe where the data came from
* Provide background on the dataset
* Finish by summarizing what you will do in the report

## Data

In the data section, you have three subsections: Data Summary, Univariate EDA, and Multivariate EDA.

### Data Summary

Write a paragraph that describes details of the data:

* What does a row in the data represent? 
* How many rows and columns are there? 
* What information do the variables record? 
* Are there any missing data?

(Make sure to write your summary as a paragraph, not a list, in your report).

### Univariate EDA

Choose **4** variables that are relevant to the research questions. For each variable: make a plot, calculate any appropriate summary statistics, and discuss the distribution in 2--3 sentences. Make sure to add a title and good labels to each plot. Plots and summary statistics depend on the type of variable.

### Multivariate EDA

Using the variables from Univariate EDA, create one or more plots that allow you to address the research questions. You may be able to address both questions with a single plot, or you may need to make a separate plot for each question. Calculate any appropriate summary statistics. Use the plot(s) and summary statistics to address the research questions in a 1--2 paragraphs.

## Modeling

In the modeling section, you have three subsections: Model, Assumption Diagnostics, and Inference.
  
### Model

First, we need to describe the regression model(s) and give the equation of the fitted model(s). For each research question, write a paragraph in which you:

* Give the population regression model. This should be written using correct notation (remember that $\varepsilon$ is included in the population model, and there are no hats on any variables or parameters). Make sure to use informative variable names (e.g., "Cost" instead of "X"). See "Writing math in R Markdown" below.
* Give the equation of the fitted regression model, using correct notation and informative variable names.
* Interpret one or more relevant coefficients of the fitted regression model, and explain what they say about the research question.

### Assumption Diagnostics

Next, we want to assess whether the regression assumptions are reasonable for our model(s). For each model, write 2 - 3 paragraphs in which you:

* Create a residual plot, QQ plot, and studentized residual plot for your fitted model. Make sure to include informative axis labels and captions (with figure numbers) for your figures.
* Use your residual plot and QQ plot to discuss whether the shape, constant variance, and normality assumptions are reasonable. Refer to your figures by their numbers in the text.
* Use your studentized residual plot to identify any potential outliers
* Discuss the independence and randomness assumptions for the college scorecard data
* Assess multicollinearity for your predictors. If multicollinearity is an issue, you should make any appropriate changes to your model.


### Inference

Finally, we want to use our fitted regression model(s) to address the research questions. For each research question, write 2 - 3 paragraphs in which you:

* Translate the research question into null and alternative hypotheses about one or more model parameters. Write these hypotheses using correct notation (see "Writing math in R Markdown" below).
* Calculate a test statistics and p-value for your hypotheses.
* Use the p-value to make a conclusion about the research question.
  * *Remember that in this class, we don't reject/fail to reject/accept p-values, nor do we compare them to cutoffs like 0.01, 0.05, etc.*
  
## Discussion

In this section, you are providing a conclusion. Write 1–2 paragraphs which summarize what you learned from your analysis, and how it addresses the original research questions. Also discuss any limitations to your analysis, and what you other relationships you might explore in future.
  
  
# Appearance and style

### Writing

The report should be written like an article or research paper: in full sentences and paragraphs, with headings for each section. You should not write your report with question numbers or as a list of bullet points. Scientific articles are generally written in third person, though "we" can also be acceptable ("we can see from Figure 1...") in some disciplines.

### Code

As with previous assignments, you will write the document in R Markdown. However, in full reports, the only output that should be visible from code chunks are figures and tables. If a code chunk does not produce a figure or table, you can hide it from the knitted document with `include=F`:

````
```{r, include=F}`r ''`

```
````

If a code chunk produces a figure or table, only the figure or table should be visible in the knitted document. You can hide the chunk but display the output with `echo=F, message=F, warning=F`:

````
```{r, echo=F, message=F, warning=F}`r ''`

```
````

### Figures

Figures should have labeled axes, and should be clear and easy to read. Figures should also be captioned and numbered; to caption a figure, use `fig.cap = "..."` in the chunk options. For example (scroll to the right on the code to see it all),

````
```{r, echo=F, message=F, warning=F, fig.cap="Figure 1: Bill depth vs. bill length for penguins near Palmer Station, Antarctica."}`r ''`
penguins %>%
  ggplot(aes(x = bill_length_mm, 
             y = bill_depth_mm)) +
  geom_point() +
  labs(x = "Bill length (mm)",
       y = "Bill depth (mm)") +
  theme_bw()
```
````

is displayed as

```{r include=F}
library(tidyverse)
library(palmerpenguins)
```

```{r, echo=F, message=F, warning=F, fig.cap="Figure 1: Bill depth vs. bill length for penguins near Palmer Station, Antarctica."}
penguins %>%
  ggplot(aes(x = bill_length_mm, 
             y = bill_depth_mm)) +
  geom_point() +
  labs(x = "Bill length (mm)",
       y = "Bill depth (mm)") +
  theme_bw()
```

Captions should provide enough information to understand what is being plotted, but interpretation can be left to the main text. Refer to figures by their number in the text. Make sure that any figures you include are discussed in the text.

### Writing math in R Markdown

Doing things like writing down models often involve parameters. Parameters in statistics are generally represented as Greek letters. How can we write Greek letters and other mathematical notation in our Markdown file so they show up when we knit?

If you want to write mathematical notation, we need to tell Markdown, “Hey, we’re going to make a math symbol!” To do that, you use dollar signs. For instance, to make $\widehat{\beta}_1$, you simply put `$\widehathat{\beta}_1$` into the white space (not a chunk) in your Markdown.

Here are some examples of writing math, which you can adapt:

| Math | Code |
| --- | --- |
| $Y = \beta_0 + \beta_1 X + \varepsilon$ | `$Y = \beta_0 + \beta_1 X + \varepsilon$` |
| $\widehat{Y} = -2.2 + 3.7 X$ | `$\widehat{Y} = -2.2 + 3.7 X$` |
| $H_0: \beta_1 = 0$ | `$H_0: \beta_1 = 0$` |
| $H_A: \beta_1 \neq 0$ | `$H_0: \beta_1 \neq 0$` |
| $H_A: \beta_1 > 0$ | `$H_0: \beta_1 > 0$` |
| $H_A: \beta_1 < 0$ | `$H_0: \beta_1 < 0$` |

# Collaboration

One purpose of the projects is to capture your knowledge and understanding of statistics and data analysis. Unlike homework and lab assignments, this project must be completed on your own, without collaborating with other students. 

You may:

* ask the instructor for help
* use resources from the class notes and internet (Wikipedia, StackExchange, etc.)

You may not:

* work with another student on the project
* ask a TA or tutor (e.g., at the MSC) for help