---
title: "Project 2"
output: 
  rmdformats::robobook:
    css: "homework.css"
    highlight: pygments
link-citations: yes
---

**Due:** Thursday, May 5, 12:00pm (noon) on Canvas

**Submission:** A template file is provided: [project_2_template.Rmd](https://sta112-s22.github.io/projects/project_2_template.Rmd)

Once you have completed the report, submit your 
knitted HTML file to Canvas.

# Overview

In this project, you will use exploratory data analysis and linear regression to investigate use of an online discussion forum. You will then write a data analysis report that communicates the results of your analysis and draws a conclusion about the research question.

This project must be completed independently. See "Collaboration" below for more details.

# Data

In online discussion forums, a wide variety of information is collected about each post. This information can help moderators monitor the site, by helping them predict which posts will receive the most attention, and which may need to be moderated or deleted due to offensive or off-topic comments.

The data here comes from a website called ScienceForums.Net (SFN), which has been open since 2002 and hosts conversations on a range of topics from biological and physical science to religion and philosophy. Each row in the data represents one 'thread', which is comprised of a series of posts stemming from an initial post. For each thread, we have some information that SFN collects such as the number of views and the number of authors. The threads present in the data are a random sample of threads from 2002-2014, with the data collected in 2014. SFN moderators are interested in using this data to determine which threads warrant the most attention.

## Variables

The dataset provided for this project contains the following columns:

* `Age`: the age of the thread (in days) when the data was collected in 2014, measured from the first post in the thread
* `State`: sometimes moderators close threads if they are inappropriate. closed indicates the thread has been closed, otherwise State is open
* `Posts`: the number of posts in the thread
* `Views`: the total number of views of the thread
* `Duration`: 	the number of days between the first and last posts in the thread
* `Authors`: the number of distinct authors posting in the thread
* `AuthorExperience`: the number of days the author of the first post in the thread had been registered on SFN when the thread began (0 indicates they registered that day)
* `DeletedPosts`: the number of posts in the thread that have been deleted by a moderator
* `Forum`: the forum in which the thread was posted (e.g., Science)
* `AuthorBanned`: whether the original author of the thread is currently banned from posting on SFN (at the time of data collection, not when the thread was first posted)

## Downloading the data

The data for this project is stored in a CSV file, and can be imported into R with the `read_csv` function in the `readr` package. The following code imports the data and saves it as a data frame called `sfn`:

```r
sfn <- read_csv("https://sta112-f21.github.io/projects/sfn.csv")
```

# Research Question

For this project, suppose you have been approached by moderators at SFN. They give you the data, and ask the following question:

What is the relationship between views and posts? Does it vary by subforum? In answering this question, the moderators want you to make sure that you account for other variables which may be associated with views and posts; for example, older threads may have more views just because they have been around longer.


# Content

As in Project 1, your written report for Project 2 will contain the following sections: Introduction, Data, Modeling, and Discussion. 

## Introduction

Write 1-2 paragraphs in which you:

* Motivate the research question
* Describe where the data came from
* Provide background on the dataset
* Finish by summarizing what you will do in the report

## Data

In the data section, you have three subsections: Data Summary, Univariate EDA, and Multivariate EDA.

### Data Summary

Write a paragraph that describes details of the data:

* What does a row in the data represent? 
* How many rows and columns are there? 
* What information do the variables record? 
* Are there any missing data?

(Make sure to write your summary as a paragraph, not a list, in your report).

### Univariate EDA

Choose **4** variables that are relevant to the research questions. For each variable: make a plot, calculate any appropriate summary statistics, and discuss the distribution in 2--3 sentences. Make sure to add a title and good labels to each plot. Plots and summary statistics depend on the type of variable.

### Multivariate EDA

Using the variables from Univariate EDA, create one or more plots that allow you to address the research question. Calculate any appropriate summary statistics. Use the plot(s) and summary statistics to address the research question in a 1--2 paragraphs.

## Modeling

In the modeling section, you have three subsections: Model, Assumption Diagnostics, and Inference.
  
### Model

First, we need to describe the regression model(s) and give the equation of the fitted model(s). Write a paragraph in which you:

* Give the population regression model. This should be written using correct notation (remember that $\varepsilon$ is included in the population model, and there are no hats on any variables or parameters). Make sure to use informative variable names. See "Writing math in R Markdown" below.
* Give the equation of the fitted regression model, using correct notation and informative variable names.
* Interpret one or more relevant coefficients of the fitted regression model, and explain what they say about the research question.

### Assumption Diagnostics

Next, we want to assess whether the regression assumptions are reasonable for our model(s). For each model, write 2 - 3 paragraphs in which you:

* Create a residual plot, QQ plot, and studentized residual plot for your fitted model. Make sure to include informative axis labels and captions (with figure numbers) for your figures.
* Use your residual plot and QQ plot to discuss whether the shape, constant variance, and normality assumptions are reasonable. Refer to your figures by their numbers in the text.
* Use your studentized residual plot to identify any potential outliers
* Discuss the independence and randomness assumptions for the college scorecard data
* Assess multicollinearity for your predictors. If multicollinearity is an issue, you should make any appropriate changes to your model.


### Inference

Finally, we want to use our fitted regression model(s) to address the research question. Write 2 - 3 paragraphs in which you:

* Translate the research question into null and alternative hypotheses about one or more model parameters. Write these hypotheses using correct notation (see "Writing math in R Markdown" below).
* Calculate a test statistics and p-value for your hypotheses.
* Use the p-value to make a conclusion about the research question.
  * *Remember that in this class, we don't reject/fail to reject/accept p-values, nor do we compare them to cutoffs like 0.01, 0.05, etc.*
  
## Discussion

In this section, you are providing a conclusion. Write 1–2 paragraphs which summarize what you learned from your analysis, and how it addresses the original research question. Also discuss any limitations to your analysis, and what other relationships you might explore in future.
  
  
# Appearance and style

### Writing

The report should be written like an article or research paper: in full sentences and paragraphs, with headings for each section. You should not write your report with question numbers or as a list of bullet points. Scientific articles are generally written in third person, though "we" can also be acceptable ("we can see from Figure 1...") in some disciplines.

### Code

As with previous assignments, you will write the document in R Markdown. However, in full reports, the only output that should be visible from code chunks are figures and tables. If a code chunk does not produce a figure or table, you can hide it from the knitted document with `include=F`:

````
```{r, include=F}`r ''`

```
````

If a code chunk produces a figure or table, only the figure or table should be visible in the knitted document. You can hide the chunk but display the output with `echo=F, message=F, warning=F`:

````
```{r, echo=F, message=F, warning=F}`r ''`

```
````

### Figures

Figures should have labeled axes, and should be clear and easy to read. Figures should also be captioned and numbered; to caption a figure, use `fig.cap = "..."` in the chunk options. For example (scroll to the right on the code to see it all),

````
```{r, echo=F, message=F, warning=F, fig.cap="Figure 1: Bill depth vs. bill length for penguins near Palmer Station, Antarctica."}`r ''`
penguins %>%
  ggplot(aes(x = bill_length_mm, 
             y = bill_depth_mm)) +
  geom_point() +
  labs(x = "Bill length (mm)",
       y = "Bill depth (mm)") +
  theme_bw()
```
````

is displayed as

```{r include=F}
library(tidyverse)
library(palmerpenguins)
```

```{r, echo=F, message=F, warning=F, fig.cap="Figure 1: Bill depth vs. bill length for penguins near Palmer Station, Antarctica."}
penguins %>%
  ggplot(aes(x = bill_length_mm, 
             y = bill_depth_mm)) +
  geom_point() +
  labs(x = "Bill length (mm)",
       y = "Bill depth (mm)") +
  theme_bw()
```

Captions should provide enough information to understand what is being plotted, but interpretation can be left to the main text. Refer to figures by their number in the text. Make sure that any figures you include are discussed in the text.

### Writing math in R Markdown

Doing things like writing down models often involve parameters. Parameters in statistics are generally represented as Greek letters. How can we write Greek letters and other mathematical notation in our Markdown file so they show up when we knit?

If you want to write mathematical notation, we need to tell Markdown, “Hey, we’re going to make a math symbol!” To do that, you use dollar signs. For instance, to make $\widehat{\beta}_1$, you simply put `$\widehathat{\beta}_1$` into the white space (not a chunk) in your Markdown.

Here are some examples of writing math, which you can adapt:

| Math | Code |
| --- | --- |
| $Y = \beta_0 + \beta_1 X + \varepsilon$ | `$Y = \beta_0 + \beta_1 X + \varepsilon$` |
| $\widehat{Y} = -2.2 + 3.7 X$ | `$\widehat{Y} = -2.2 + 3.7 X$` |
| $H_0: \beta_1 = 0$ | `$H_0: \beta_1 = 0$` |
| $H_A: \beta_1 \neq 0$ | `$H_0: \beta_1 \neq 0$` |
| $H_A: \beta_1 > 0$ | `$H_0: \beta_1 > 0$` |
| $H_A: \beta_1 < 0$ | `$H_0: \beta_1 < 0$` |

# Collaboration

One purpose of the projects is to capture your knowledge and understanding of statistics and data analysis. Unlike homework and lab assignments, this project must be completed on your own, without collaborating with other students. 

You may:

* ask the instructor for help
* use resources from the class notes and internet (Wikipedia, StackExchange, etc.)

You may not:

* work with another student on the project
* ask a TA or tutor (e.g., at the MSC) for help