---
title: "HW 1 Solutions"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(nycflights13)
library(tidyverse)
```

Start by grading problem 1 for all students, then problem 2 for all students, etc.

Don't put in their final score yet. I need to grade their plagiarism tutorials.

If they get an answer wrong, or their answer could be improved, leave them a short comment explaining what they did wrong.

## Question 1 (5 pts)

**Question:** Summarize the `flights` data. Describe:

* Who (or which organization) originally collected the data?
* What does a row in the data represent?
* How many rows and columns are in the data?
* What information do the variables record?
* Are there any missing data?

**Solution:**

```{r, eval=F}
# getting dimensions of the data
glimpse(flights)

# You can view documentation with ?flights in the console

# getting the dimension after removing NAs
flights %>%
  drop_na() %>%
  glimpse()

# 336776 - 327346
```

The `flights` dataset contains information on 336776 different flights from three New York airports (JFK, LGA, and EWR) in 2013. Each row represents a flight, and there are 19 columns capturing information about departure and landing times, delays, the origin and destination, and the airline. The data was collected by the Bureau of Transportation Statistics, part of the Research and Innovative Technology Administration (RITA), and is provided in the `nycflights13` data. 9430 rows in the data contain missing information.

**Grading**:

When grading this question, please comment on any parts which are missing or incorrect. I ask for 5 different things in the bulleted list above. Take off 1 point for each answer that is missing / incorrect.

## Question 2 (3 pts)

**Question:** Create a histogram for the distribution of departure delay, and describe the distribution.

**Solution:** 

```{r}
flights %>%
  ggplot(aes(x = dep_delay)) +
  geom_histogram() +
  labs(x = "Departure delay (minutes)",
       title = "Histogram for departure delay") +
  theme_bw()
```

The distribution of departure delay is unimodal and highly right skewed, with a peak around 0 minutes. Almost all flights in the data have a departure delay below 200 minutes. These may be outliers.

**Grading:** 

To get full credit, they only need to make a histogram and describe some aspect of the distribution. In future, they will be asked to describe shape, center, spread, and comment on any potential outliers. They do not need to change the theme, axis labels, or title of the histogram. They do not need to count the number of outliers.

They may use `drop_na` before creating the histogram. Whether or not they use `drop_na` does not affect their score.

Take off 1 point for an incorrect plot. Take off 1 point for a missing/incorrect interpretation.

## Question 3 (3 pts)

**Question:** Create a bar chart showing the number of flights departing from each of the three New York airports (EWR, LGA, and JFK). Which airport had the most flights?

**Solution:**

```{r}
flights %>%
  ggplot(aes(x = origin)) +
  geom_bar() +
  labs(x = "Origin airport",
       title = "Distribution of flights by NYC airport") +
  theme_bw()
```

EWR had the most departing flights.

**Grading:** 

To get full credit, they only need to make a bar chart and report the airport with the most departing flights. They do not need to change the theme, axis labels, or title of the bar chart.

They may use `drop_na` before creating the bar chart. Whether or not they use `drop_na` does not affect their score.

Take off 1 point for an incorrect plot. Take off 1 point for a missing/incorrect interpretation.

## Question 4 (6 pts)

**Question:** Create a plot showing the distribution of departure delay for each airport. For each airport, also report the mean departure delay and the standard deviation of departure delays. How does the distribution of departure delay differ across the three New York airports (EWR, LGA, and JFK)?

**Solution:** Acceptable answers for the plot are either faceted histograms or grouped box plots (see below).

```{r}
# Option 1: faceted histograms
flights %>%
  ggplot(aes(x = dep_delay)) +
  geom_histogram() +
  facet_wrap(~origin) +
  labs(x = "Departure delay (minutes)",
       title = "Departure delay by origin airport") +
  theme_bw()

# Option 2: boxplots
flights %>%
  ggplot(aes(x = origin, y = dep_delay)) +
  geom_boxplot() +
  labs(x = "Origin airport",
       y = "Departure delay (minutes)",
       title = "Departure delay by origin airport")

# calculating mean and sd
flights %>%
  drop_na() %>%
  group_by(origin) %>%
  summarize(mean_delay = mean(dep_delay),
            sd_delay = sd(dep_delay))
```

The distributions of departure delay look similar across airports, with all three distributions unimodal and strongly right skewed. The three distributions have similar means and standard deviations.

**Grading:** 2 pts for the plot, 2 points for the summary statistics, and 2 points for the interpretation. 

* I didn't ask for them to discuss specific features of the plot in their interpretation, so any reasonable interpretation gets full credit. But take off 1 point for incorrect interpretation
* Take off 1 point if they don't remove NAs before calculating the summary statistics. Whether or not they use `drop_na` does not affect their score for the plot. 
* Take off 1 - 2 points if the plot doesn't actually visualize the distribution of departure delay for each airport



## Question 5 (8 pts)

**Question:** Below is a scatterplot showing arrival delay vs. departure delay. I have:

* colored and shaped each flight by the New York airport it departed from
* changed the axis labels
* added a title
* changed the legend label (to "Origin airport")
* changed the plot theme (e.g., notice the background is white, not gray)
* changed the opacity of the points
* changed the colors used to represent each airport

Recreate this plot as closely as possible.

**Solution:**

```{r}
flights %>%
  drop_na() %>%
  ggplot(aes(x = dep_delay, 
             y = arr_delay, 
             color = origin,
             shape = origin)) +
  geom_point(alpha = 0.7) +
  labs(x = "Departure delay (minutes)", 
       y = "Arrival delay (minutes)",
       color = "Origin airport",
       shape = "Origin airport",
       title = "Arrival delay vs. departure delay in NYC airports") +
  theme_bw() +
  scale_color_manual(values = c("#999999", "#E69F00", "#56B4E9"))
```

**Grading:**

There are 8 pieces in this plot: the basic scatterplot of arrival delay vs. departure delay, and then the 7 components that I changed. We'll grade based on the number of complete pieces. Here's what they need for each piece to be complete:

* basic scatterplot: correctly displays arrival delay on y axis and departure delay on x axis
* color and shape the points: correctly colors and shapes by origin
* changing axis labels: both x and y axis labels are close to mine
* adding a title: title is close to mine
* changing legend label: changed to "Origin airport" (capitalization not important), and just one legend (not separate legends for color and shape)
* changed the plot theme: any theme other than the basic ggplot theme is ok
* changed the opacity: any opacity other than 1 is ok
* changed the colors: any colors that are somewhat similar to mine are ok

Scoring: Take off 1 point for each incomplete piece

# Part II -- ZSR tutorial (5 pts)

I'll grade this part separately