---
title: "Homework 4 Solutions"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(Stat2Data)
data("MetroCommutes")
```

### Question 1 (3 pts)

```{r}
MetroCommutes %>%
  ggplot(aes(x = City, y = Time)) +
  geom_boxplot() +
  theme_bw()
```

The distribution of commute times for each city looks somewhat right-skewed, with a few points in the tails which may be potential outliers. Houston and Minneapolis appear to have slightly shorter commute times than Boston and Washington.

**Grading:** 1 pt for plot, 1 pt for describing boxplots, 1 pt for comparing between cities.

### Question 2 (1 pt)

```{r}
MetroCommutes %>%
  count(City)
```

We have 500 observations in each city.

**Grading:** They can get the numbers different ways. They just need to report 500 in each city.


### Question 3 (3 pts)

$$\text{Time} = \beta_0 + \beta_1 \text{IsHouston} + \beta_2 \text{IsMinnneapolis} + \beta_3 \text{IsWashington} + \varepsilon$$
There are 4 parameters ( $\beta_0$, $\beta_1$, $\beta_2$, and $\beta_3$ ).

**Grading:** 2 pts for population model, 1 pt for number of parameters

* Lose 1 pt if they don't make Boston the base category
* Lose 1 pt if they don't include $\varepsilon$
* Lose 1 pt if they don't use indicator variables (e.g. if they do $\beta_2 \text{City}$ )

### Question 4 (4 pts)

```{r}
city_lm <- lm(Time ~ City, data = MetroCommutes)
summary(city_lm)
```

$$\widehat{\text{Time}} = 25.746 - 0.648 \ \text{IsHouston} - 3.004 \ \text{IsMinnneapolis} + 5.944 \ \text{IsWashington}$$
Average commute times in each city:

* Boston: 25.746
* Houston: 25.746 - 0.648 = 25.098
* Minneapolis: 25.746 - 3.004 = 22.742
* Washington: 25.746 + 5.944 = 31.69

**Grading:** 2 pts for equation of fitted model, 2 pts for mean in each city

* Lose 1 pt if they don't put hat on the response, and/or if they add hats to the predictors, and/or if they keep $\varepsilon$ in the equation
* Lose 1 pt if they think the estimated coefficients on the cities are the means (e.g., they report that -0.648 is the mean for Houston)


### Question 5 (4 pts)

```{r}
MetroCommutes %>%
  mutate(resids = residuals(city_lm)) %>%
  ggplot(aes(x = City, y = resids)) +
  geom_boxplot() +
  theme_bw() +
  labs(x = "City", 
       y = "Residuals")

MetroCommutes %>%
  mutate(resids = residuals(city_lm)) %>%
  group_by(City) %>%
  summarize(residual_std_dev = sd(resids))
```

Looking at the boxplots, Houston and Minneapolis seem to have slightly less variability in the residuals. Looking at the estimated standard deviations, Washington has a slightly larger standard deviation than the others. But the ratio of the largest to smallest is 21.2/16.9 < 2, so we're not concerned about violations of the constant variance assumption.

**Grading:** 1 pt for plot, 1 pt for standard deviations, 1 pt for calculating ratio, 1 pt for conclusion

### Question 6 (2 pts)

```{r}
MetroCommutes %>%
  mutate(resids = residuals(city_lm)) %>%
  ggplot(aes(sample = resids)) +
  geom_qq() +
  geom_qq_line() +
  theme_bw() +
  labs(x = "Theoretical normal quantiles", 
       y = "Observed residual quantiles")
```

The normality assumption appears to be violated here -- there is some deviation from normality in the tails, particularly the right tail.

**Grading:** 1 pt for plot, 1 pt for interpretation

### Question 7 (4 pts)

```{r}
city_lm <- lm(sqrt(Time) ~ City, data = MetroCommutes)
summary(city_lm)
```

$$\widehat{\sqrt{\text{Time}}} = 4.759 - 0.029 \ \text{IsHouston} - 0.270 \ \text{IsMinnneapolis} + 0.562 \ \text{IsWashington}$$

```{r}
MetroCommutes %>%
  mutate(resids = residuals(city_lm)) %>%
  ggplot(aes(sample = resids)) +
  geom_qq() +
  geom_qq_line() +
  theme_bw() +
  labs(x = "Theoretical normal quantiles", 
       y = "Observed residual quantiles")
```

The normality assumption looks much better now. It still isn't perfect (there is a little deviation in the right tail), but it looks pretty good.

**Grading:** 2 pts for equation of fitted line, 1 pt for new plot, 1 pt for interpretation

* Lose 1 pt if they don't put hat on the response, and/or if they add hats to the predictors, and/or if they keep $\varepsilon$ in the equation, and/or if they don't square root time in the equation


### Question 8 (3 pts)

$H_0: \beta_1 = \beta_2 = \beta_3 = 0$

$H_A: \text{at least one of } \beta_1, \beta_2, \beta_3 \neq 0$

**Grading:** 

* Lose 1 pt if they include $\beta_0$ in the hypotheses
* Lose 1 pt if they don't correctly negate the null for $H_A$ (e.g., if they say all of them are not 0)
* Lose 1 pt if they don't write the hypotheses in terms of the parameters


### Question 9 (4 pts)

```{r}
anova(city_lm)
```

Test statistic: 20.993

Degrees of freedom: 3 and 1996

p-value: $2.17 \times 10^{-13}$

Conclusion: we have strong evidence for a difference in mean commute times between the different cities

**Grading:** 1 pt for test statistic, 1 pt for degrees of freedom, 1 pt for p-value, 1 pt for conclusion