---
title: "Homework 6 Solutions"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
team_data <- read_csv("https://sta112-s22.github.io/homework/team_data.csv")
```

### Question 1

$H_0: \beta_5 = 0 \hspace{1cm} H_A: \beta_5 \neq 0$

### Question 2

**Full model:**

$\text{performance} = \beta_0 + \beta_1 \text{size} + \beta_2 \text{females} + \beta_3 \text{diversity} + \beta_4 \text{testosterone} \\ \hspace{3cm} + \beta_5 \text{diversity} \cdot \text{testosterone} + \varepsilon$

**Reduced model:**

$\text{performance} = \beta_0 + \beta_1 \text{size} + \beta_2 \text{females} + \beta_3 \text{diversity} + \beta_4 \text{testosterone} + \varepsilon$


### Question 3 

Numerator degrees of freedom: 1 (we are only testing a single parameter)

Denominator degrees of freedom: 68 (= 74 - 6, because there are 74 observations and 6 parameters in the full model)



### Question 4 

```{r}
m1 <- lm(performance ~ size + females + diversity*testosterone,
         data = team_data)
m0 <- lm(performance ~ size + females + diversity + testosterone,
         data = team_data)
anova(m0, m1)
```

Test statistic: $F = 21.1$ 

p-value: $1.94 \times 10^{-5}$

conclusion: we have strong evidence that the relationship between diversity and group performance depends on the level of testosterone

### Question 5

```{r}
summary(m1)
```

Degrees of freedom: 68

Test statistic: $t = -4.594$

p-value: $1.94 \times 10^{-5}$

### Question 6 

The $F = 21.1 = (-4.594)^2 = t^2$


### Question 7 

We'll do a 95% confidence interval (you could do a different level if you want). The critical value $t^*$ (from a $t_{68}$ distribution) is 2.00:

```{r}
qt(0.025, 68, lower.tail=F)
```

So, our confidence interval is 

$$-0.148 \pm 2.00 \cdot 0.0322 = (-0.212, -0.084)$$

We are 95% confident that for a one-unit increase in testosterone, the slope on diversity decreases by between 0.084 and 0.212 units. 



