---
title: Multicollinearity, and choosing a model
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

### Agenda

.large[
* Reminders:
  * Project 1, Part 4 due Friday
  * HW 6 due Friday
* Today: 
  * Finish multicollinearity
  * Start discussing model choice
]

---

### Project data

.large[

* `CONTROL`: whether the school is public or private
* `SATVRMID`: midpoint of SAT critical reading scores of students attending the school
* `SATMTMID`: midpoint of SAT math scores
* `SATWRMID`: midpoint of SAT writing scores
* `ACTCMMID`: midpoint of the ACT cumulative scores
* `ACTENMID`: midpoint of ACT English scores
* `ACTMTMID`: midpoint of ACT math scores
* `+` others...

.question[
Which predictors might exhibit multicollinearity?
]
]

---

### Project data

.large[

* `CONTROL`: whether the school is public or private
* `SATVRMID`: midpoint of SAT critical reading scores of students attending the school
* `SATMTMID`: midpoint of SAT math scores
* `SATWRMID`: midpoint of SAT writing scores
* `ACTCMMID`: midpoint of the ACT cumulative scores
* `ACTENMID`: midpoint of ACT English scores
* `ACTMTMID`: midpoint of ACT math scores
* `+` others...

There is probably multicollinearity with all the test score variables
]

---

### Project data

.large[

* `CONTROL`: whether the school is public or private
* `SATVRMID`: midpoint of SAT critical reading scores of students attending the school
* `SATMTMID`: midpoint of SAT math scores
* `SATWRMID`: midpoint of SAT writing scores
* `ACTCMMID`: midpoint of the ACT cumulative scores
* `ACTENMID`: midpoint of ACT English scores
* `ACTMTMID`: midpoint of ACT math scores
* `+` others...

.question[
How can we visually investigate correlated predictors?
]
]

---

### Pairs plot

.large[
A *pairs plot* lets us vizualize the relationships between pairs of variables:
]

```{r, echo=F, message=F, warning=F, fig.align='center', fig.width=9, fig.height=6}
library(tidyverse)
library(GGally)

scorecard <- read_csv("https://sta112-s22.github.io/projects/scorecard.csv")

scorecard %>%
  dplyr::select(SATVRMID, SATMTMID, SATWRMID, ACTCMMID, ACTENMID, ACTMTMID) %>%
  ggpairs(lower = list(continuous = "cor"),
          upper = list(continuous = "points")) +
  theme_minimal()
```

---

### Pairs plot

```{r, echo=F, message=F, warning=F, fig.align='center', fig.width=8, fig.height=5}
library(tidyverse)
library(GGally)

scorecard <- read_csv("https://sta112-s22.github.io/projects/scorecard.csv")

scorecard %>%
  dplyr::select(SATVRMID, SATMTMID, SATWRMID, ACTCMMID, ACTENMID, ACTMTMID) %>%
  ggpairs(lower = list(continuous = "cor"),
          upper = list(continuous = "points")) +
  theme_minimal()
```

.large[
.question[
What does this pairs plot tell us?
]
]

---

### Pairs plot

```{r, echo=F, message=F, warning=F, fig.align='center', fig.width=8, fig.height=5}
library(tidyverse)
library(GGally)

scorecard <- read_csv("https://sta112-s22.github.io/projects/scorecard.csv")

scorecard %>%
  dplyr::select(SATVRMID, SATMTMID, SATWRMID, ACTCMMID, ACTENMID, ACTMTMID) %>%
  ggpairs(lower = list(continuous = "cor"),
          upper = list(continuous = "points")) +
  theme_minimal()
```

.large[
As expected, all the test score variables are super highly correlated
]

---

### Multicollinearity

.large[
Suppose I fit the model

$$\text{Earnings} = \beta_0 + \beta_1 \text{SATVerbal} + \beta_2 \text{SATWriting} + \beta_3 \text{SATMath} \\  + \beta_4 \text{ACTCumulative} + \beta_5 \text{ACTEnglish} \\ + \beta_7 \text{ACTMath} + \beta_9 \text{ACTWriting} + \varepsilon$$

.question[
How do I measure the multicollinearity of the predictors?
]
]

--

.large[
With variance inflation factors!
]

---

### VIFs in R

.large[
Fortunately, R makes it easy to calculate VIFs:

```{r message=F, warning=F}
library(car)
m1 <- lm(MD_EARN_WNE_P10 ~ SATVRMID + SATMTMID + 
           SATWRMID + ACTCMMID + ACTENMID + 
           ACTMTMID + ACTWRMID,
         data = scorecard)
vif(m1)
```

* The `vif` function (from the `car` package) calculates a VIF for each predictor
* The variance inflation factor for SAT verbal score is 29.78, e.g.
]

---

### Handling multicollinearity

.large[
We have a few options for handling multicollinearity:

* Remove variables
  * Example: just pick one of the test score variables, and discard the rest, since they are so highly correlated
]

---

### Handling multicollinearity

.large[
We have a few options for handling multicollinearity:

* Remove variables
  * Example: just pick one of the test score variables, and discard the rest, since they are so highly correlated
* Create a new variable
  * Example: take an average or sum of the test score variables, instead of using all of them individually
]

---

### Handling multicollinearity

.large[
We have a few options for handling multicollinearity:

* Remove variables
  * Example: just pick one of the test score variables, and discard the rest, since they are so highly correlated
* Create a new variable
  * Example: take an average or sum of the test score variables, instead of using all of them individually
* Ignore it
  * If we don't care about interpreting coefficients or doing inference, multicollinearity doesn't matter as much
]

---

### Multiple regression so far

.large[
* Polynomial regression and categorical predictors
* Model interpretation and interaction terms
* Prediction and inference
* Assumptions, diagnostics, and multicollinearity

.question[
Final topic in multiple regression: how do you know which model to use?
]

* Today: choosing a model for inference (hypothesis testing)
* Friday: automatic variable selection techniques
]

---

### Class activity

.large[
[https://sta112-s22.github.io/class_activities/ca_lecture_25.html](https://sta112-s22.github.io/class_activities/ca_lecture_25.html)
]

---

### Class activity

.large[
.question[
What model can the researchers use to investigate their research questions?
]
]

--

.large[
$$\text{genetic diversity} = \beta_0 + \beta_1 \text{mass} + \beta_2 \text{latitude} + \\ \hspace{4cm} \beta_3 \text{IsMigratory} + \beta_4 \text{breeding range} +\\ \hspace{4cm} \beta_5 \text{breeding range} \cdot \text{IsMigratory} + \varepsilon$$
]

---

### Class activity

.large[

$$\text{genetic diversity} = \beta_0 + \beta_1 \text{mass} + \beta_2 \text{latitude} + \\ \hspace{4cm} \beta_3 \text{IsMigratory} + \beta_4 \text{breeding range} +\\ \hspace{4cm} \beta_5 \text{breeding range} \cdot \text{IsMigratory} + \varepsilon$$

.question[
Which hypotheses allow the researchers to test whether there is any relationship between breeding range and genetic diversity, after accounting for migratory status, body size, and latitude?
]
]

--

.large[
$H_0: \beta_4 = \beta_5 = 0$

$H_A: \text{ at least one of } \beta_4, \beta_5 \neq 0$
]

---

### Class activity

.large[

$H_0: \beta_4 = \beta_5 = 0$

$H_A: \text{ at least one of } \beta_4, \beta_5 \neq 0$

.question[
What test can the researchers use the test these hypotheses?
]
]

--

.large[
A nested F test

.question[
What will be the degrees of freedom?
]
]

--

.large[
2 and 381
]

---

### Class activity

.large[

$$\text{genetic diversity} = \beta_0 + \beta_1 \text{mass} + \beta_2 \text{latitude} + \\ \hspace{4cm} \beta_3 \text{IsMigratory} + \beta_4 \text{breeding range} +\\ \hspace{4cm} \beta_5 \text{breeding range} \cdot \text{IsMigratory} + \varepsilon$$

.question[
Which hypotheses allow the researchers to test whether the slope of the relationship between breeding range and genetic diversity is different for migratory and non-migratory species, after accounting for body size and latitude?
]
]

--

.large[
$H_0: \beta_5 = 0$

$H_A: \beta_5 \neq 0$
]

---

### Class activity

.large[

$H_0: \beta_5 = 0$

$H_A: \beta_5 \neq 0$

.question[
What test can the researchers use the test these hypotheses?
]
]

--

.large[
A nested F test or a t test

.question[
What will be the degrees of freedom?
]
]

--

.large[
F test: 1 and 381

t test: 381
]

