---
title: Introduction to multiple linear regression
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

### Agenda

.large[
* Mask policy
* Statistics classes next semester
* Multiple linear regression
]

---

### Masks

.large[
* On March 19, masks will become optional in most spaces on campus
* However, faculty can request that masks be worn in classroom spaces
* Please let me know if there are any circumstances or needs that I should be aware of when making my decision
  * Anonymous survey sent out
]

---

### Statistics classes next semester

.large[
* STA 214 (Generalized linear models)
  * Requires MST 111 (Calc 1)
* STA 310 (Probability)
  * Requires MST 112 (Calc 2)
* STA 362 (Multivariate statistics)
  * Requires MST 121 or MST 205 (Linear algebra)
* STA 363 (Statistical learning)
  * Requires MST 121 or MST 205 (Linear algebra)
]

---

### Relationships between > 2 variables

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=8, fig.height=5}
library(palmerpenguins)
library(tidyverse)
library(latex2exp)
library(Stat2Data)
library(knitr)

penguins %>%
  drop_na() %>%
  ggplot(aes(x = body_mass_g,
             y = bill_length_mm,
             color = species)) +
  geom_point(size = 2) +
  labs(x = "Body mass (g)",
       y = "Bill length (mm)",
       color = "Species") +
  theme_bw() +
  theme(text = element_text(size = 25))

hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })
```

.large[
.question[
How can we model the relationship between body mass, species, and bill length?
]
]

---

### Regression with a single quantitative predictor

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=6, fig.height=4}

penguins %>%
  drop_na() %>%
  ggplot(aes(x = body_mass_g,
             y = bill_length_mm)) +
  geom_point() +
  labs(x = "Body mass (g)",
       y = "Bill length (mm)") +
  theme_bw() +
  theme(text = element_text(size = 20))
```

.large[
**Model:**

.center[
$\text{bill length} = \beta_0 + \beta_1 \text{body mass} + \varepsilon$
]
]

---

### Regression with a single categorical predictor

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=6, fig.height=4}

penguins %>%
  drop_na() %>%
  ggplot(aes(x = species,
             y = bill_length_mm)) +
  geom_boxplot() +
  labs(x = "Species",
       y = "Bill length (mm)") +
  theme_bw() +
  theme(text = element_text(size = 20))
```


.large[
**Model:**

.center[
$\text{bill length} = \beta_0 + \beta_1 \text{IsChinstrap} + \beta_2 \text{IsGentoo} + \varepsilon$
]
]

---

### Regression with $> 2$ variables

.large[
$\text{bill length} = \beta_0 + \beta_1 \text{body mass} + \varepsilon$

$\text{bill length} = \beta_0 + \beta_1 \text{IsChinstrap} + \beta_2 \text{IsGentoo} + \varepsilon$

.question[
How can we include both species *and* body mass in the model?
]
]

---

## Multiple regression

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4.5}

penguins_no_nas <- penguins %>%
  drop_na()

length_lm <- lm(bill_length_mm ~ species + body_mass_g, 
                data = penguins_no_nas)

penguins_no_nas %>%
  mutate(yhat = predict(length_lm)) %>%
  ggplot(aes(x = body_mass_g,
             y = bill_length_mm,
             color = species)) +
  geom_point(size = 2, alpha = 0.5) +
  geom_line(aes(y = yhat), 
            lwd = 1.5) +
  labs(x = "Body mass (g)",
       y = "Bill length (mm)",
       color = "Species") +
  theme_bw() +
  theme(text = element_text(size = 25))
```


.large[
**Model:**

\begin{align}
\text{bill length} &= \beta_0 + \beta_1 \text{IsChinstrap} + \beta_2 \text{IsGentoo} \\
& \hspace{0.5cm} + \beta_3 \text{body mass} + \varepsilon
\end{align}
]

---

### Multiple regression model

.large[
\begin{align}
\text{bill length} &= \beta_0 + \beta_1 \text{IsChinstrap} + \beta_2 \text{IsGentoo} \\
& \hspace{0.5cm} + \beta_3 \text{body mass} + \varepsilon
\end{align}

**This gives us a separate regression line for each species of penguin.**

.question[
What is the regression line for Adelie penguins?
]
]

--

.large[
$\text{bill length} = \beta_0 + \beta_3 \text{body mass} + \varepsilon$

.question[
What about for Chinstrap and Gentoo penguins?
]
]

---

### Multiple regression model

.large[
\begin{align}
\text{bill length} &= \beta_0 + \beta_1 \text{IsChinstrap} + \beta_2 \text{IsGentoo} \\
& \hspace{0.5cm} + \beta_3 \text{body mass} + \varepsilon
\end{align}

**This gives us a separate regression line for each species of penguin:**

$$\text{bill length} = \begin{cases} \beta_0 + \beta_3 \text{body mass} +  \varepsilon & \text{species = Adelie} \\ \beta_0 + \beta_1 + \beta_3 \text{body mass} + \varepsilon & \text{species = Chinstrap} \\ \beta_0 + \beta_2 + \beta_3 \text{body mass} +  \varepsilon & \text{species = Gentoo} \end{cases}$$

]

---

### Fitted model

.large[
\begin{align}
\widehat{\text{bill length}} &= 24.91 + 9.91 \ \text{IsChinstrap} + 3.54 \ \text{IsGentoo} \\
& \hspace{0.5cm} + 0.0038 \ \text{body mass}
\end{align}

.question[
How would I interpret $\widehat{\beta}_0 = 24.91$?
]
]

--

.large[
The estimated bill length for an Adelie penguin weighing 0g. (not that meaningful)

.question[
What about $\widehat{\beta}_1 = 9.91$ and $\widehat{\beta}_2 = 3.54$?
]
]

---

### Interpreting regression coefficients

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4.5}
library(latex2exp)
penguins_no_nas %>%
  mutate(yhat = predict(length_lm)) %>%
  ggplot(aes(x = body_mass_g,
             y = bill_length_mm)) +
  geom_point(size = 2, alpha = 0) +
  geom_line(aes(y = yhat, color = species), 
            lwd = 1.5) +
  labs(x = "Body mass (g)",
       y = "Bill length (mm)",
       color = "Species") +
  geom_segment(aes(x = 3500, y = 38,
                   xend = 3500, yend = 48),
               arrow = arrow(length = unit(0.3, "cm"))) +
  geom_segment(aes(x = 4500, y = 42,
                   xend = 4500, yend = 45),
               arrow = arrow(length = unit(0.3, "cm"))) +
  annotate("text", x = 3000, y = 41, 
           label = TeX("$9.91", output='character'), 
           parse=TRUE, size = 7) +
  annotate("text", x = 5000, y = 44, 
           label = TeX("$3.54", output='character'), 
           parse=TRUE, size = 7) +
  theme_bw() +
  theme(text = element_text(size = 25))
```


.large[

* $9.91 =$ estimated mean difference in bill length for a Chinstrap and Adelie penguin with the same body mass
* $3.54 =$ estimated mean difference in bill length for a Gentoo and Adelie penguin with the same body mass
]

---

### Interpreting regression coefficients

.large[
\begin{align}
\widehat{\text{bill length}} &= 24.91 + 9.91 \ \text{IsChinstrap} + 3.54 \ \text{IsGentoo} \\
& \hspace{0.5cm} + 0.0038 \ \text{body mass}
\end{align}

.question[
How would I interpret $\widehat{\beta}_3 = 0.0038$?
]
]

--

.large[
The average change in bill length associated with an increase of 1g in body mass, holding species constant.

*Note that we assume the slope is the same for each species!*
]

---

### Predictions

.large[
\begin{align}
\widehat{\text{bill length}} &= 24.91 + 9.91 \ \text{IsChinstrap} + 3.54 \ \text{IsGentoo} \\
& \hspace{0.5cm} + 0.0038 \ \text{body mass}
\end{align}

.question[
What is the predicted bill length for a Gentoo penguin with a body mass of 5000 g?
]
]

--

.large[
$24.91 + 3.54 + 0.0038(5000) = 47.45$
]

---

### Fitting the model in R

.large[
```{r}
m1 <- lm(bill_length_mm ~ species + body_mass_g, 
                 data = penguins_no_nas)
```

* `bill_length_mm ~ species + body_mass_g` is a `formula` in R
* Still `response ~ predictor(s)`
* `+` means "include more variables in the model"
* Need to remove missing values before fitting the model
]

---

### Fitting the model in R

.large[
```{r, output.lines = 10:14}
m1 <- lm(bill_length_mm ~ species + body_mass_g, 
                 data = penguins_no_nas)
summary(m1)
```

\begin{align}
\widehat{\text{bill length}} &= 24.91 + 9.91 \ \text{IsChinstrap} + 3.54 \ \text{IsGentoo} \\
& \hspace{0.5cm} + 0.0038 \ \text{body mass}
\end{align}
]

---

### Class activity

.large[
[https://sta112-s22.github.io/class_activities/ca_lecture_19.html](https://sta112-s22.github.io/class_activities/ca_lecture_19.html)
]

---

### Class activity

.large[
.question[
What is the population regression model?
]
]

--

.large[
\begin{align}
\text{Weight} &= \beta_0 + \beta_1 \text{IsEnlarged} + \beta_2 \text{IsReduced} + \\
& \hspace{1cm} \beta_3 \text{WingLength} + \varepsilon
\end{align}
]

---

### Class activity

.large[
.question[
What is the equation of the fitted regression model?
]
]

--

.large[
\begin{align}
\widehat{\text{Weight}} &= 1.94 - 0.46 \text{IsEnlarged} + 0.90 \text{IsReduced} + \\
& \hspace{1cm} 0.45 \text{WingLength}
\end{align}
]

---

### Class activity

.large[

\begin{align}
\widehat{\text{Weight}} &= 1.94 - 0.46 \text{IsEnlarged} + 0.90 \text{IsReduced} + \\
& \hspace{1cm} 0.45 \text{WingLength}
\end{align}

.question[
What is the average change in weight associated with a 1mm increase in wing length, for sparrows from enlarged nests?
]
]

--

.large[
Answer: 0.45

(We fit the same slope for each group of sparrows, and it is estimated to be 0.45)
]

---

### Class activity

.large[

\begin{align}
\widehat{\text{Weight}} &= 1.94 - 0.46 \text{IsEnlarged} + 0.90 \text{IsReduced} + \\
& \hspace{1cm} 0.45 \text{WingLength}
\end{align}

.question[
What is the estimated difference in weight for sparrows from enlarged vs. control nests, with the same wing length?
]
]

--

.large[
We estimate that sparrows from enlarged nests weigh 0.46g less, on average, than sparrows from control nests with the same wing length
]

---

### Benefits of multiple linear regression

.large[
.question[
What are some benefits of including multiple predictors in my regression model?
]
]

--

.large[
Two main reasons:

* Improve the ability to predict the response
* Account for other variables that may change the relationship (e.g., penguin species)
]

---

### Improving model fit

```{r echo=F, message=F, warning=F, fig.width=7, fig.height=4.5, fig.align='center'}
penguins_no_nas %>%
  ggplot(aes(x = body_mass_g, y = bill_length_mm)) +
  geom_point(size=2) +
  geom_smooth(se=F, method="lm", lwd=1.5) +
  labs(x = "Body mass (g)",
       y = "Bill length (mm)") +
  theme_bw() +
  theme(text = element_text(size = 25))
```

.large[
$$\widehat{\text{bill length}} = 27.15 + 0.004 \ \text{body mass}$$

.question[
How do I measure how well body mass predicts bill length?
]
]

---

### Improving model fit

```{r echo=F, message=F, warning=F, fig.width=7, fig.height=4.5, fig.align='center'}
penguins_no_nas %>%
  ggplot(aes(x = body_mass_g, y = bill_length_mm)) +
  geom_point(size=2) +
  geom_smooth(se=F, method="lm", lwd=1.5) +
  labs(x = "Body mass (g)",
       y = "Bill length (mm)") +
  theme_bw() +
  theme(text = element_text(size = 25))
```

.large[
$$\widehat{\text{bill length}} = 27.15 + 0.004 \ \text{body mass}$$

$R^2 = 0.348$, $R^2_{adj} = 0.346$
]

---

### Improving model fit

```{r echo=F, message=F, warning=F, fig.width=7, fig.height=4.5, fig.align='center'}
length_lm <- lm(bill_length_mm ~ species + body_mass_g, 
                data = penguins_no_nas)

penguins_no_nas %>%
  mutate(yhat = predict(length_lm)) %>%
  ggplot(aes(x = body_mass_g,
             y = bill_length_mm,
             color = species)) +
  geom_point(size = 2, alpha = 0.5) +
  geom_line(aes(y = yhat), 
            lwd = 1.5) +
  labs(x = "Body mass (g)",
       y = "Bill length (mm)",
       color = "Species") +
  theme_bw() +
  theme(text = element_text(size = 25))
```

.large[
But we should probably fit a line for each species...

.question[
Do you think $R^2_{adj}$ will increase or decrease when we add species to the model?
]
]

---

### Improving model fit

```{r echo=F, message=F, warning=F, fig.width=7, fig.height=4.5, fig.align='center'}
length_lm <- lm(bill_length_mm ~ species + body_mass_g, 
                data = penguins_no_nas)

penguins_no_nas %>%
  mutate(yhat = predict(length_lm)) %>%
  ggplot(aes(x = body_mass_g,
             y = bill_length_mm,
             color = species)) +
  geom_point(size = 2, alpha = 0.5) +
  geom_line(aes(y = yhat), 
            lwd = 1.5) +
  labs(x = "Body mass (g)",
       y = "Bill length (mm)",
       color = "Species") +
  theme_bw() +
  theme(text = element_text(size = 25))
```

.large[
\begin{align}
\widehat{\text{bill length}} &= 24.91 + 9.91 \ \text{IsChinstrap} + 3.54 \ \text{IsGentoo} \\
& \hspace{0.5cm} + 0.0038 \ \text{body mass}
\end{align}
]

.large[
$R^2 = 0.806$, $R^2_{adj} = 0.804$
]

---

### Improving model fit

.pull-left[
```{r echo=F, message=F, warning=F, fig.width=6, fig.height=4, fig.align='center'}
penguins_no_nas %>%
  ggplot(aes(x = body_mass_g, y = bill_length_mm)) +
  geom_point(size=2) +
  geom_smooth(se=F, method="lm", lwd=1.5) +
  labs(x = "Body mass (g)",
       y = "Bill length (mm)") +
  theme_bw() +
  theme(text = element_text(size = 25))
```

.large[
$R^2_{adj} = 0.346$
]
]

.pull-right[
```{r echo=F, message=F, warning=F, fig.width=6, fig.height=4, fig.align='center'}

penguins_no_nas %>%
  mutate(yhat = predict(length_lm)) %>%
  ggplot(aes(x = body_mass_g,
             y = bill_length_mm,
             color = species)) +
  geom_point(size = 2, alpha = 0.5) +
  geom_line(aes(y = yhat), 
            lwd = 1.5) +
  labs(x = "Body mass (g)",
       y = "Bill length (mm)",
       color = "Species") +
  theme_bw() +
  theme(text = element_text(size = 25))
```

.large[
$R^2_{adj} = 0.804$
]
]

.large[
* Our ability to predict bill length increases substantially when we add Species to the model
* Make sure to compare $R^2_{adj}$ (rather than $R^2$ ) when comparing models
]

---

### Accounting for other variables

.pull-left[
```{r echo=F, message=F, warning=F, fig.width=6, fig.height=4, fig.align='center'}
penguins_no_nas %>%
  ggplot(aes(x = body_mass_g, y = bill_length_mm)) +
  geom_point(size=2) +
  geom_smooth(se=F, method="lm", lwd=1.5) +
  labs(x = "Body mass (g)",
       y = "Bill length (mm)") +
  theme_bw() +
  theme(text = element_text(size = 25))
```

]

.pull-right[
```{r echo=F, message=F, warning=F, fig.width=6, fig.height=4, fig.align='center'}

penguins_no_nas %>%
  mutate(yhat = predict(length_lm)) %>%
  ggplot(aes(x = body_mass_g,
             y = bill_length_mm,
             color = species)) +
  geom_point(size = 2, alpha = 0.5) +
  geom_line(aes(y = yhat), 
            lwd = 1.5) +
  labs(x = "Body mass (g)",
       y = "Bill length (mm)",
       color = "Species") +
  theme_bw() +
  theme(text = element_text(size = 25))
```

]

.large[
The slope looks pretty similar regardless of whether we add Species to the model. But that isn't always the case...
]

---

### Accounting for other variables

```{r echo=F, message=F, warning=F, fig.width=7, fig.height=5, fig.align='center'}
penguins_raw_no_nas <- penguins_raw %>%
  mutate(species_short = ifelse(Species == "Adelie Penguin (Pygoscelis adeliae)", "Adelie",
                                ifelse(Species == "Gentoo penguin (Pygoscelis papua)", "Gentoo", "Chinstrap"))) %>%
  select(`Culmen Length (mm)`, `Culmen Depth (mm)`, species_short) %>%
  drop_na()

penguins_raw_no_nas %>%
  ggplot(aes(x = `Culmen Length (mm)`, y = `Culmen Depth (mm)`)) +
  geom_point(size=2) +
  geom_smooth(se=F, method="lm", lwd=1.5) +
  theme_bw() +
  theme(text = element_text(size=25))
```

.large[
There appears to be a weak, negative relationship between culmen length and culmen depth.
]

---

### Accounting for other variables

```{r echo=F, message=F, warning=F, fig.width=7, fig.height=5, fig.align='center'}
penguins_raw_no_nas %>%
  ggplot(aes(x = `Culmen Length (mm)`, y = `Culmen Depth (mm)`,
             color = species_short)) +
  geom_point(size=2, alpha = 0.5) +
  geom_smooth(se=F, method="lm", lwd=1.5) +
  labs(color = "Species") +
  theme_bw() +
  theme(text = element_text(size=25))
```

.large[
But when we include Species in the model, we see *positive* relationships between culmen length and culmen depth.
]

---

### Simpson's paradox

.pull-left[
```{r echo=F, message=F, warning=F, fig.width=6, fig.height=4, fig.align='center'}

penguins_raw_no_nas %>%
  ggplot(aes(x = `Culmen Length (mm)`, y = `Culmen Depth (mm)`)) +
  geom_point(size=2) +
  geom_smooth(se=F, method="lm", lwd=1.5) +
  theme_bw() +
  theme(text = element_text(size=25))
```
]

.pull-right[
```{r echo=F, message=F, warning=F, fig.width=6, fig.height=4, fig.align='center'}
penguins_raw_no_nas %>%
  ggplot(aes(x = `Culmen Length (mm)`, y = `Culmen Depth (mm)`,
             color = species_short)) +
  geom_point(size=2, alpha = 0.5) +
  geom_smooth(se=F, method="lm", lwd=1.5) +
  labs(color = "Species") +
  theme_bw() +
  theme(text = element_text(size=25))
```
]

.large[
**Simpson's paradox:** The apparent relationship between variables changes when you look within groups
]

---

### Accounting for other variables

.pull-left[
```{r echo=F, message=F, warning=F, fig.width=6, fig.height=4, fig.align='center'}

penguins_raw_no_nas %>%
  ggplot(aes(x = `Culmen Length (mm)`, y = `Culmen Depth (mm)`)) +
  geom_point(size=2) +
  geom_smooth(se=F, method="lm", lwd=1.5) +
  theme_bw() +
  theme(text = element_text(size=25))
```
]

.pull-right[
```{r echo=F, message=F, warning=F, fig.width=6, fig.height=4, fig.align='center'}
penguins_raw_no_nas %>%
  ggplot(aes(x = `Culmen Length (mm)`, y = `Culmen Depth (mm)`,
             color = species_short)) +
  geom_point(size=2, alpha = 0.5) +
  geom_smooth(se=F, method="lm", lwd=1.5) +
  labs(color = "Species") +
  theme_bw() +
  theme(text = element_text(size=25))
```
]

.large[
* The model *without* Species estimates the relationship between culmen length and culmen depth in the population of all Adelie, Chinstrap, and Gentoo penguins
* The model *with* Species estimates the relationship between culmen length and culmen depth *after accounting for species*
]