---
title: Regression with a categorical predictor
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

```{r include=F}
library(tidyverse)
library(Stat2Data)
data("CO2Germany")
```

### Agenda

.large[
* Project:
  * Part 3 released
  * Part 2 feedback over weekend
* Today: 
  * wrapping up $R^2$ and $R^2_{adj}$
  * categorical predictors
* Faculty guest on Monday. Please don't wear perfume, cologne, or any scented products
]
    
---

### $R^2$: the coefficient of determination

.large[
$$SSE = \sum_i (y_i - \widehat{y}_i)^2$$
$$SSTotal = \sum_i (y_i -\overline{y})^2$$
]

.large[
The proportion of variability in $Y$ "explained" by our regression model is 

$$R^2 = 1 - \dfrac{SSE}{SSTotal}$$
]

.large[
* $R^2$ is between 0 and 1
* A higher $R^2$ means our model is a "better fit" to the data
* For regression with one variable, $R^2 = r^2$ (the squared correlation)
]

---

### Comparing fits

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=5}

library(Stat2Data)
data("CO2Germany")

CO2Germany %>%
  ggplot(aes(x = Day, y = CO2)) +
  geom_point(size = 2.5) +
  geom_smooth(se=F, method="lm", formula = y ~ I(x) + I(x^2), lwd=1.2) +
  geom_smooth(se=F, method="lm", formula = y ~ I(x) + I(x^2) + I(x^3), lwd=1.2,
              color = "red", lty=2) +
  theme_bw() +
  theme(text = element_text(size = 25))
```

---

### Comparing $R^2$

.large[
```{r}
co2_lm <- lm(CO2 ~ I(Day) + I(Day^2), 
               data = CO2Germany)
summary(co2_lm)$r.squared
```
]

.large[
```{r}
co2_lm <- lm(CO2 ~ I(Day) + I(Day^2) + I(Day^3), 
               data = CO2Germany)
summary(co2_lm)$r.squared
```
]

.large[
.question[
$R^2$ has increased very slightly -- does that mean that a polynomial of order 3 is a better fit?
]
]

---

### $R^2$

.large[
$R^2$ **will always increase when you add more parameters to the model.**

Why? Because

$$R^2 = 1 - \dfrac{SSE}{SSTotal}$$
and adding more parameters gives us more flexibility to decrease SSE.
]

---

### Adjusted $R^2$

.large[
We can account for the number of parameters in the model with *adjusted* $R^2$:

$$R^2_{adj} = 1 - \dfrac{SSE/(n-p)}{SSTotal/(n-1)}$$
]

.large[
* $n =$ number of observations in data (sample size)
* $p =$ number of parameters ( $\beta$ terms) in model
]

---

### $R^2_{adj}$

.large[
```{r}
co2_lm <- lm(CO2 ~ I(Day) + I(Day^2), 
               data = CO2Germany)
summary(co2_lm)$adj.r.squared
```
]

.large[
```{r}
co2_lm <- lm(CO2 ~ I(Day) + I(Day^2) + I(Day^3), 
               data = CO2Germany)
summary(co2_lm)$adj.r.squared
```
]

.large[
We can see that $R^2_{adj}$ is slightly higher for the polynomial of order 2
]

---

### Properties of $R^2_{adj}$

.large[
* $R^2_{adj} < R^2$ (always)
* $R^2_{adj}$ accounts for the number of parameters in the model
* $R^2$ can *never* decrease when predictors are added
* $R^2_{adj}$ *can* decrease if unhelpful predictors are added
]

---

### Regression so far

.large[
Simple linear regression:

$$y = \beta_0 + \beta_1 x + \varepsilon$$
Polynomial regression:

$$y = \beta_0 + \beta_1 x + \beta_2 x^2 + \cdots + \beta_k x^k + \varepsilon$$

where $x$ and $y$ are both **quantitative**.

.question[
What if $x$ is *categorical*?
]
]

---

### Categorical predictor

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=6, fig.height=4}
library(palmerpenguins)
library(tidyverse)

penguins_no_nas <- penguins %>%
  drop_na()

penguins_no_nas %>%
  ggplot(aes(x = species, y = bill_length_mm)) +
  geom_boxplot(lwd=1.5) +
  theme_classic() +
  labs(x = "Species", 
       y = "Bill length (mm)") +
  theme(text = element_text(size = 25))
```

.large[
* How do we model the relationship between species and bill length?
* Can we test for a relationship between bill length and species?
]

---

### Categorical predictor

.large[
Imagine population like this:
]

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=6, fig.height=4}

set.seed(1)
penguins_no_nas %>%
  ggplot(aes(x = species, y = bill_length_mm)) +
  geom_jitter(alpha = 0.5, size=2) +
  theme_classic() +
  geom_segment(aes(x = 0.6, y = 38.8, xend = 1.4, yend = 38.8), lwd=1.5, color="blue") + 
  geom_segment(aes(x = 1.6, y = 48.8, xend = 2.4, yend = 48.8), lwd=1.5, color="blue") + 
  geom_segment(aes(x = 2.6, y = 47.6, xend = 3.4, yend = 47.6), lwd=1.5, color="blue") + 
  labs(x = "Species", 
       y = "Bill length (mm)") +
  theme(text = element_text(size = 25))
```

.large[
If $\text{species} = \text{Adelie}$, 

$$\hspace{1cm} \text{bill length} = \text{mean Adelie bill length} + \varepsilon$$
]

---

### Categorical predictor

.large[
Imagine population like this:
]

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=6, fig.height=4}

set.seed(1)
penguins_no_nas %>%
  ggplot(aes(x = species, y = bill_length_mm)) +
  geom_jitter(alpha = 0.5, size=2) +
  theme_classic() +
  geom_segment(aes(x = 0.6, y = 38.8, xend = 1.4, yend = 38.8), lwd=1.5, color="blue") + 
  geom_segment(aes(x = 1.6, y = 48.8, xend = 2.4, yend = 48.8), lwd=1.5, color="blue") + 
  geom_segment(aes(x = 2.6, y = 47.6, xend = 3.4, yend = 47.6), lwd=1.5, color="blue") + 
  labs(x = "Species", 
       y = "Bill length (mm)") +
  theme(text = element_text(size = 25))
```

.large[
If $\text{species} = \text{Chinstrap}$, 

$$\hspace{0.3cm} \text{bill length} = \text{mean Chinstrap bill length} + \varepsilon$$
]

---

### Categorical predictor

.large[
Imagine population like this:
]

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=6, fig.height=4}

set.seed(1)
penguins_no_nas %>%
  ggplot(aes(x = species, y = bill_length_mm)) +
  geom_jitter(alpha = 0.5, size=2) +
  theme_classic() +
  geom_segment(aes(x = 0.6, y = 38.8, xend = 1.4, yend = 38.8), lwd=1.5, color="blue") + 
  geom_segment(aes(x = 1.6, y = 48.8, xend = 2.4, yend = 48.8), lwd=1.5, color="blue") + 
  geom_segment(aes(x = 2.6, y = 47.6, xend = 3.4, yend = 47.6), lwd=1.5, color="blue") + 
  labs(x = "Species", 
       y = "Bill length (mm)") +
  theme(text = element_text(size = 25))
```

.large[

If $\text{species} = \text{Gentoo}$, 

$$\hspace{1cm} \text{bill length} = \text{mean Gentoo bill length} + \varepsilon$$
]


---

### Categorical predictor

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=5, fig.height=3}

library(latex2exp)

set.seed(1)
penguins_no_nas %>%
  ggplot(aes(x = species, y = bill_length_mm)) +
  geom_jitter(alpha = 0) +
  theme_classic() +
  geom_segment(aes(x = 0.6, y = 38, xend = 1.4, yend = 38), lwd=1.5, color="blue") + 
  geom_segment(aes(x = 1.6, y = 50, xend = 2.4, yend = 50), lwd=1.5, color="blue") + 
  geom_segment(aes(x = 2.6, y = 47, xend = 3.4, yend = 47), lwd=1.5, color="blue") + 
  labs(x = "Species", 
       y = "Bill length (mm)") +
  theme(text = element_text(size = 25)) +
  theme(axis.title.y = element_blank()) +
  scale_y_continuous(breaks = c(38),
                     labels = c(TeX("$\\beta_0")))
```

.large[
If $\text{species} = \text{Adelie}, \hspace{1cm} \text{bill length} = \beta_0 + \varepsilon$

* $\beta_0$ = true mean bill length of Adelie penguins (in the population)
]

---

### Categorical predictor

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=5, fig.height=3}

library(latex2exp)

set.seed(1)
penguins_no_nas %>%
  ggplot(aes(x = species, y = bill_length_mm)) +
  geom_jitter(alpha = 0) +
  theme_classic() +
  geom_segment(aes(x = 0.6, y = 38, xend = 1.4, yend = 38), lwd=1.5, color="blue") + 
  geom_segment(aes(x = 1.6, y = 50, xend = 2.4, yend = 50), lwd=1.5, color="blue") + 
  geom_segment(aes(x = 2.6, y = 47, xend = 3.4, yend = 47), lwd=1.5, color="blue") + 
  labs(x = "Species", 
       y = "Bill length (mm)") +
  theme(text = element_text(size = 25)) +
  theme(axis.title.y = element_blank()) +
  scale_y_continuous(breaks = c(38),
                     labels = c(TeX("$\\beta_0")))
```

.large[
If $\text{species} = \text{Adelie}, \hspace{1cm} \text{bill length} = \beta_0 + \varepsilon$

* $\beta_0$ = true mean bill length of Adelie penguins (in the population)

.question[
What notation could I use for Chinstrap penguins?
]
]

---

### Categorical predictor

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=5, fig.height=3}

set.seed(1)
penguins_no_nas %>%
  ggplot(aes(x = species, y = bill_length_mm)) +
  geom_jitter(alpha = 0) +
  theme_classic() +
  geom_segment(aes(x = 0.6, y = 38, xend = 1.4, yend = 38), lwd=1.5, color="blue") + 
  geom_segment(aes(x = 1.6, y = 50, xend = 2.4, yend = 50), lwd=1.5, color="blue") + 
  geom_segment(aes(x = 2.6, y = 47, xend = 3.4, yend = 47), lwd=1.5, color="blue") + 
  labs(x = "Species", 
       y = "Bill length (mm)") +
  theme(text = element_text(size = 25)) +
  theme(axis.title.y = element_blank()) +
  scale_y_continuous(breaks = c(38, 50),
                     labels = c(TeX("$\\beta_0"), 
                                TeX("$\\beta_0 + \\beta_1"))) +
  geom_segment(aes(x = 1.45, y = 38,
                   xend = 1.45, yend = 50),
               arrow = arrow(length = unit(0.3, "cm"))) +
  annotate("text", x = 1.3, y = 44, 
           label = TeX("$\\beta_1", output='character'), 
           parse=TRUE, size = 7)
```

.large[
If $\text{species} = \text{Adelie}, \hspace{1cm} \text{bill length} = \beta_0 + \varepsilon$

* $\beta_0$ = true mean bill length of Adelie penguins (in the population)

If $\text{species} = \text{Chinstrap}, \hspace{0.3cm} \text{bill length} = \beta_0 + \beta_1 + \varepsilon$
]

---

### Categorical predictor

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=5, fig.height=3}

set.seed(1)
penguins_no_nas %>%
  ggplot(aes(x = species, y = bill_length_mm)) +
  geom_jitter(alpha = 0) +
  theme_classic() +
  geom_segment(aes(x = 0.6, y = 38, xend = 1.4, yend = 38), lwd=1.5, color="blue") + 
  geom_segment(aes(x = 1.6, y = 50, xend = 2.4, yend = 50), lwd=1.5, color="blue") + 
  geom_segment(aes(x = 2.6, y = 47, xend = 3.4, yend = 47), lwd=1.5, color="blue") + 
  labs(x = "Species", 
       y = "Bill length (mm)") +
  theme(text = element_text(size = 25)) +
  theme(axis.title.y = element_blank()) +
  scale_y_continuous(breaks = c(38, 50),
                     labels = c(TeX("$\\beta_0"), 
                                TeX("$\\beta_0 + \\beta_1"))) +
  geom_segment(aes(x = 1.45, y = 38,
                   xend = 1.45, yend = 50),
               arrow = arrow(length = unit(0.3, "cm"))) +
  annotate("text", x = 1.3, y = 44, 
           label = TeX("$\\beta_1", output='character'), 
           parse=TRUE, size = 7)
```

.large[
If $\text{species} = \text{Adelie}, \hspace{1cm} \text{bill length} = \beta_0 + \varepsilon$

If $\text{species} = \text{Chinstrap}, \hspace{0.3cm} \text{bill length} = \beta_0 + \beta_1 + \varepsilon$

.question[
What about Gentoo penguins?
]
]

---

### Categorical predictor

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=5, fig.height=3}

set.seed(1)
penguins_no_nas %>%
  ggplot(aes(x = species, y = bill_length_mm)) +
  geom_jitter(alpha = 0) +
  theme_classic() +
  geom_segment(aes(x = 0.6, y = 38, xend = 1.4, yend = 38), lwd=1.5, color="blue") + 
  geom_segment(aes(x = 1.6, y = 50, xend = 2.4, yend = 50), lwd=1.5, color="blue") + 
  geom_segment(aes(x = 2.6, y = 47, xend = 3.4, yend = 47), lwd=1.5, color="blue") + 
  labs(x = "Species", 
       y = "Bill length (mm)") +
  theme(text = element_text(size = 25)) +
  theme(axis.title.y = element_blank()) +
  scale_y_continuous(breaks = c(38, 50, 47),
                     labels = c(TeX("$\\beta_0"), 
                                TeX("$\\beta_0 + \\beta_1"), 
                                TeX("$\\beta_0 + \\beta_2"))) +
  geom_segment(aes(x = 1.45, y = 38,
                   xend = 1.45, yend = 50),
               arrow = arrow(length = unit(0.3, "cm"))) +
  geom_segment(aes(x = 2.45, y = 38,
                   xend = 2.45, yend = 47),
               arrow = arrow(length = unit(0.3, "cm"))) +
  annotate("text", x = 1.3, y = 44, 
           label = TeX("$\\beta_1", output='character'), 
           parse=TRUE, size = 7) +
  annotate("text", x = 2.3, y = 43, 
           label = TeX("$\\beta_2", output='character'), 
           parse=TRUE, size = 7)
```

.large[
If $\text{species} = \text{Adelie}, \hspace{1cm} \text{bill length} = \beta_0 + \varepsilon$

If $\text{species} = \text{Chinstrap}, \hspace{0.3cm} \text{bill length} = \beta_0 + \beta_1 + \varepsilon$

If $\text{species} = \text{Gentoo}, \hspace{1cm} \text{bill length} = \beta_0 + \beta_2 + \varepsilon$
]

---

### Categorical predictor

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=6, fig.height=4}

set.seed(1)
penguins_no_nas %>%
  ggplot(aes(x = species, y = bill_length_mm)) +
  geom_jitter(alpha = 0) +
  theme_classic() +
  geom_segment(aes(x = 0.6, y = 38, xend = 1.4, yend = 38), lwd=1.5, color="blue") + 
  geom_segment(aes(x = 1.6, y = 50, xend = 2.4, yend = 50), lwd=1.5, color="blue") + 
  geom_segment(aes(x = 2.6, y = 47, xend = 3.4, yend = 47), lwd=1.5, color="blue") + 
  labs(x = "Species", 
       y = "Bill length (mm)") +
  theme(text = element_text(size = 25)) +
  theme(axis.title.y = element_blank()) +
  scale_y_continuous(breaks = c(38, 50, 47),
                     labels = c(TeX("$\\beta_0"), 
                                TeX("$\\beta_0 + \\beta_1"), 
                                TeX("$\\beta_0 + \\beta_2"))) +
  geom_segment(aes(x = 1.45, y = 38,
                   xend = 1.45, yend = 50),
               arrow = arrow(length = unit(0.3, "cm"))) +
  geom_segment(aes(x = 2.45, y = 38,
                   xend = 2.45, yend = 47),
               arrow = arrow(length = unit(0.3, "cm"))) +
  annotate("text", x = 1.3, y = 44, 
           label = TeX("$\\beta_1", output='character'), 
           parse=TRUE, size = 7) +
  annotate("text", x = 2.3, y = 43, 
           label = TeX("$\\beta_2", output='character'), 
           parse=TRUE, size = 7)
```

.large[
$$\text{bill length} = \begin{cases} \beta_0 + \varepsilon & \text{species = Adelie} \\ \beta_0 + \beta_1 + \varepsilon & \text{species = Chinstrap} \\ \beta_0 + \beta_2 + \varepsilon & \text{species = Gentoo} \end{cases}$$
]

--

.large[
.question[
Can we write this more concisely?
]
]

---

### Indicator variables

.large[
$$\text{IsChinstrap} = \begin{cases} 0 & \text{species } \neq \text{Chinstrap} \\ 1 & \text{species = Chinstrap} \end{cases}$$

$$\text{IsGentoo} = \begin{cases} 0 & \text{species } \neq \text{Gentoo} \\ 1 & \text{species = Gentoo} \end{cases}$$
.question[
Do I need another indicator variable to represent Adelie penguins?
]
]

---

### Indicator variables

.large[
$$\text{IsChinstrap} = \begin{cases} 0 & \text{species } \neq \text{Chinstrap} \\ 1 & \text{species = Chinstrap} \end{cases}$$

$$\text{IsGentoo} = \begin{cases} 0 & \text{species } \neq \text{Gentoo} \\ 1 & \text{species = Gentoo} \end{cases}$$

Then:

| Species | IsChinstrap | IsGentoo |
| --- | --- | --- |
| Adelie | 0 | 0 |
| Chinstrap | 1 | 0 |
| Gentoo | 0 | 1 |
]

---

### Indicator variables

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=6, fig.height=3.5}

set.seed(1)
penguins_no_nas %>%
  ggplot(aes(x = species, y = bill_length_mm)) +
  geom_jitter(alpha = 0) +
  theme_classic() +
  geom_segment(aes(x = 0.6, y = 38, xend = 1.4, yend = 38), lwd=1.5, color="blue") + 
  geom_segment(aes(x = 1.6, y = 50, xend = 2.4, yend = 50), lwd=1.5, color="blue") + 
  geom_segment(aes(x = 2.6, y = 47, xend = 3.4, yend = 47), lwd=1.5, color="blue") + 
  labs(x = "Species", 
       y = "Bill length (mm)") +
  theme(text = element_text(size = 25)) +
  theme(axis.title.y = element_blank()) +
  scale_y_continuous(breaks = c(38, 50, 47),
                     labels = c(TeX("$\\beta_0"), 
                                TeX("$\\beta_0 + \\beta_1"), 
                                TeX("$\\beta_0 + \\beta_2"))) +
  geom_segment(aes(x = 1.45, y = 38,
                   xend = 1.45, yend = 50),
               arrow = arrow(length = unit(0.3, "cm"))) +
  geom_segment(aes(x = 2.45, y = 38,
                   xend = 2.45, yend = 47),
               arrow = arrow(length = unit(0.3, "cm"))) +
  annotate("text", x = 1.3, y = 44, 
           label = TeX("$\\beta_1", output='character'), 
           parse=TRUE, size = 7) +
  annotate("text", x = 2.3, y = 43, 
           label = TeX("$\\beta_2", output='character'), 
           parse=TRUE, size = 7)
```

.large[
$$\text{bill length} = \begin{cases} \beta_0 + \varepsilon & \text{species = Adelie} \\ \beta_0 + \beta_1 + \varepsilon & \text{species = Chinstrap} \\ \beta_0 + \beta_2 + \varepsilon & \text{species = Gentoo} \end{cases}$$

Or, more concisely:

$$\text{bill length} = \beta_0 + \beta_1 \text{IsChinstrap} + \beta_2 \text{IsGentoo} + \varepsilon$$
]

---

### Estimated model

.large[
$$\widehat{\text{bill length}} = 38.82 + 10.01 \ \text{IsChinstrap} + 8.74 \ \text{IsGentoo}$$
]

.large[
.question[
* What is the estimated average bill length for each of the three species of penguin?
* There are 333 penguins in the dataset used to fit the model. How many degrees of freedom does the fitted model have?
]
]

---

### Estimated model

.large[
$$\widehat{\text{bill length}} =  38.82 + 10.01 \ \text{IsChinstrap} + 8.74 \ \text{IsGentoo}$$

| Species | $\widehat{\text{bill length}}$ |
| --- | --- |
| Adelie | 38.82 | 
| Chinstrap | 38.82 + 10.01 = 48.83 | 
| Gentoo | 38.82 + 8.74 = 47.56 |
]

--

.large[
**Degrees of freedom:** 333 - 3 = 330
* (Lose one degree of freedom for each parameter estimated)
]

---

### Fitting the model in R

.large[
R will automatically create indicator variables:

```{r}
lm(bill_length_mm ~ species, data = penguins_no_nas)
```

$$\widehat{\text{bill length}} =  38.82 + 10.01 \ \text{IsChinstrap} + 8.74 \ \text{IsGentoo}$$
]

---

### Model assumptions

.large[

$$\text{bill length} = \beta_0 + \beta_1 \text{IsChinstrap} + \beta_2 \text{IsGentoo} + \varepsilon$$

.question[
What model assumptions do you think we make?
]
]

---

### Model assumptions

.large[

$$\text{bill length} = \beta_0 + \beta_1 \text{IsChinstrap} + \beta_2 \text{IsGentoo} + \varepsilon$$

where $\varepsilon \sim N(0, \sigma_{\varepsilon}^2)$ and the $\varepsilon$ are independent of each other.

* No notion of shape assumption with single categorical predictor
* Still assume constant variance, normality, independence, and randomness
]

---

### Assessing constant variance

.large[
For a single categorical predictor, we can create a boxplot of residuals for each group, and calculate the standard deviation of residuals in each group.
]

.pull-left[
```{r echo=F, fig.align='center', fig.width=6, fig.height=4}
penguin_lm <- lm(bill_length_mm ~ species, 
                 data = penguins_no_nas)

penguins_no_nas %>%
  mutate(resids = residuals(penguin_lm)) %>%
  ggplot(aes(x = species, y = resids)) +
  geom_boxplot() +
  theme_bw() +
  labs(x = "Species", 
       y = "Residuals") +
  theme(text = element_text(size = 25))
```
]

.pull-right[
.large[
```{r echo=F, message=F, warning=F}
library(knitr)
penguins_no_nas %>%
  mutate(resids = residuals(penguin_lm)) %>%
  group_by(species) %>%
  summarize(residual_std_dev = round(sd(resids), 2)) %>%
  kable()
```
]
]

---

### Assessing constant variance

.pull-left[
```{r echo=F, fig.align='center', fig.width=6, fig.height=4}
penguin_lm <- lm(bill_length_mm ~ species, 
                 data = penguins_no_nas)

penguins_no_nas %>%
  mutate(resids = residuals(penguin_lm)) %>%
  ggplot(aes(x = species, y = resids)) +
  geom_boxplot() +
  theme_bw() +
  labs(x = "Species", 
       y = "Residuals") +
  theme(text = element_text(size = 25))
```
]

.pull-right[
.large[
```{r echo=F, message=F, warning=F}
library(knitr)
penguins_no_nas %>%
  mutate(resids = residuals(penguin_lm)) %>%
  group_by(species) %>%
  summarize(residual_std_dev = round(sd(resids), 2)) %>%
  kable()
```
]
]

.large[
Looking for similar variability between groups. Rule of thumb: not too concerned about constant variance if

$$\dfrac{\text{largest standard deviation}}{\text{smallest standard deviation}} < 2$$
]

---

### Assessing normality

.large[
Can use a QQ plot again:
]

```{r echo=F, fig.align='center', fig.width=6, fig.height=4}
penguins_no_nas %>%
  mutate(resids = residuals(penguin_lm)) %>%
  ggplot(aes(sample = resids)) +
  geom_qq() +
  geom_qq_line() +
  theme_bw() +
  labs(x = "Theoretical normal quantiles", 
       y = "Observed residual quantiles") +
  theme(text = element_text(size = 25))
```

---

### Class activity

.large[
Practice regression with a single categorical predictor, with experimental data on fruit flies:

[https://sta112-s22.github.io/class_activities/ca_lecture_16.html](https://sta112-s22.github.io/class_activities/ca_lecture_16.html)
]