---
title: Interactions
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

### Agenda

.large[
* HW 5 released today
* Today:
  * Visualizing and choosing interactions
  * Interactions between quantitative predictors
]

---

### Interactions: Allowing the slope to be different

.large[
$$\text{bill length} = \begin{cases} \beta_0 + \beta_3 \text{body mass} +  \varepsilon & \text{if Adelie} \\ \beta_0 + \beta_1 + (\beta_3 + \beta_4) \text{body mass} + \varepsilon & \text{if Chinstrap} \\ \beta_0 + \beta_2 + (\beta_3 + \beta_5) \text{body mass} +  \varepsilon & \text{if Gentoo} \end{cases}$$
]

.large[
Written more concisely:

\begin{align}
\text{bill length} &= \beta_0 + \beta_1 \text{IsChinstrap} + \beta_2 \text{IsGentoo} \\
& \hspace{0.5cm} + \beta_3 \text{body mass} + \beta_4 \text{body mass} \cdot \text{IsChinstrap} \\
& \hspace{0.5cm} + \beta_5 \text{body mass} \cdot \text{IsGentoo} + \varepsilon
\end{align}
]

---

### Interactions

.large[
\begin{align}
\text{bill length} &= \beta_0 + \beta_1 \text{IsChinstrap} + \beta_2 \text{IsGentoo} \\
& \hspace{0.5cm} + \beta_3 \text{body mass} + \beta_4 \text{body mass} \cdot \text{IsChinstrap} \\
& \hspace{0.5cm} + \beta_5 \text{body mass} \cdot \text{IsGentoo} + \varepsilon
\end{align}
]

.large[
* The terms $\text{body mass} \cdot \text{IsChinstrap}$ and $\text{body mass} \cdot \text{IsGentoo}$ denote an *interaction* between body mass and species
* Interactions allow the slope for body mass to depend on species
* $\text{body mass} \cdot \text{IsChinstrap}$ literally means multiplication
]

.large[
.question[
What are some advantages and disadvantages of adding interaction terms?
]
]

---

### Interactions: pros and cons

.large[

**Pros:** Interactions can allow us to

* better capture the true relationships in the data
* better predict the response
* answer different research question (e.g., is there a difference in slope for different penguin species?)

**Cons:** Interactions make the model more complex

* often harder to interpret
* more variability in estimating the model
]

---

### Do we need an interaction?

.large[
Ways to decide on interaction terms:

* Visualization
* Measuring predictive ability (e.g., $R^2_{adj}$)
* Research question
]

---

### Visualization

.large[
For interaction between categorical and quantitative predictors, we can plot the fitted regression lines for each group (with the interaction). Check whether the slopes look different.
]

.pull-left[
```{r echo=F, message=F, warning=F, fig.align='center', fig.height=4, fig.width=6.5}
library(tidyverse)
library(palmerpenguins)
library(knitr)

hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

penguins_no_nas <- penguins %>%
  drop_na()
penguins %>%
  ggplot(aes(x = flipper_length_mm, y = bill_length_mm, color = species)) +
  geom_point(alpha = 0.5) +
  geom_smooth(se=F, method="lm", lwd=2) +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  labs(x = "Flipper length (mm)",
       y = "Bill length (mm)")
```
]

.pull-right[
```{r echo=F, message=F, warning=F, fig.align='center', fig.height=4, fig.width=6.5}
penguins %>%
  ggplot(aes(x = body_mass_g, y = bill_depth_mm, color = species)) +
  geom_point(alpha = 0.5) +
  geom_smooth(se=F, method="lm", lwd=2) +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  labs(x = "Body mass (g)",
       y = "Bill depth (mm)")
```
]

.large[
.question[
Which plot suggests we might need an interaction?
]
]

--

.large[
The one on the left -- the slopes look different in the different groups
]

---

### Measuring predictive ability

.large[
Compare $R^2_{adj}$ with and without the interaction term.
]

.pull-left[
```{r echo=F, message=F, warning=F, fig.align='center', fig.height=4, fig.width=6.5}
library(tidyverse)
library(palmerpenguins)
penguins %>%
  ggplot(aes(x = flipper_length_mm, y = bill_length_mm, color = species)) +
  geom_point(alpha = 0.5) +
  geom_smooth(se=F, method="lm", lwd=2) +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  labs(x = "Flipper length (mm)",
       y = "Bill length (mm)")
```

.large[
With interaction (different slopes): $R^2_{adj} = 0.782$
Without interaction (same slopes): $R^2_{adj} = 0.774$
]
]

.pull-right[
```{r echo=F, message=F, warning=F, fig.align='center', fig.height=4, fig.width=6.5}
penguins %>%
  ggplot(aes(x = body_mass_g, y = bill_depth_mm, color = species)) +
  geom_point(alpha = 0.5) +
  geom_smooth(se=F, method="lm", lwd=2) +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  labs(x = "Body mass (g)",
       y = "Bill depth (mm)")
```

.large[
With interaction (different slopes): $R^2_{adj} = 0.8008$
Without interaction (same slopes): $R^2_{adj} = 0.8012$
]
]

---

### Measuring predictive ability

.pull-left[
```{r echo=F, message=F, warning=F, fig.align='center', fig.height=3.5, fig.width=6.5}
library(tidyverse)
library(palmerpenguins)
penguins %>%
  ggplot(aes(x = flipper_length_mm, y = bill_length_mm, color = species)) +
  geom_point(alpha = 0.5) +
  geom_smooth(se=F, method="lm", lwd=2) +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  labs(x = "Flipper length (mm)",
       y = "Bill length (mm)")
```

.large[
With interaction (different slopes): $R^2_{adj} = 0.782$
Without interaction (same slopes): $R^2_{adj} = 0.774$
]
]

.pull-right[
```{r echo=F, message=F, warning=F, fig.align='center', fig.height=3.5, fig.width=6.5}
penguins %>%
  ggplot(aes(x = body_mass_g, y = bill_depth_mm, color = species)) +
  geom_point(alpha = 0.5) +
  geom_smooth(se=F, method="lm", lwd=2) +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  labs(x = "Body mass (g)",
       y = "Bill depth (mm)")
```

.large[
With interaction (different slopes): $R^2_{adj} = 0.8008$
Without interaction (same slopes): $R^2_{adj} = 0.8012$
]
]

.large[
.question[
If the interaction only increases $R^2_{adj}$ by a small amount, we might still prefer the simpler model.
]
]

---

### Considering research questions

.large[
Common types of research questions:

* Description (want to describe the overall relationship between 2 or more variables)
* Prediction (want to do a good job predicting the response)
* Inference (want to test hypotheses about the relationship)
]

---

### Description

.large[
**Example:** What is the relationship between flipper length and bill length, accounting for species?
]

.pull-left[
```{r echo=F, message=F, warning=F, fig.align='center', fig.height=4, fig.width=6.5}
penguins %>%
  ggplot(aes(x = flipper_length_mm, y = bill_length_mm, color = species)) +
  geom_point(alpha = 0.5) +
  geom_smooth(se=F, method="lm", lwd=2) +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  labs(x = "Flipper length (mm)",
       y = "Bill length (mm)")
```

.large[
With interaction
]

]

.pull-right[
```{r echo=F, message=F, warning=F, fig.align='center', fig.height=4, fig.width=6.5}
library(moderndive)
penguins %>%
  ggplot(aes(x = flipper_length_mm, y = bill_length_mm, color = species)) +
  geom_point(alpha = 0.5) +
  geom_parallel_slopes(se=F, lwd=2) +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  labs(x = "Flipper length (mm)",
       y = "Bill length (mm)")
```

.large[
Without interaction
]

]

.large[
.question[
Do we need the interaction to describe the general relationship between flipper length and bill length?
]
]

---

### Inference

.large[
**Example:** Is there a difference in the slope on flipper length, for different species?
]

.pull-left[
```{r echo=F, message=F, warning=F, fig.align='center', fig.height=4, fig.width=6.5}
penguins %>%
  ggplot(aes(x = flipper_length_mm, y = bill_length_mm, color = species)) +
  geom_point(alpha = 0.5) +
  geom_smooth(se=F, method="lm", lwd=2) +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  labs(x = "Flipper length (mm)",
       y = "Bill length (mm)")
```

.large[
With interaction
]

]

.pull-right[
```{r echo=F, message=F, warning=F, fig.align='center', fig.height=4, fig.width=6.5}
library(moderndive)
penguins %>%
  ggplot(aes(x = flipper_length_mm, y = bill_length_mm, color = species)) +
  geom_point(alpha = 0.5) +
  geom_parallel_slopes(se=F, lwd=2) +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  labs(x = "Flipper length (mm)",
       y = "Bill length (mm)")
```

.large[
Without interaction
]

]

.large[
.question[
Do we need to include the interaction to answer this question?
]
]

---

### Inference

.large[
**Example:** Is there a difference in the slope on flipper length, for different species?
]

.pull-left[
```{r echo=F, message=F, warning=F, fig.align='center', fig.height=4, fig.width=6.5}
penguins %>%
  ggplot(aes(x = flipper_length_mm, y = bill_length_mm, color = species)) +
  geom_point(alpha = 0.5) +
  geom_smooth(se=F, method="lm", lwd=2) +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  labs(x = "Flipper length (mm)",
       y = "Bill length (mm)")
```

]

.pull-right[
```{r echo=F, message=F, warning=F, fig.align='center', fig.height=4, fig.width=6.5}
library(moderndive)
penguins %>%
  ggplot(aes(x = flipper_length_mm, y = bill_length_mm, color = species)) +
  geom_point(alpha = 0.5) +
  geom_parallel_slopes(se=F, lwd=2) +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  labs(x = "Flipper length (mm)",
       y = "Bill length (mm)")
```

]

.large[
.question[
Do we need to include the interaction to answer this question?
]

Yes -- we can't check if the slopes are different without fitting different slopes!
]

---

### Fitting the model with interactions

.large[
```{r message=F}
m1 <- lm(bill_length_mm ~ species*flipper_length_mm,
         data = penguins_no_nas)
```

* `*` means "include both variables *and* their interaction"
]

---

### Fitting the model with interactions

.large[
```{r output.lines = 11:16}
summary(m1)
```

\begin{align}
\widehat{\text{bill length}} &= 13.04 - 7.44 \ \text{IsChinstrap} - 33.52 \ \text{IsGentoo} \\
& \hspace{1cm} + 0.136 \ \text{flipper length} \\
& \hspace{1cm} + 0.085 \ \text{flipper length} \cdot \text{IsChinstrap} \\
& \hspace{1cm} + 0.178 \ \text{flipper length} \cdot \text{IsGentoo}
\end{align}

]

---

### Class activity

---

### Class activity

.large[
\begin{align}
\widehat{\text{Height}} &= 40.689 - 7.728 \ \text{IsMale} + 0.118 \ \text{Age} \\
& \hspace{1cm} + 0.064 \ \text{Age} \cdot \text{IsMale}
\end{align}

.question[
What is the estimated height for a girl aged 120 months (10 years)?
]
]

--

.large[
$40.689 + 0.118(120) = 54.85 inches$ (about 4 feet 6 inches)
]

---

### Class activity

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=6.5, fig.height=4}
library(Stat2Data)
data("Kids198")
kids <- Kids198 %>%
  mutate(Sex = ifelse(Sex == 0, "male", "female"))

kids %>%
  ggplot(aes(x = Age, y = Height, color = Sex)) +
  geom_point() +
  geom_smooth(se=F, method="lm", lwd=2) +
  theme_bw() +
  theme(text = element_text(size = 25))
```

.large[
With interaction: $R^2_{adj} = 0.763$

Without interaction: $R^2_{adj} = 0.732$
]

.large[
.question[
Do you think an interaction is needed?
]
]

---

### Class activity

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7.5, fig.height=4.5}
library(Stat2Data)
data("Kids198")
kids <- Kids198 %>%
  mutate(Sex = ifelse(Sex == 0, "male", "female"))

kids %>%
  ggplot(aes(x = Age, y = Height, color = Sex)) +
  geom_point() +
  geom_smooth(se=F, method="lm", lwd=2) +
  theme_bw() +
  theme(text = element_text(size = 25))
```

.large[
I think an interaction is important
* slopes look different 
* captures that boys and girls start off with similar heights, but boys eventually become taller.
]

---

### Interactions with a categorical predictor

.large[
\begin{align}
\text{bill length} &= \beta_0 + \beta_1 \text{IsChinstrap} + \beta_2 \text{IsGentoo} \\
& \hspace{0.5cm} + \beta_3 \text{body mass} + \beta_4 \text{body mass} \cdot \text{IsChinstrap} \\
& \hspace{0.5cm} + \beta_5 \text{body mass} \cdot \text{IsGentoo} + \varepsilon
\end{align}
]

.large[
The slope on body mass depends on species:

* Slope for Adelie penguins: $\beta_3$
* Slope for Chinstrap penguins: $\beta_3 + \beta_4$
* Slope for Gentoo penguins: $\beta_3 + \beta_5$
]

---

### Interactions with a categorical predictor

.large[
\begin{align}
\text{bill length} &= \beta_0 + \beta_1 \text{IsChinstrap} + \beta_2 \text{IsGentoo} \\
& \hspace{0.5cm} + \beta_3 \text{body mass} + \beta_4 \text{body mass} \cdot \text{IsChinstrap} \\
& \hspace{0.5cm} + \beta_5 \text{body mass} \cdot \text{IsGentoo} + \varepsilon
\end{align}
]

.large[
The slope on body mass depends on species:

$\text{slope} = \beta_3 + \beta_4 \ \text{IsChinstrap} + \beta_5 \ \text{IsGentoo}$

.question[
We can do something similar for two quantitative variables
]
]

---

### Interactions with two quantitative variables

.large[

Additive model (no interactions):

$$\text{bill length} = \beta_0 + \beta_1 \text{body mass} + \beta_2 \text{flipper length} + \varepsilon$$

* Slope on body mass is $\beta_1$
* Slope on body mass *does not depend on* flipper length
]

---

### Interactions with two quantitative variables

.large[

$$\text{bill length} = \beta_0 + (\text{slope}) \ \text{body mass} + \beta_2 \text{flipper length} + \varepsilon$$

**Making the slope depend on flipper length:**

slope on body mass = $\beta_1 + \beta_3 \text{flipper length}$

**Substituting:**

\begin{align}
\text{bill length} &= \beta_0 + (\beta_1 + \beta_3 \text{flipper length}) \ \text{body mass} \\
& \hspace{1cm} + \beta_2 \text{flipper length} + \varepsilon \\
&= \beta_0 + \beta_1 \text{body mass} + \beta_2 \text{flipper length} \\
& \hspace{1cm} + \beta_3 \text{body mass} \cdot \text{flipper length} + \varepsilon
\end{align}
]

---

### Interactions with two quantitative variables

.large[
\begin{align}
\text{bill length} &= \beta_0 + \beta_1 \text{body mass} + \beta_2 \text{flipper length} \\
& \hspace{1cm} + \beta_3 \text{body mass} \cdot \text{flipper length} + \varepsilon
\end{align}
]

.large[
* $\text{body mass} \cdot \text{flipper length}$ is the *interaction* between body mass and flipper length
* As before, it literally means multiplication
]

---

### Interactions with two quantitative variables

.large[
\begin{align}
\widehat{\text{bill length}} &= -26.75 + 0.0059 \ \text{body mass} + 0.339 \ \text{flipper length} \\
& \hspace{1cm} - 0.000026 \ \text{body mass} \cdot \text{flipper length}
\end{align}
]

.large[
.question[
What is the predicted bill length for a penguin with body mass 5000g and flipper length 200mm?
]

.abox[
44.55 mm
]

.bbox[
70.55 mm
]

.cbox[
71.3 mm
]
]

--

.large[
$-26.75 + 0.0059(5000) + 0.339(200) - 0.000026(5000)(200)$

$= 44.55$
]

---

### Interactions with two quantitative variables

.large[
\begin{align}
\widehat{\text{bill length}} &= -26.75 + 0.0059 \ \text{body mass} + 0.339 \ \text{flipper length} \\
& \hspace{1cm} - 0.000026 \ \text{body mass} \cdot \text{flipper length}
\end{align}
]

.large[
.question[
What is the change in bill length associated with a 1g increase in body mass, for penguins with flipper length = 200mm?
]

.abox[
0.0059
]

.bbox[
0.005874
]

.cbox[
0.0007
]
]

--

.large[
$0.0059 - 0.000026(200) = 0.0007$
]

---

### Interactions with two quantitative variables

.large[
\begin{align}
\widehat{\text{bill length}} &= -26.75 + 0.0059 \ \text{body mass} + 0.339 \ \text{flipper length} \\
& \hspace{1cm} - 0.000026 \ \text{body mass} \cdot \text{flipper length}
\end{align}
]

.large[
.question[
Is the slope on body mass higher or lower for penguins with longer flippers?
]

.abox[
Higher
]

.bbox[
Lower
]
]

--

.large[
Lower, because the coefficient -0.000026 is negative.
]

---

### Visualizing interactions

.large[
With a categorical variable, we can fit a different slope for each group:
]

```{r echo=F, message=F, warning=F, fig.align='center', fig.height=5, fig.width=7}
penguins %>%
  ggplot(aes(x = flipper_length_mm, y = bill_length_mm, color = species)) +
  geom_point(alpha = 0.5) +
  geom_smooth(se=F, method="lm", lwd=2) +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  labs(x = "Flipper length (mm)",
       y = "Bill length (mm)")
```

.large[
.question[
How would this work with two quantitative variables?
]
]

---

### Visualizing interactions

.large[
**Binarize:**

$$LongFlipper = \begin{cases} 1 & \text{flipper length} > 200 \\ 0 & \text{flipper length} \leq 200 \end{cases}$$
]

```{r echo=F, message=F, warning=F, fig.align='center', fig.height=5, fig.width=7}
penguins %>%
  drop_na() %>%
  mutate(LongFlipper = ifelse(flipper_length_mm > 200, "1", "0")) %>%
  ggplot(aes(x = body_mass_g, y = bill_length_mm, color = LongFlipper)) +
  geom_point(alpha = 0.5) +
  geom_smooth(se=F, method="lm", lwd=2) +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  labs(x = "Body mass (g)",
       y = "Bill length (mm)")
```