---
title: Assumptions, diagnostics, and multicollinearity
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

### Agenda

.large[
* Exam 2 next Wednesday (April 6)
  * polynomial regression
  * categorical data
  * multiple regression
* Today: 
  * multiple regression assumptions
  * Diagnostics
  * Multicollinearity
]

---

### Data

.large[
Data on the number of doctors, hospitals, and beds in 53 randomly selected US counties. Variables include:

* `MDs`: number of doctors in the county
* `Hospitals`: number of hospitals in the county
* `Beds`: number of beds in the hospitals

**Goal:** Predict the number of doctors using the number of hospitals and the number of beds
]

---

### Multiple regression model

.large[
```{r include=F}
library(tidyverse)
library(Stat2Data)
data("CountyHealth")
```

$$\text{MDs} = \beta_0 + \beta_1 \ \text{Hospitals} + \beta_2 \ \text{Beds} + \varepsilon$$
.question[
What assumptions does our multiple regression model make?
]
]

--

.large[
The same assumptions as before:

* Shape
* Constant variance
* Normality
* Independence
* Randomness
]

---

### Shape and constant variance

.large[
$$\text{MDs} = \beta_0 + \beta_1 \ \text{Hospitals} + \beta_2 \ \text{Beds} + \varepsilon$$

**Shape:** the shape of the regression model is at least approximately correct
  * In this case, the number of doctors has a linear relationship with the number of hospitals and the number of beds
  
**Constant variance:** The variance of the noise term $\varepsilon$ is the same for all numbers of hospitals and beds

.question[
How do we assess the shape and constant variance assumptions?
]
]

--

.large[
A residual plot!
]

---

### Residual plot

.large[
$\widehat{\text{MDs}} = -472.08 + 117.40 \ \text{Hospitals} + 1.26 \ \text{Beds}$

.question[
What goes on the x and y axes of my residual plot?
]
]

--

.large[
* x axis: the predicted response ( $\widehat{\text{MDs}}$ )
* y axis: the residuals ( $\text{MDs} - \widehat{\text{MDs}}$ )
]

---

### Residual plot

.large[

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4.5}
m1 <- lm(MDs ~ Hospitals + Beds, data = CountyHealth)

CountyHealth %>%
  mutate(preds = predict(m1),
         resids = residuals(m1)) %>%
  ggplot(aes(x = preds, y = resids)) +
  geom_point(size=2.5) +
  geom_abline(slope = 0, intercept = 0, color="blue", lwd=1.2) +
  theme_bw() +
  labs(x = "Predicted number of doctors", 
       y = "Residual") +
  theme(text = element_text(size = 25))
```

.question[
How do the shape and constant variance assumptions look?
]
]

--

.large[
Not good -- violations to both shape and constant variance
]

---

### Residual plot

.large[

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4.5}
m1 <- lm(MDs ~ Hospitals + Beds, data = CountyHealth)

CountyHealth %>%
  mutate(preds = predict(m1),
         resids = residuals(m1)) %>%
  ggplot(aes(x = preds, y = resids)) +
  geom_point(size=2.5) +
  geom_abline(slope = 0, intercept = 0, color="blue", lwd=1.2) +
  theme_bw() +
  labs(x = "Predicted number of doctors", 
       y = "Residual") +
  theme(text = element_text(size = 25))
```

.question[
How do we address violations of shape and constant variance?
]
]

--

.large[
Transformations!
]

---

### Residual plot

.large[

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4.5}
m1 <- lm(MDs ~ Hospitals + Beds, data = CountyHealth)

CountyHealth %>%
  mutate(preds = predict(m1),
         resids = residuals(m1)) %>%
  ggplot(aes(x = preds, y = resids)) +
  geom_point(size=2.5) +
  geom_abline(slope = 0, intercept = 0, color="blue", lwd=1.2) +
  theme_bw() +
  labs(x = "Predicted number of doctors", 
       y = "Residual") +
  theme(text = element_text(size = 25))
```

.question[
What transformation would you recommend?
]
]

--

.large[
Maybe log or square root the response
]

---

### Transforming the data

.large[
$\widehat{\sqrt{\text{MDs}}} = 3.58 + 2.58 \ \text{Hospitals} + 0.012 \ \text{Beds}$


```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4.5}
m1 <- lm(sqrt(MDs) ~ Hospitals + Beds, data = CountyHealth)

CountyHealth %>%
  mutate(preds = predict(m1),
         resids = residuals(m1)) %>%
  ggplot(aes(x = preds, y = resids)) +
  geom_point(size=2.5) +
  geom_abline(slope = 0, intercept = 0, color="blue", lwd=1.2) +
  theme_bw() +
  labs(x = "Predicted number of doctors", 
       y = "Residual") +
  theme(text = element_text(size = 25))
```

This looks better!
]

---

### Normality

.large[
$$\sqrt{\text{MDs}} = \beta_0 + \beta_1 \ \text{Hospitals} + \beta_2 \ \text{Beds} + \varepsilon$$

**Normality assumption:** the noise term $\varepsilon$ follows a normal distribution

.question[
How do we assess the normality assumption?
]
]

--

.large[
With a QQ plot!
]

---

### QQ plot

.large[
$\widehat{\sqrt{\text{MDs}}} = 3.58 + 2.58 \ \text{Hospitals} + 0.012 \ \text{Beds}$

.question[
What goes on the x and y axes of my QQ plot?
]
]

--

.large[
* x axis: theoretical normal quantiles
* y axis: observed residual quantiles
]

---

### QQ plot


```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4.5}
m1 <- lm(sqrt(MDs) ~ Hospitals + Beds, data = CountyHealth)

CountyHealth %>%
  mutate(resids = residuals(m1)) %>%
  ggplot(aes(sample = resids)) +
  geom_qq(size=2.5) +
  geom_qq_line(color="blue", lwd=1.2) +
  theme_bw() +
  labs(x = "Theoretical normal quantiles", 
       y = "Observed residual quantiles") +
  theme(text = element_text(size = 25))
```

.large[
.question[
How does the QQ plot look?
]
]

--

.large[
Pretty good!
]

---

### Independence and randomness

.large[
$$\sqrt{\text{MDs}} = \beta_0 + \beta_1 \ \text{Hospitals} + \beta_2 \ \text{Beds} + \varepsilon$$

**Independence assumption:** Observations are independent -- one point falling above or below the line has no influence on the location of another point

**Randomness assumption:** The data are obtained using a random process, such as a random sample or randomized experiment

.question[
How do we assess the independence and randomness assumptions?
]
]

--

.large[
Think about the data generating process (can't check these assumptions with plots)
]

---

### Independence and randomness

.large[

**Independence assumption:** Observations are independent -- one point falling above or below the line has no influence on the location of another point

**Randomness assumption:** The data are obtained using a random process, such as a random sample or randomized experiment

.question[
Do independence and randomness seem reasonable for the county health data?
]
]

--

.large[
* Independence: reasonable to assume the number of hospitals/doctors for one county doesn't affect the number of hospitals/doctors for another county
* Randomness: we are told the data come from a random sample
]

---

### Comparing models

.large[
With both variables:

* $\widehat{\sqrt{\text{MDs}}} = 3.58 + 2.58 \ \text{Hospitals} + 0.012 \ \text{Beds}$
* $R^2_{adj} = 0.919$

With just Beds:

* $\widehat{\sqrt{\text{MDs}}} = 8.49 + 0.018 \ \text{Beds}$
* $R^2_{adj} = 0.899$

With just Hospitals:

* $\widehat{\sqrt{\text{MDs}}} = -2.75 + 6.88 \ \text{Hospitals}$
* $R^2_{adj} = 0.849$

]

---

### Comparing models

.large[
.question[
Each variable is strongly associated with the response, but using both in the model doesn't improve the $R^2_{adj}$ by much. Why the small change?
]
]

--

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=6.5, fig.height=4.5}
CountyHealth %>%
  ggplot(aes(x = Beds, y = Hospitals)) +
  geom_point(size = 2.5) +
  theme_bw() +
  theme(text = element_text(size=25))
```

---

### Multicollinearity

.large[
Beds and Hospitals are strongly related, so using both doesn't provide much more information than using just one.

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=6.5, fig.height=4}
CountyHealth %>%
  ggplot(aes(x = Beds, y = Hospitals)) +
  geom_point(size = 2.5) +
  theme_bw() +
  theme(text = element_text(size=25))
```

.question[
**Multicollinearity:** one or more predictors is strongly correlated with some combination of the other predictors.
]

]

---

### Multicollinearity

.large[
.question[
**Multicollinearity:** one or more predictors is strongly correlated with some combination of the other predictors.
]
```{r echo=F, message=F, warning=F, fig.align='center', fig.width=6.5, fig.height=4}
CountyHealth %>%
  ggplot(aes(x = Beds, y = Hospitals)) +
  geom_point(size = 2.5) +
  theme_bw() +
  theme(text = element_text(size=25))
```

$r = 0.91$ (correlation between Beds and Hospitals)
]

---

### Measuring multicollinearity

.large[
```{r echo=F, message=F, warning=F, fig.align='center', fig.width=6.5, fig.height=4}
CountyHealth %>%
  ggplot(aes(x = Beds, y = Hospitals)) +
  geom_point(size = 2.5) +
  geom_smooth(method = "lm", se=F, lwd=1.2) +
  theme_bw() +
  theme(text = element_text(size=25))
```

$\widehat{\text{Beds}} = -514.68 + 349.15 \ \text{Hospitals}$

$R^2 = 0.82$ (for predicting Beds from Hospitals)

**Variance inflation factor:** $VIF = \dfrac{1}{1 - R^2} = \dfrac{1}{1 - 0.82} = 5.56$
]

---

### Variance inflation factors

.large[
**Motivation:** High multicollinearity means a predictor can be estimated well by some combination of other predictors

**Variance inflation factor:** Let $X_i$ be a predictor in the model, and $R_i^2$ the coefficient of determination for a model estimating $X_i$ from the other predictors. The *variance inflation factor* of $X_i$ is

$$VIF_i = \dfrac{1}{1 - R_i^2}$$
* Higher VIF means more multicollinearity
* Rule of thumb: worried when $VIF_i > 5$ (i.e. $R_i^2 > 0.8$ )
]

---

### Class activity

---

### Class activity

.large[
* $\widehat{\text{Hgt97}} = 40.59 + 1.10 \ \text{Hgt96} \hspace{1cm} R^2_{adj} =$
]

---

### Class activity

.large[
* $\widehat{\text{Hgt97}} = 40.59 + 1.10 \ \text{Hgt96} \hspace{1cm} R^2_{adj} = 0.949$
* $\widehat{\text{Hgt97}} = 40.24 − 1.26 \ \text{Diam96} + 1.12 \ \text{Hgt96} \hspace{1cm} R^2_{adj} =$
]

---

### Class activity

.large[
* $\widehat{\text{Hgt97}} = 40.59 + 1.10 \ \text{Hgt96} \hspace{1cm} R^2_{adj} = 0.949$
* $\widehat{\text{Hgt97}} = 40.24 − 1.26 \ \text{Diam96} + 1.12 \ \text{Hgt96} \hspace{1cm} R^2_{adj} = 0.942$

.question[
Why doesn't $R^2_{adj}$ increase when I add diameter to the model?
]
]

---

### Class activity

.large[
.question[
Why doesn't $R^2_{adj}$ increase when I add diameter to the model?
]
]

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=5}
data("Pines")
Pines %>%
  ggplot(aes(x = Diam96, y = Hgt96)) +
  geom_point(size=2) +
  theme_bw() +
  theme(text = element_text(size = 25))
```

.large[
Because diameter and height in 1996 are so correlated.
]

---

### Class activity

.large[
$$\widehat{\text{Hgt97}} = 40.24 − 1.26 \ \text{Diam96} + 1.12 \ \text{Hgt96}$$
.question[
What is the variance inflation factor for Diam96?
]
]

--

.large[
```{r}
m1 <- lm(Diam96 ~ Hgt96, data = Pines)
summary(m1)$r.squared
```

$VIF = \dfrac{1}{1 - 0.813} = 5.35$
]

---

### Why is multicollinearity a problem?

.large[
* Inflates variability of estimated coefficients
* Interpretation of individual terms is difficult

$$\widehat{\sqrt{\text{MDs}}} = 3.58 + 2.58 \ \text{Hospitals} + 0.012 \ \text{Beds}$$

**Usual interpretation:** Holding Beds fixed, an increase of 1 hospital is associated with an increase of 2.58 units in $\sqrt{\text{MDs}}$

**Problem:** When Beds and Hospitals are highly correlated, doesn't make sense to fix one and change the other. (Can't really change the number of hospitals without changing the number of beds)

]

