---
title: Logistic regression interpretation
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

### Agenda

.large[
* No class this Friday
* No HW due this week
* Today: interpreting logistic regression models
]

---

### Class activity, Part I

.large[
[https://sta112-s22.github.io/class_activities/ca_lecture_29.html](https://sta112-s22.github.io/class_activities/ca_lecture_29.html)
]

---

### Class activity

.large[
$$\log \left( \dfrac{\widehat{\pi}}{1 - \widehat{\pi}} \right) = -19.21 + 5.45 \ \text{GPA}$$

.question[
What is the estimated odds of acceptance for a student with 4.0 GPA?
]
]

--

.large[
$\exp\{ -19.21 + 5.45 \cdot 4 \} = 13.330$
]

---

### Class activity

.large[
$$\log \left( \dfrac{\widehat{\pi}}{1 - \widehat{\pi}} \right) = -19.21 + 5.45 \ \text{GPA}$$

.question[
Calculate the odds ratio comparing odds of acceptance for a student with a GPA of 4.0 to a student with a GPA of 3.0.
]
]

--

.large[
$\dfrac{\exp\{ -19.21 + 5.45 \cdot 4 \}}{\exp\{ -19.21 + 5.45 \cdot 3 \}} = 232.76$

So the odds of acceptance are 232.76 times higher for a student with a GPA of 4.0, compared to a student with a GPA of 3.0
]

---

### Class activity

.large[
$$\log \left( \dfrac{\widehat{\pi}}{1 - \widehat{\pi}} \right) = -19.21 + 5.45 \ \text{GPA}$$

.question[
Calculate the odds ratio comparing odds of acceptance for a student with a GPA of 3.8 to a student with a GPA of 2.8.
]
]

--

.large[
$\dfrac{\exp\{ -19.21 + 5.45 \cdot 3.8 \}}{\exp\{ -19.21 + 5.45 \cdot 2.8 \}} = 232.76$

So the odds of acceptance are 232.76 times higher for a student with a GPA of 3.8, compared to a student with a GPA of 2.8

.question[
This is the same odds ratio as comparing a 4.0 GPA to a 3.0 GPA!
]
]

---

### Class activity

.large[
$$\log \left( \dfrac{\widehat{\pi}}{1 - \widehat{\pi}} \right) = -19.21 + 5.45 \ \text{GPA}$$

Odds ratio for increasing GPA by one point:

\begin{align}
\text{odds ratio} &= \dfrac{\exp\{ -19.21 + 5.45 \cdot (GPA + 1) \}}{\exp\{ -19.21 + 5.45 \cdot (GPA) \}} \\ 
&= \dfrac{\exp \{ -19.21\} \exp\{5.45 GPA \} \exp\{5.45\} }{\exp \{ -19.21\} \exp\{5.45 GPA \}} \\
&= \exp\{5.45\} \\
&= 232.76
\end{align}

.question[
So, increasing GPA by one point is associated with an increase in the odds of acceptance by a factor of 232.76!
]
]

---

### Interpreting coefficients

.large[
$$\log \left( \dfrac{\widehat{\pi}}{1 - \widehat{\pi}} \right) = \widehat{\beta}_0 + \widehat{\beta}_1 x$$

Interpretation of $\widehat{\beta}_1$ :

* A one unit increase in $x$ is associated with an increase of $\widehat{\beta}_1$ in the log odds
* A one unit increase in $x$ is associated with an increase in the odds by a factor of $\exp\{\widehat{\beta}_1\}$

.question[
How do you think we interpret $\widehat{\beta}_0$?
]
]

---

### Interpreting coefficients

.large[
$$\log \left( \dfrac{\widehat{\pi}}{1 - \widehat{\pi}} \right) = \widehat{\beta}_0 + \widehat{\beta}_1 x$$

Interpretation of $\widehat{\beta}_1$ :

* A one unit increase in $x$ is associated with an increase of $\widehat{\beta}_1$ in the log odds
* A one unit increase in $x$ is associated with an increase in the odds by a factor of $\exp\{\widehat{\beta}_1\}$

Interpretation of $\widehat{\beta}_0$ :

* The estimated log odds when $x = 0$ are $\widehat{\beta}_0$
* The estimated odds when $x = 0$ are $\exp\{\widehat{\beta}_0\}$
]

---

### Converting from log odds to probabilities

.large[
$$\log \left( \dfrac{\widehat{\pi}}{1 - \widehat{\pi}} \right) = \widehat{\beta}_0 + \widehat{\beta}_1 x$$
So,

$$\dfrac{\widehat{\pi}}{1 - \widehat{\pi}} = \exp\{\widehat{\beta}_0 + \widehat{\beta}_1 x \}$$

and thus

$$\widehat{\pi} = \dfrac{\exp\{\widehat{\beta}_0 + \widehat{\beta}_1 x \}}{1 + \exp\{\widehat{\beta}_0 + \widehat{\beta}_1 x \}}$$
]

---

### Shape of the fitted curve

.large[

.pull-left[
$\widehat{\pi} = \dfrac{\exp\{\widehat{\beta}_0 +  x \}}{1 + \exp\{\widehat{\beta}_0 + x \}} \hspace{0.5cm}$

$\widehat{\beta}_0 = -3, -1, 1$

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=6, fig.height=4}
library(tidyverse)

expit <- function(p){
  return(exp(p)/(1 + exp(p)))
}

library(knitr)
hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

data.frame(x = seq(-5, 6, length.out=100),
           y1 = expit(-1 + seq(-5, 6, length.out=100)),
           y2 = expit(-3 + seq(-5, 6, length.out=100)),
           y3 = expit(1 + seq(-5, 6, length.out=100))) %>%
  ggplot(aes(x = x, y = y1)) +
  geom_line(lwd = 1.2) +
  geom_line(aes(y = y2), lwd=1.7, lty = 2, color="blue") +
  geom_line(aes(y = y3), lwd=1.7, lty = 3, color="red") +
  theme_bw() +
  labs(x = "x", y = "P(y = 1)") +
  annotate("text", x = -2, y = 0.4, label="1", size=8) +
  annotate("text", x = 0, y = 0.4, label="-1", size=9) +
  annotate("text", x = 3.5, y = 0.4, label="-3", size=8) +
  theme(text = element_text(size = 25))
```
]

.pull-right[
$\widehat{\pi} = \dfrac{\exp\{-1 +  \widehat{\beta}_1 \ x \}}{1 + \exp\{-1 +  \widehat{\beta}_1 \ x \}} \hspace{0.5cm}$

$\widehat{\beta}_1 = 0.5, 1, 2$

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=6, fig.height=4}

data.frame(x = seq(-5, 6, length.out=100),
           y1 = expit(-1 + 0.5*seq(-5, 6, length.out=100)),
           y2 = expit(-1 + seq(-5, 6, length.out=100)),
           y3 = expit(-1 + 2*seq(-5, 6, length.out=100))) %>%
  ggplot(aes(x = x, y = y1)) +
  geom_line(lwd = 1.2) +
  geom_line(aes(y = y2), lwd=1.7, lty = 2, color="blue") +
  geom_line(aes(y = y3), lwd=1.7, lty = 3, color="red") +
  theme_bw() +
  labs(x = "x", y = "P(y = 1)") +
  annotate("text", x = 3, y = 0.75, label="0.5", size=8) +
  annotate("text", x = 1.5, y = 0.75, label="1", size=9) +
  annotate("text", x = 0.5, y = 0.75, label="2", size=8) +
  theme(text = element_text(size = 25))
```
]
]

---

### Fitting logistic regression in R

.large[
```{r message=F, warning=F}
library(Stat2Data)
data("MedGPA")

med_glm <- glm(Acceptance ~ GPA, data = MedGPA,
               family = binomial)
```
]

.large[
* `glm` stands for "Generalized Linear Model"
* Use `family = binomial` for binary responses
]

---

### Fitting logistic regression in R

.large[
```{r output.lines = 9:12}
library(Stat2Data)
data("MedGPA")

med_glm <- glm(Acceptance ~ GPA, data = MedGPA,
               family = binomial)
summary(med_glm)
```
]

.large[
$\log \left( \dfrac{\widehat{\pi}}{1 - \widehat{\pi}} \right) = -19.21 + 5.45 \ \text{GPA}$ 
]

---

### Class activity, Part II

.large[
[https://sta112-s22.github.io/class_activities/ca_lecture_29.html](https://sta112-s22.github.io/class_activities/ca_lecture_29.html)
]

---

### Class activity

.large[
$\pi = P(Accepted = 1)$

$\log \left( \dfrac{\widehat{\pi}}{1 - \widehat{\pi}} \right) = -8.712 + 0.246 \ \text{MCAT}$ 

.question[
What is the change in the odds of acceptance associated with an increase of 1 point on the MCAT?
]
]

--

.large[
**Answer:** An increase of 1 point on the MCAT is associated with an increase in the odds of acceptance by a factor of $\exp\{ 0.246 \} = 1.28$.
]

---

### Class activity

.large[
$\pi = P(Accepted = 1)$

$\log \left( \dfrac{\widehat{\pi}}{1 - \widehat{\pi}} \right) = -8.712 + 0.246 \ \text{MCAT}$ 

.question[
What is the estimated probability that a student with an MCAT score of 40 is accepted?
]
]

--

.large[
**Answer:** $\widehat{\pi} = \dfrac{ \exp\{-8.712 + 0.246 \cdot 40 \} }{1 +  \exp\{-8.712 + 0.246 \cdot 40 \}} = 0.76$

So we estimate that a student with an MCAT score of 40 has a 76% chance of being accepted to medical school.
]

---

### Class activity

.large[
$\pi = P(Accepted = 1)$

$\log \left( \dfrac{\widehat{\pi}}{1 - \widehat{\pi}} \right) = -8.712 + 0.246 \ \text{MCAT}$ 

.question[
For approximately what MCAT score would a student have a roughly 50-50 chance of being accepted to medical school?
]
]

--

.large[
**Answer:** We need odds = 1, so $\dfrac{\widehat{\pi}}{1 - \widehat{\pi}} = \exp\{ -8.712 + 0.246 \ \text{MCAT} \} = 1$

So $-8.712 + 0.246 \ \text{MCAT} = 0$, which means $\text{MCAT} = 8.712/0.246 = 35.4$
]