---
title: Outliers and influential points
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

### Agenda

.large[
* Exam 1 next Monday, in class
  * Covers material up through today (inclusive)
  * Closed notes
  * Bring a calculator (cannot use phone or laptop)
  * I won't ask you to write R code, but you may need to interpret R output
* Today:
  * Outliers and influential points
]

---

### Anything unusual?

.large[
Votes for Bush and Buchanan, in Florida counties in 2000:
]

```{r echo=F, message=F, warning=F, fig.width = 7, fig.height=5, fig.align='center', message=F}
library(tidyverse)
library(Stat2Data)
data("PalmBeach")

PalmBeach %>%
  ggplot(aes(x = Bush, y = Buchanan)) +
  geom_point(size = 2.5) +
  labs(x = "Votes for Bush",
       y = "Votes for Buchanan") +
  theme_bw() +
  theme(text= element_text(size=25))
```

.large[
.question[
Does anything look unusual in this scatterplot?
]
]

---

### Anything unusual?

.large[
**Outlier:** A point which doesn't follow the general relationship between the predictor and response.
]

```{r echo=F, message=F, warning=F, fig.width = 7, fig.height=4.5, fig.align='center', message=F}
library(tidyverse)
library(Stat2Data)
data("PalmBeach")

PalmBeach %>%
  ggplot(aes(x = Bush, y = Buchanan)) +
  geom_point(size = 2.5) +
  labs(x = "Votes for Bush",
       y = "Votes for Buchanan") +
  theme_bw() +
  theme(text= element_text(size=25))
```

.large[
.question[
Why might the potential outlier be a problem?
]
]

---

### The impact of outliers

```{r echo=F, fig.width=7, fig.height=5, fig.align='center'}
lm_1 <- lm(Buchanan ~ Bush, data = PalmBeach)
lm_2 <- lm(Buchanan ~ Bush, 
           data = PalmBeach %>%
             filter(County != "PALM BEACH"))

PalmBeach %>%
  mutate(pred_1 = predict(lm_1),
         pred_2 = predict(lm_2, 
                          newdata = PalmBeach)) %>%
  ggplot(aes(x = Bush)) +
  geom_point(aes(y = Buchanan), 
             size = 2.5) +
  geom_line(aes(y = pred_1), color = "blue",
            lwd = 1.2) +
  geom_line(aes(y = pred_2), color = "red",
            lwd = 1.2, lty=2) +
  labs(x = "Votes for Bush",
       y = "Votes for Buchanan") +
  theme_bw() +
  theme(text= element_text(size=25))
```

.large[
* Blue, solid line: potential outlier is included in the data
* Red, dashed line: potential outlier is excluded
]

---

### Influential points

.large[
Outliers which substantially change the fitted line are called **influential.**
]

```{r echo=F, fig.width=7, fig.height=4.5, fig.align='center'}
lm_1 <- lm(Buchanan ~ Bush, data = PalmBeach)
lm_2 <- lm(Buchanan ~ Bush, 
           data = PalmBeach %>%
             filter(County != "PALM BEACH"))

PalmBeach %>%
  mutate(pred_1 = predict(lm_1),
         pred_2 = predict(lm_2, 
                          newdata = PalmBeach)) %>%
  ggplot(aes(x = Bush)) +
  geom_point(aes(y = Buchanan), 
             size = 2.5) +
  geom_line(aes(y = pred_1), color = "blue",
            lwd = 1.2) +
  geom_line(aes(y = pred_2), color = "red",
            lwd = 1.2, lty=2) +
  labs(x = "Votes for Bush",
       y = "Votes for Buchanan") +
  theme_bw() +
  theme(text= element_text(size=25))
```

.large[
.question[
Why would an outlier change the fitted line?
]
]

---

### Impact of influential points

.large[
$\widehat{\beta}_0$ and $\widehat{\beta}_1$ minimize $\sum \limits_{i=1}^n (y_i - \widehat{\beta}_0 - \widehat{\beta}_1 x_i)^2$
]

```{r echo=F, fig.width=7, fig.height=5, fig.align='center'}
lm_1 <- lm(Buchanan ~ Bush, data = PalmBeach)
lm_2 <- lm(Buchanan ~ Bush, 
           data = PalmBeach %>%
             filter(County != "PALM BEACH"))

PalmBeach %>%
  mutate(pred_1 = predict(lm_1),
         pred_2 = predict(lm_2, 
                          newdata = PalmBeach)) %>%
  ggplot(aes(x = Bush)) +
  geom_segment(aes(x = 152846, y = 796.81,
                   xend = 152846, yend = 3407)) +
  geom_point(aes(y = Buchanan), 
             size = 2.5) +
  geom_line(aes(y = pred_1), color = "blue",
            lwd = 1.2) +
  geom_line(aes(y = pred_2), color = "red",
            lwd = 1.2, lty=2) +
  labs(x = "Votes for Bush",
       y = "Votes for Buchanan") +
  theme_bw() +
  theme(text= element_text(size=25))
```

---

### Potential for influence

.large[
.question[
Which point do you think is more influential?
]
]

.pull-left[
```{r echo=F, fig.align='center', fig.width=6, fig.height=4}
set.seed(1)

x  = 1:100
y = 1:100 + rnorm(100, sd=10)

data.frame(x = x,
           y = y) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(size = 2.5) +
  geom_point(aes(x = 50, y = 120), color="red",
             size = 2.5) +
  theme_bw() +
  theme(text = element_text(size = 25))
```
]

.pull-right[
```{r echo=F, fig.align='center', fig.width=6, fig.height=4}
set.seed(1)

x  = 1:100
y = 1:100 + rnorm(100, sd=10)

data.frame(x = x,
           y = y) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(size = 2.5) +
  geom_point(aes(x = 100, y = 30), color="red",
             size = 2.5) +
  theme_bw() +
  theme(text = element_text(size = 25))
```
]

.large[
.abox[
(A) The one on the left
]

.bbox[
(B) The one on the right
]
]

---

### Potential for influence

.large[
Influence depends on how far $x$ is from $\overline{x}$.
]

.pull-left[
<ins> Some influence </ins>
```{r echo=F, fig.align='center', fig.width=6, fig.height=4}
set.seed(1)

x  = 1:100
y = 1:100 + rnorm(100, sd=10)

data.frame(x = x,
           y = y) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(size = 2.5) +
  geom_point(aes(x = 50, y = 120), color="red",
             size = 2.5) +
  theme_bw() +
  theme(text = element_text(size = 25))
```
]

.pull-right[
<ins> More influence </ins>
```{r echo=F, fig.align='center', fig.width=6, fig.height=4}
set.seed(1)

x  = 1:100
y = 1:100 + rnorm(100, sd=10)

data.frame(x = x,
           y = y) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(size = 2.5) +
  geom_point(aes(x = 100, y = 30), color="red",
             size = 2.5) +
  theme_bw() +
  theme(text = element_text(size = 25))
```
]

---

### Influential points

.large[
.question[
Which point do you think is more influential?
]
]

.pull-left[
```{r echo=F, fig.align='center', fig.width=6, fig.height=4}
set.seed(1)

x  = 1:100
y = 1:100 + rnorm(100, sd=10)

data.frame(x = x,
           y = y) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(size = 2.5) +
  geom_point(aes(x = 50, y = 120), color="red",
             size = 2.5) +
  theme_bw() +
  theme(text = element_text(size = 25))
```
]

.pull-right[
```{r echo=F, fig.align='center', fig.width=6, fig.height=4}
set.seed(1)

x  = 1:100
y = 1:100 + rnorm(100, sd=10)

data.frame(x = x,
           y = y) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(size = 2.5) +
  geom_point(aes(x = 100, y = 100), color="red",
             size = 2.5) +
  theme_bw() +
  theme(text = element_text(size = 25))
```
]

.large[
.abox[
(A) The one on the left
]

.bbox[
(B) The one on the right
]
]

---

### Influential points

.large[
Influence depends on both $x$ and $y$.
]

.pull-left[
<ins> Some influence </ins>
```{r echo=F, fig.align='center', fig.width=6, fig.height=4}
set.seed(1)

x  = 1:100
y = 1:100 + rnorm(100, sd=10)

data.frame(x = x,
           y = y) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(size = 2.5) +
  geom_point(aes(x = 50, y = 120), color="red",
             size = 2.5) +
  theme_bw() +
  theme(text = element_text(size = 25))
```
]

.pull-right[
<ins> Not influential </ins>
```{r echo=F, fig.align='center', fig.width=6, fig.height=4}
set.seed(1)

x  = 1:100
y = 1:100 + rnorm(100, sd=10)

data.frame(x = x,
           y = y) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(size = 2.5) +
  geom_point(aes(x = 100, y = 100), color="red",
             size = 2.5) +
  theme_bw() +
  theme(text = element_text(size = 25))
```
]

---

### Identifying outliers

.large[
The normality assumption can be used to help identify outliers.

.question[
Suppose the residuals follow a $N(0, 1)$ distribution. What values are "unusual" for a $N(0, 1)$ distribution?
]
]

--

.large[
Values $< -2$ or $> 2$ are somewhat unusual. Values $< -3$ or $> 3$ are very unusual.
]

---

### Studentized residuals

.large[
**Studentized residuals:** $\dfrac{y_i - \widehat{y}_i}{\widehat{\sigma}_{\varepsilon}}$

where $\widehat{\sigma}_{\varepsilon}$ is computed without observation $i$. 

* Under the normality assumption, studentized residuals should approximately follow a $N(0, 1)$ distribution.
]

---

### Studentized residual plot

.large[
**Studentized residuals:** $\dfrac{y_i - \widehat{y}_i}{\widehat{\sigma}_{\varepsilon}}$

**Studentized residual plot:** plot studentized residuals on vertical axis, and $\widehat{y}$ on horizontal axis. Add lines at -3, -2, 2, and 3.
]

```{r echo=F, fig.width=7, fig.height=4.5, fig.align='center'}
vote_lm <- lm(Buchanan ~ Bush, data = PalmBeach)

PalmBeach %>%
  mutate(studentized_resids = rstudent(vote_lm),
         yhat = predict(vote_lm)) %>%
  ggplot(aes(x = yhat, y = studentized_resids)) +
  geom_point(size = 2.5) +
  geom_abline(slope = 0, intercept = 0, 
              color = "blue", lwd=1.2) +
  geom_abline(slope = 0, intercept = c(-3, -2, 2, 3),
              color = "orange", lty = 2, lwd=1.2) +
  labs(x = "Predicted votes for Buchanan",
       y = "Studentized residuals") +
  theme_bw() +
  theme(text = element_text(size = 25))
```

---

### Studentized residual plot

.large[
**Studentized residual plot:** plot studentized residuals on vertical axis, and $\widehat{y}$ on horizontal axis. Add lines at -3, -2, 2, and 3.
]

```{r echo=F, fig.width=7, fig.height=4.5, fig.align='center'}
vote_lm <- lm(Buchanan ~ Bush, data = PalmBeach)

PalmBeach %>%
  mutate(studentized_resids = rstudent(vote_lm),
         yhat = predict(vote_lm)) %>%
  ggplot(aes(x = yhat, y = studentized_resids)) +
  geom_point(size = 2.5) +
  geom_abline(slope = 0, intercept = 0, 
              color = "blue", lwd=1.2) +
  geom_abline(slope = 0, intercept = c(-3, -2, 2, 3),
              color = "orange", lty = 2, lwd=1.2) +
  labs(x = "Predicted votes for Buchanan",
       y = "Studentized residuals") +
  theme_bw() +
  theme(text = element_text(size = 25))
```

.large[
Points $< -3$ or $> 3$ are suspected outliers.
]

---

### Caution: reliance on normality

.large[
A studentized residual plot relies on the normality assumption. If the normality assumption doesn't hold, points $< -3$ or $> 3$ might not be outliers.
]

.pull-left[
```{r, echo=F, message=F, fig.width=6, fig.height=5.5, fig.align='center'}
data("CountyHealth")
health_lm <- lm(MDs ~ Hospitals, data = CountyHealth)
CountyHealth %>%
  mutate(studentized_resids = rstudent(health_lm),
         yhat = predict(health_lm)) %>%
  ggplot(aes(x = yhat, y = studentized_resids)) +
  geom_point(size = 2.5) +
  geom_abline(slope = 0, intercept = 0, 
              color = "blue", lwd=1.2) +
  geom_abline(slope = 0, intercept = c(-3, -2, 2, 3),
              color = "orange", lty = 2, lwd=1.2) +
  labs(x = "Predicted number of doctors",
       y = "Studentized residuals",
       title = "Studentized residual plot \n (before transforming)") +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  scale_y_continuous(limits = c(-4, 5))
```
]

.pull-right[
```{r, echo=F, message=F, fig.width=6, fig.height=5.5, fig.align='center'}
health_lm <- lm(sqrt(MDs) ~ Hospitals, data = CountyHealth)
CountyHealth %>%
  mutate(studentized_resids = rstudent(health_lm),
         yhat = predict(health_lm)) %>%
  ggplot(aes(x = yhat, y = studentized_resids)) +
  geom_point(size = 2.5) +
  geom_abline(slope = 0, intercept = 0, 
              color = "blue", lwd=1.2) +
  geom_abline(slope = 0, intercept = c(-3, -2, 2, 3),
              color = "orange", lty = 2, lwd=1.2) +
  labs(x = "Predicted sqrt(number of doctors)",
       y = "Studentized residuals",
       title = "Studentized residual plot \n (after transforming)") +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  scale_y_continuous(limits = c(-4, 5))
```
]

---

### Handling outliers

.large[
.question[
Suppose we identify a suspected outlier. What should we do about it?
]
]

--

.large[
If the observation is clearly a measurement error:
* Remove it

Otherwise:
* Fit line with and without point, see how the estimated line changes
* If the estimated line changes substantially, report both fits
]

---

### Class activity

.large[
[https://sta112-s22.github.io/class_activities/ca_lecture_13.html](https://sta112-s22.github.io/class_activities/ca_lecture_13.html)
]

---

### Class activity

```{r echo=F, message=F, fig.width=7, fig.height=4.5, fig.align='center'}
library(gapminder)

gapminder_2007 <- gapminder %>%
  filter(year == 2007) %>%
  mutate(log_gdp_percap = log(gdpPercap))
  
gap_lm <- lm(lifeExp ~ log_gdp_percap, data = gapminder_2007)

studentized_residual_plot <- function(fitted_model, data){
  data %>%
    mutate(studentized_resids = rstudent(fitted_model),
           yhat = predict(fitted_model)) %>%
    ggplot(aes(x = yhat, y = studentized_resids)) +
    geom_point(size = 2.5) +
    geom_abline(slope = 0, intercept = 0, 
                color = "blue", lwd=1.2) +
    geom_abline(slope = 0, intercept = c(-3, -2, 2, 3),
                color = "orange", lty = 2, lwd=1.2) +
    labs(x = "Predicted response",
         y = "Studentized residuals") +
    theme_bw() +
    theme(text = element_text(size = 20))
}

studentized_residual_plot(gap_lm, gapminder_2007)
```

.large[
.question[
Are there any suspected outliers?
]
]

--

.large[
Yes -- there are a few points with studentized residuals $< -3$
]

---

### Class activity

```{r echo=F, message=F, fig.width=7, fig.height=4.5, fig.align='center'}
gapminder_2007_small <- gapminder_2007 %>%
  mutate(studentized_resids = rstudent(gap_lm)) %>%
  filter(studentized_resids > -3, studentized_resids < 3)

gapminder_2007 %>%
  ggplot(aes(x = log_gdp_percap, y = lifeExp)) +
  geom_point() +
  geom_smooth(se=F, method="lm", color="blue") +
  geom_smooth(data = gapminder_2007_small, 
              se=F, method="lm", color = "red", lty = 2) +
  labs(x = "log(GDP per capita)", 
       y = "Life expectancy") +
  theme_bw()
```

.large[
.question[
Do the outliers seem to substantially change the fitted line?
]
]

--

.large[
Not really -- the fitted lines are pretty close
]

---

### Class activity

```{r echo=F, message=F, fig.width=7, fig.height=4.5, fig.align='center'}

gapminder_2007 %>%
  ggplot(aes(x = log_gdp_percap, y = lifeExp)) +
  geom_point() +
  geom_smooth(se=F, method="lm", color="blue") +
  geom_smooth(data = gapminder_2007_small, 
              se=F, method="lm", color = "red", lty = 2) +
  labs(x = "log(GDP per capita)", 
       y = "Life expectancy") +
  theme_bw()
```

.large[
.question[
What should we do about the suspected outliers?
]
]

--

.large[
Seems reasonable to leave them in the data.
]