---
title: Regression and hypothesis testing with a categorical predictor
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

### Agenda

.large[
* Recap: regression with a categorical predictor
* Partitioning variance
* Hypothesis testing
]
    
---

### Categorical predictor

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=6, fig.height=4}
library(palmerpenguins)
library(tidyverse)
library(latex2exp)
library(Stat2Data)

penguins_no_nas <- penguins %>%
  drop_na()

penguins_no_nas %>%
  ggplot(aes(x = species, y = bill_length_mm)) +
  geom_boxplot(lwd=1.5) +
  theme_classic() +
  labs(x = "Species", 
       y = "Bill length (mm)") +
  theme(text = element_text(size = 25))
```

.large[
* How do we model the relationship between species and bill length?
* Can we test for a relationship between bill length and species?
]

---

### Categorical predictor

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=6, fig.height=4}

set.seed(1)
penguins_no_nas %>%
  ggplot(aes(x = species, y = bill_length_mm)) +
  geom_jitter(alpha = 0) +
  theme_classic() +
  geom_segment(aes(x = 0.6, y = 38, xend = 1.4, yend = 38), lwd=1.5, color="blue") + 
  geom_segment(aes(x = 1.6, y = 50, xend = 2.4, yend = 50), lwd=1.5, color="blue") + 
  geom_segment(aes(x = 2.6, y = 47, xend = 3.4, yend = 47), lwd=1.5, color="blue") + 
  labs(x = "Species", 
       y = "Bill length (mm)") +
  theme(text = element_text(size = 25)) +
  theme(axis.title.y = element_blank()) +
  scale_y_continuous(breaks = c(38, 50, 47),
                     labels = c(TeX("$\\beta_0"), 
                                TeX("$\\beta_0 + \\beta_1"), 
                                TeX("$\\beta_0 + \\beta_2"))) +
  geom_segment(aes(x = 1.45, y = 38,
                   xend = 1.45, yend = 50),
               arrow = arrow(length = unit(0.3, "cm"))) +
  geom_segment(aes(x = 2.45, y = 38,
                   xend = 2.45, yend = 47),
               arrow = arrow(length = unit(0.3, "cm"))) +
  annotate("text", x = 1.3, y = 44, 
           label = TeX("$\\beta_1", output='character'), 
           parse=TRUE, size = 7) +
  annotate("text", x = 2.3, y = 43, 
           label = TeX("$\\beta_2", output='character'), 
           parse=TRUE, size = 7)
```

.large[
$$\text{bill length} = \begin{cases} \beta_0 + \varepsilon & \text{species = Adelie} \\ \beta_0 + \beta_1 + \varepsilon & \text{species = Chinstrap} \\ \beta_0 + \beta_2 + \varepsilon & \text{species = Gentoo} \end{cases}$$
]

---

### Indicator variables

.large[
$$\text{IsChinstrap} = \begin{cases} 0 & \text{species } \neq \text{Chinstrap} \\ 1 & \text{species = Chinstrap} \end{cases}$$

$$\text{IsGentoo} = \begin{cases} 0 & \text{species } \neq \text{Gentoo} \\ 1 & \text{species = Gentoo} \end{cases}$$

Then:

| Species | IsChinstrap | IsGentoo |
| --- | --- | --- |
| Adelie | 0 | 0 |
| Chinstrap | 1 | 0 |
| Gentoo | 0 | 1 |
]

---

### Indicator variables

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=6, fig.height=3.5}

set.seed(1)
penguins_no_nas %>%
  ggplot(aes(x = species, y = bill_length_mm)) +
  geom_jitter(alpha = 0) +
  theme_classic() +
  geom_segment(aes(x = 0.6, y = 38, xend = 1.4, yend = 38), lwd=1.5, color="blue") + 
  geom_segment(aes(x = 1.6, y = 50, xend = 2.4, yend = 50), lwd=1.5, color="blue") + 
  geom_segment(aes(x = 2.6, y = 47, xend = 3.4, yend = 47), lwd=1.5, color="blue") + 
  labs(x = "Species", 
       y = "Bill length (mm)") +
  theme(text = element_text(size = 25)) +
  theme(axis.title.y = element_blank()) +
  scale_y_continuous(breaks = c(38, 50, 47),
                     labels = c(TeX("$\\beta_0"), 
                                TeX("$\\beta_0 + \\beta_1"), 
                                TeX("$\\beta_0 + \\beta_2"))) +
  geom_segment(aes(x = 1.45, y = 38,
                   xend = 1.45, yend = 50),
               arrow = arrow(length = unit(0.3, "cm"))) +
  geom_segment(aes(x = 2.45, y = 38,
                   xend = 2.45, yend = 47),
               arrow = arrow(length = unit(0.3, "cm"))) +
  annotate("text", x = 1.3, y = 44, 
           label = TeX("$\\beta_1", output='character'), 
           parse=TRUE, size = 7) +
  annotate("text", x = 2.3, y = 43, 
           label = TeX("$\\beta_2", output='character'), 
           parse=TRUE, size = 7)
```

.large[
$$\text{bill length} = \begin{cases} \beta_0 + \varepsilon & \text{species = Adelie} \\ \beta_0 + \beta_1 + \varepsilon & \text{species = Chinstrap} \\ \beta_0 + \beta_2 + \varepsilon & \text{species = Gentoo} \end{cases}$$

Or, more concisely:

$$\text{bill length} = \beta_0 + \beta_1 \text{IsChinstrap} + \beta_2 \text{IsGentoo} + \varepsilon$$
]

---

### Diamonds Data

.large[
Data on sales price and characteristics for 351 diamonds from an online gem retailer:
]

```{r echo=F, fig.align='center', fig.width=7, fig.height=5}
data("Diamonds")
Diamonds %>%
  ggplot(aes(x = Clarity, y = PricePerCt)) +
  geom_boxplot() +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  labs(x = "Clarity", 
       y = "Price per carat ($)")
``` 

---

### Diamonds Data

```{r echo=F, fig.align='center', fig.width=7, fig.height=4.5}
data("Diamonds")
Diamonds %>%
  ggplot(aes(x = Clarity, y = PricePerCt)) +
  geom_boxplot() +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  labs(x = "Clarity", 
       y = "Price per carat ($)")
``` 

.large[
.question[
There are 8 levels of diamond clarity in the data. How many indicator variables do we need to represent them?
]
]

--

.large[
$7 \ (= 8 - 1)$
]

---

### Diamonds Data


```{r echo=F, fig.align='center', fig.width=7, fig.height=4.5}
data("Diamonds")
Diamonds %>%
  ggplot(aes(x = Clarity, y = PricePerCt)) +
  geom_boxplot() +
  theme_bw() +
  theme(text = element_text(size = 25)) +
  labs(x = "Clarity", 
       y = "Price per carat ($)")
``` 

.large[
.question[
How many parameters are in the model?
]
]

--

.large[
8 (intercept + 1 for each indicator variable)
]

---

### Class activity, Part I

.large[
\begin{align}
\text{Price per carat} &= \beta_0 + \beta_1 \ \text{IsSI1} + \beta_2 \ \text{IsSI2} + \beta_3 \ \text{IsSI3} + \beta_4 \ \text{IsVS1} \\
& \hspace{1cm} + \beta_5 \ \text{IsVS2} + \beta_6 \ \text{IsVVS1} + \beta_7 \ \text{IsVVS2} + \varepsilon
\end{align}
]

---

### Class activity

.large[
.question[
What is the estimated mean price for a diamond of clarity level SI3?
]
]

--

.large[
6630.53 - 3714.53 = $2916
]

---

### Class activity

.large[
.question[
What is the $R^2$ and $R^2_{adj}$ for the fitted model?
]
]

--

.large[
$R^2 = 0.03696$ 

$R^2_{adj} = 0.0173$
]

---

### Class activity

.large[
.question[
Why is $R^2_{adj} < R^2$?
]
]

--

.large[
Because we add a penalty term for the number of parameters ( $p$ ) in the model (in this case, 8).

$R^2 = 1 - \dfrac{SSE}{SSTotal}$

$R^2_{adj} = 1 - \dfrac{SSE/(n - p)}{SSTotal/(n - 1)}$
]

---

### Partitioning variability and degrees of freedom

.large[
**Variability:** $\hspace{3cm} \sum \limits_{i=1}^n (y_i - \overline{y})^2 = \sum \limits_{i=1}^n (\widehat{y}_i - \overline{y})^2 + \sum \limits_{i=1}^n (y_i - \widehat{y}_i)^2$

$\hspace{3cm} SSTotal \hspace{0.5cm} = \hspace{0.5cm} SSModel \hspace{0.5cm} + \hspace{0.5cm} SSE$

**Degrees of freedom:** 

$\hspace{4cm} n-1 \hspace{0.5cm} = \hspace{0.5cm} (p - 1)  \hspace{0.5cm} + \hspace{0.5cm} (n - p)$
]

---

### Hypothesis testing

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=6, fig.height=4}
library(palmerpenguins)
library(tidyverse)

penguins_no_nas <- penguins %>%
  drop_na()

penguins_no_nas %>%
  ggplot(aes(x = species, y = bill_length_mm)) +
  geom_boxplot() +
  theme_classic() +
  labs(x = "Species", 
       y = "Bill length (mm)") +
  theme(text = element_text(size = 25))
```

.large[
Is there a relationship between species and bill length?
]

---

### Intuition

.large[
More evidence for a relationship when *between-group* variability is larger than *within-group* variability.
]

.pull-left[
```{r echo=F, message=F, warning=F, fig.align='center', fig.width=6, fig.height=4}
penguins_no_nas %>%
  mutate(bill_length_mm = bill_length_mm + 
           ifelse(species == "Adelie", 0, 
                  ifelse(species == "Chinstrap", 10, 5))) %>%
  ggplot(aes(x = species, y = bill_length_mm)) +
  geom_boxplot(lwd=1) +
  theme_classic() +
  labs(x = "Species", 
       y = "Bill length (mm)") +
  theme(text = element_text(size = 25),
        axis.text.y = element_blank())
```

.large[
Large between-group variability, relative to within-group variability.
]
]

.pull-right[
```{r echo=F, message=F, warning=F, fig.align='center', fig.width=6, fig.height=4}
penguins_no_nas %>%
  mutate(bill_length_mm = bill_length_mm + 
           rnorm(nrow(penguins_no_nas), sd=20)) %>%
  ggplot(aes(x = species, y = bill_length_mm)) +
  geom_boxplot(lwd=1) +
  theme_classic() +
  labs(x = "Species", 
       y = "Bill length (mm)") +
  theme(text = element_text(size = 25),
        axis.text.y = element_blank())
```

.large[
Small between-group variability, relative to within-group variability.
]
]

---

### Between-group vs. within-group variability

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4.5}
penguins_no_nas %>%
  ggplot(aes(x = species, y = bill_length_mm)) +
  geom_boxplot(lwd=1) +
  theme_classic() +
  labs(x = "Species", 
       y = "Bill length (mm)") +
  theme(text = element_text(size = 25))
```

.large[
**Between-group variance:** $\frac{1}{p-1} \sum \limits_{i=1}^n (\widehat{y}_i - \overline{y})^2 = \dfrac{SSModel}{p-1}$ 

**Within-group variance:** $\frac{1}{n-p} \sum \limits_{i=1}^n (y_i - \widehat{y}_i)^2 = \dfrac{SSE}{n-p}$ 
]

---

### Between-group vs. within-group variability

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4.5}
penguins_no_nas %>%
  ggplot(aes(x = species, y = bill_length_mm)) +
  geom_boxplot(lwd=1) +
  theme_classic() +
  labs(x = "Species", 
       y = "Bill length (mm)") +
  theme(text = element_text(size = 25))
```

.large[
**Test statistic:** $F = \dfrac{\text{between group variance}}{\text{within group variance}} = \dfrac{SSModel/(p-1)}{SSE/(n-p)}$
]

---

### Partitioning variability

.large[
**Variability:** $\hspace{3cm} \sum \limits_{i=1}^n (y_i - \overline{y})^2 = \sum \limits_{i=1}^n (\widehat{y}_i - \overline{y})^2 + \sum \limits_{i=1}^n (y_i - \widehat{y}_i)^2$

$\hspace{3cm} SSTotal \hspace{0.5cm} = \hspace{0.5cm} SSModel \hspace{0.5cm} + \hspace{0.5cm} SSE$


**Test statistic:** $F = \dfrac{\frac{1}{p-1} \sum \limits_{i=1}^n (\widehat{y}_i - \overline{y})^2}{\frac{1}{n-p} \sum \limits_{i=1}^n (y_i - \widehat{y}_i)^2} = \dfrac{\frac{1}{p-1} SSModel}{\frac{1}{n-p} SSE} = \dfrac{MSModel}{MSE}$ 
]

---

### Analysis of variance

.large[

**Test statistic:** $F = \dfrac{\frac{1}{p-1} \sum \limits_{i=1}^n (\widehat{y}_i - \overline{y})^2}{\frac{1}{n-p} \sum \limits_{i=1}^n (y_i - \widehat{y}_i)^2} = \dfrac{\frac{1}{p-1} SSModel}{\frac{1}{n-p} SSE} = \dfrac{MSModel}{MSE}$ 

**Analysis of variance (ANOVA) table:**

| Source | df | SS | MS | F |
| --- | --- | --- | --- | --- |
| Model | $p-1$ | $SSModel$ | $MSModel = \frac{SSModel}{p-1}$ | $F = \frac{MSModel}{MSE}$ |
| Residual | $n - p$ | $SSE$ | $MSE = \frac{SSE}{n-p}$ | |
| Total | $n-1$ | $SSTotal$ | | |
]

---

### Example

.large[
$$\widehat{\text{bill length}} = 38.82 + 10.01 \ \text{IsChinstrap} + 8.74 \ \text{IsGentoo}$$
]

.large[

$$\sum \limits_{i=1}^n (\widehat{\text{bill length}}_i - \overline{\text{bill length}})^2 = 7015.4 \hspace{1cm}$$ 

$$\sum \limits_{i=1}^n (\text{bill length}_i - \widehat{\text{bill length}}_i)^2 = 2913.5 \hspace{1cm}$$
]

---

### Example

.large[

$$\sum \limits_{i=1}^n (\widehat{\text{bill length}}_i - \overline{\text{bill length}})^2 = 7015.4 \hspace{1cm}$$ 

$$\sum \limits_{i=1}^n (\text{bill length}_i - \widehat{\text{bill length}}_i)^2 = 2913.5 \hspace{1cm}$$

.question[
There are 333 penguins in the dataset used to fit the model. Calculate all the numbers for the ANOVA table:
]

| Source | df | SS | MS | F |
| --- | --- | --- | --- | --- |
| Model | |  | | |
| Residual | |  |  | |
| Total |  | | | |
]

---

### Example

.large[

$$\sum \limits_{i=1}^n (\widehat{\text{bill length}}_i - \overline{\text{bill length}})^2 = 7015.4 \hspace{1cm}$$ 

$$\sum \limits_{i=1}^n (\text{bill length}_i - \widehat{\text{bill length}}_i)^2 = 2913.5 \hspace{1cm}$$

| Source | df | SS | MS | F |
| --- | --- | --- | --- | --- |
| Model | 2 | 7015.4 | 3507.7 | 397.3 |
| Residual | 330 | 2913.5 | 8.8 | |
| Total | 332  | 9928.9 | | |

]

---

### Class activity, Part II

---

### Class activity

.large[
.question[
Fill in the ANOVA table for the Diamonds data:
]
]

--

.large[

| Source | df | SS | MS | F |
| --- | --- | --- | --- | --- |
| Model | 7 | 108.44 | 15.49 | 1.88 |
| Residual | 343 | 2825.74 | 8.24 | |
| Total | 350  | 2934.18 | | |

]