---
title: Introduction to multiple linear regression
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

### Agenda

---

### Recap: multiple linear regression

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4.5}

library(palmerpenguins)
library(tidyverse)
library(latex2exp)
library(Stat2Data)
library(knitr)

hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

penguins_no_nas <- penguins %>%
  drop_na()

length_lm <- lm(bill_length_mm ~ species + body_mass_g, 
                data = penguins_no_nas)

penguins_no_nas %>%
  mutate(yhat = predict(length_lm)) %>%
  ggplot(aes(x = body_mass_g,
             y = bill_length_mm,
             color = species)) +
  geom_point(size = 2, alpha = 0.5) +
  geom_line(aes(y = yhat), 
            lwd = 1.5) +
  labs(x = "Body mass (g)",
       y = "Bill length (mm)",
       color = "Species") +
  theme_bw() +
  theme(text = element_text(size = 25))
```


.large[
**Model:**

\begin{align}
\text{bill length} &= \beta_0 + \beta_1 \text{IsChinstrap} + \beta_2 \text{IsGentoo} \\
& \hspace{0.5cm} + \beta_3 \text{body mass} + \varepsilon
\end{align}
]

---

### Recap: multiple linear regression

.large[
\begin{align}
\text{bill length} &= \beta_0 + \beta_1 \text{IsChinstrap} + \beta_2 \text{IsGentoo} \\
& \hspace{0.5cm} + \beta_3 \text{body mass} + \varepsilon
\end{align}

With one categorical and one quantitative predictor, we get a separate regression line for each species of penguin:

$$\text{bill length} = \begin{cases} \beta_0 + \beta_3 \text{body mass} +  \varepsilon & \text{species = Adelie} \\ \beta_0 + \beta_1 + \beta_3 \text{body mass} + \varepsilon & \text{species = Chinstrap} \\ \beta_0 + \beta_2 + \beta_3 \text{body mass} +  \varepsilon & \text{species = Gentoo} \end{cases}$$

]

---

### Two quantitative predictors

.large[
What if I want to fit a regression model with two *quantitative* predictors, body mass and flipper length?

.question[
What would the population regression model look like?
]
]

--

.large[
$$\text{bill length} = \beta_0 + \beta_1 \text{body mass} + \beta_2 \text{flipper length} + \varepsilon$$
]

---

### Fitting the model

.large[
```{r, output.lines=10:14}
m1 <- lm(bill_length_mm ~ body_mass_g + 
           flipper_length_mm,
         data = penguins_no_nas)
summary(m1)
```

$$\widehat{\text{bill length}} = -3.98 + 0.00055 \ \text{body mass} + 0.227 \ \text{flipper length}$$
]

---

### Interpreting coefficients

.large[
$$\widehat{\text{bill length}} = -3.98 + 0.00055 \ \text{body mass} + 0.227 \ \text{flipper length}$$

.question[
How do I interpret the estimated intercept, $\widehat{\beta}_0 = -3.98$?
]
]

--

.large[
-3.98 is the estimated bill length when both body mass = 0 and flipper length = 0 (not that meaningful)
]

---

### Interpreting coefficients

.large[
$$\widehat{\text{bill length}} = -3.98 + 0.00055 \ \text{body mass} + 0.227 \ \text{flipper length}$$
As with a categorical predictor, we get a separate regression line for each flipper length.

.question[
Suppose a penguin has a flipper length of 200mm. What is the estimated regression line for the relationship between body mass and bill length?
]
]

--

.large[
The estimated regression line is

\begin{align}
\widehat{\text{bill length}} &= -3.98 + 0.00055 \ \text{body mass} + 0.227(200) \\
&= -3.98 + 45.4 + 0.00055 \ \text{body mass} \\
&= 41.42 + 0.00055 \ \text{body mass}
\end{align}
]

---

### Interpreting coefficients

.large[
$$\widehat{\text{bill length}} = -3.98 + 0.00055 \ \text{body mass} + 0.227 \ \text{flipper length}$$
]

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=5}
penguins_no_nas %>%
  ggplot(aes(x = body_mass_g, y = bill_length_mm)) +
  geom_point(alpha = 0) +
  geom_abline(intercept = 41.42, slope = 0.00055,
              color = "blue", lwd=1.2) +
  geom_abline(intercept = 48.23, slope = 0.00055,
              color = "blue", lwd=1.2) +
  labs(x = "Body mass (g)", y = "Bill length (mm)") +
  theme_bw() +
  theme(text = element_text(size=25)) +
  annotate("text", x = 5300, y = 41, 
           label = c("Flipper length = 200mm"), 
           parse=F, size = 7) +
annotate("text", x = 5300, y = 55, 
           label = c("Flipper length = 230mm"), 
           parse=F, size = 7)
```

.large[
.question[
Are longer flippers associated with longer or shorter bills?
]
]

---

### Interpreting coefficients

.large[
$$\widehat{\text{bill length}} = -3.98 + 0.00055 \ \text{body mass} + 0.227 \ \text{flipper length}$$
]

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=5}
penguins_no_nas %>%
  ggplot(aes(x = body_mass_g, y = bill_length_mm)) +
  geom_point(alpha = 0) +
  geom_abline(intercept = 41.42, slope = 0.00055,
              color = "blue", lwd=1.2) +
  geom_abline(intercept = 48.23, slope = 0.00055,
              color = "blue", lwd=1.2) +
  labs(x = "Body mass (g)", y = "Bill length (mm)") +
  theme_bw() +
  theme(text = element_text(size=25)) +
  annotate("text", x = 5300, y = 41, 
           label = c("Flipper length = 200mm"), 
           parse=F, size = 7) +
  annotate("text", x = 5300, y = 55, 
           label = c("Flipper length = 230mm"), 
           parse=F, size = 7) +
  geom_segment(aes(x = 4000, y = 44.5,
                   xend = 4000, yend = 49.5),
               arrow = arrow(length = unit(0.3, "cm")))
```

.large[
For a fixed body mass, penguins with longer flippers tend to have longer bills.
]

---

### Interpreting coefficients

.large[
$$\widehat{\text{bill length}} = -3.98 + 0.00055 \ \text{body mass} + 0.227 \ \text{flipper length}$$
]

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=5}
penguins_no_nas %>%
  ggplot(aes(x = body_mass_g, y = bill_length_mm)) +
  geom_point(alpha = 0) +
  geom_abline(intercept = 41.42, slope = 0.00055,
              color = "blue", lwd=1.2) +
  geom_abline(intercept = 48.23, slope = 0.00055,
              color = "blue", lwd=1.2) +
  labs(x = "Body mass (g)", y = "Bill length (mm)") +
  theme_bw() +
  theme(text = element_text(size=25)) +
  annotate("text", x = 5300, y = 41, 
           label = c("Flipper length = 200mm"), 
           parse=F, size = 7) +
  annotate("text", x = 5300, y = 55, 
           label = c("Flipper length = 230mm"), 
           parse=F, size = 7) +
  geom_segment(aes(x = 4000, y = 44.5,
                   xend = 4000, yend = 49.5),
               arrow = arrow(length = unit(0.3, "cm")))
```

.large[
.question[
What does $\widehat{\beta}_2 = 0.227$ mean?
]
]

---

### Interpreting coefficients

.large[
$$\widehat{\text{bill length}} = -3.98 + 0.00055 \ \text{body mass} + 0.227 \ \text{flipper length}$$
]

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=5}
penguins_no_nas %>%
  ggplot(aes(x = body_mass_g, y = bill_length_mm)) +
  geom_point(alpha = 0) +
  geom_abline(intercept = 41.42, slope = 0.00055,
              color = "blue", lwd=1.2) +
  geom_abline(intercept = 41.647, slope = 0.00055,
              color = "blue", lwd=1.2) +
  labs(x = "Body mass (g)", y = "Bill length (mm)") +
  theme_bw() +
  theme(text = element_text(size=25)) +
  annotate("text", x = 5300, y = 43.7, 
           label = c("Flipper length = 200mm"), 
           parse=F, size = 7) +
  annotate("text", x = 4900, y =45, 
           label = c("Flipper length = 201mm"), 
           parse=F, size = 7) +
  annotate("text", x = 4000, y =43.5, 
           label = "0.227", 
           parse=F, size = 7) +
  geom_segment(aes(x = 4000, y = 43.6,
                   xend = 4000, yend = 43.8),
               arrow = arrow(length = unit(0.3, "cm"))) +
  scale_y_continuous(limits=c(42.7, 45.2))
```

.large[
For a fixed body mass, and increase of 1mm in flipper length is associated with an increase of 0.227mm in bill length.
]

---

### Interpreting coefficients

.large[
$$\widehat{\text{bill length}} = -3.98 + 0.00055 \ \text{body mass} + 0.227 \ \text{flipper length}$$
]

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=5}
penguins_no_nas %>%
  ggplot(aes(x = body_mass_g, y = bill_length_mm)) +
  geom_point(alpha = 0) +
  geom_abline(intercept = 41.42, slope = 0.00055,
              color = "blue", lwd=1.2) +
  geom_abline(intercept = 41.647, slope = 0.00055,
              color = "blue", lwd=1.2) +
  labs(x = "Body mass (g)", y = "Bill length (mm)") +
  theme_bw() +
  theme(text = element_text(size=25)) +
  annotate("text", x = 5300, y = 43.7, 
           label = c("Flipper length = 200mm"), 
           parse=F, size = 7) +
  annotate("text", x = 4900, y =45, 
           label = c("Flipper length = 201mm"), 
           parse=F, size = 7) +
  scale_y_continuous(limits=c(42.7, 45.2))
```

.large[
.question[
What does $\widehat{\beta}_1 = 0.00055$ mean?
]
]

---

### Interpreting coefficients

.large[
$$\widehat{\text{bill length}} = -3.98 + 0.00055 \ \text{body mass} + 0.227 \ \text{flipper length}$$
]

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=5}
penguins_no_nas %>%
  ggplot(aes(x = body_mass_g, y = bill_length_mm)) +
  geom_point(alpha = 0) +
  geom_abline(intercept = 41.42, slope = 0.00055,
              color = "blue", lwd=1.2) +
  geom_abline(intercept = 41.647, slope = 0.00055,
              color = "blue", lwd=1.2) +
  labs(x = "Body mass (g)", y = "Bill length (mm)") +
  theme_bw() +
  theme(text = element_text(size=25)) +
  annotate("text", x = 5300, y = 43.7, 
           label = c("Flipper length = 200mm"), 
           parse=F, size = 7) +
  annotate("text", x = 4900, y =45, 
           label = c("Flipper length = 201mm"), 
           parse=F, size = 7) +
  scale_y_continuous(limits=c(42.7, 45.2))
```

.large[
For a fixed flipper length, an increase of 1g in body mass is associated with an increase of 0.00055mm in bill length.
]

---

### Making predictions

.large[
$$\widehat{\text{bill length}} = -3.98 + 0.00055 \ \text{body mass} + 0.227 \ \text{flipper length}$$

.question[
What is the estimated bill length for a penguin with a body mass of 5000g and a flipper length of 200mm?
]

.abox[
(A) 48.15
]

.bbox[
(B) 41.42
]

.cbox[
(C) 44.17
]
]

--

.large[
$-3.98 + 0.00055(5000) + 0.227(200) = 44.17$
]

---

### Class activity

---

### Class activity

.large[
.question[
What is the equation of the estimated regression line?
]
]

--

.large[
$\widehat{\text{GPA}} = 0.635 + 0.498 \text{HSGPA} + 0.0012 \text{SATV}$
]

---

### Class activity

.large[
.question[
What is the estimated college GPA for a student with a high school GPA of 3.5, and a verbal SAT score of 600?
]
]

--

.large[
$0.635 + 0.498(3.5) + 0.0012(600) = 3.098$
]

---

### Class activity

.large[
.question[
Is there any verbal SAT score for which we would predict a college GPA of at least 3.5, for a student with a high school GPA of 3.5?
]
]

--

.large[
No: solving for SATV, we get

$0.635 + 0.498(3.5) + 0.0012(\text{SATV}) = 3.5$, so

$0.0012(\text{SATV}) = 3.5 - 0.635 - 0.498(3.5) = 1.22$, so

$\text{SATV} = 1.22/0.0012 = 1017 > 800$
]

---

### Class activity

.large[
.question[
Interpret the estimated slopes 0.498 and 0.0012.
]
]

--

.large[
* Holding verbal SAT score fixed, an increase of 1 point in high school GPA is associated with an increase of 0.498 points in college GPA.
* Holding high school GPA fixed, an increase of 1 point on the verbal SAT is associated with an increase of 0.0012 points in college GPA.
]

.large[
.question[
0.0012 is close to 0. Does that mean that verbal SAT score doesn't matter?
]
]

--

.large[
Not necessarily -- the scale of SAT scores is just very large (e.g., scores from 200 to 800)
]

---

### Interactions

.large[
\begin{align}
\text{bill length} &= \beta_0 + \beta_1 \text{IsChinstrap} + \beta_2 \text{IsGentoo} \\
& \hspace{0.5cm} + \beta_3 \text{body mass} + \varepsilon
\end{align}

We get a separate regression line for each species of penguin:

$$\text{bill length} = \begin{cases} \beta_0 + \beta_3 \text{body mass} +  \varepsilon & \text{species = Adelie} \\ \beta_0 + \beta_1 + \beta_3 \text{body mass} + \varepsilon & \text{species = Chinstrap} \\ \beta_0 + \beta_2 + \beta_3 \text{body mass} +  \varepsilon & \text{species = Gentoo} \end{cases}$$

]

.large[
.question[
What are we assuming about the slopes of these different regression lines?
]
]

--

.large[
That each line has the same slope
]

---

### Interactions

.large[
\begin{align}
\text{bill length} &= \beta_0 + \beta_1 \text{IsChinstrap} + \beta_2 \text{IsGentoo} \\
& \hspace{0.5cm} + \beta_3 \text{body mass} + \varepsilon
\end{align}

We get a separate regression line for each species of penguin:

$$\text{bill length} = \begin{cases} \beta_0 + \beta_3 \text{body mass} +  \varepsilon & \text{if Adelie} \\ \beta_0 + \beta_1 + \beta_3 \text{body mass} + \varepsilon & \text{if Chinstrap} \\ \beta_0 + \beta_2 + \beta_3 \text{body mass} +  \varepsilon & \text{if Gentoo} \end{cases}$$

]

.large[
.question[
How can we modify our regression model to allow a different slope for each species?
]
]

---

### Allowing the slope to be different

.large[
$$\text{bill length} = \begin{cases} \beta_0 + \beta_3 \text{body mass} +  \varepsilon & \text{if Adelie} \\ \beta_0 + \beta_1 + (\beta_3 + \beta_4) \text{body mass} + \varepsilon & \text{if Chinstrap} \\ \beta_0 + \beta_2 + (\beta_3 + \beta_5) \text{body mass} +  \varepsilon & \text{if Gentoo} \end{cases}$$
]

.large[
* Slope for Adelie penguins: $\beta_3$
* Slope for Chinstrap penguins: $\beta_3 + \beta_4$
* Slope for Gentoo penguins: $\beta_3 + \beta_5$
]

---

### Allowing the slope to be different

.large[
$$\text{bill length} = \begin{cases} \beta_0 + \beta_3 \text{body mass} +  \varepsilon & \text{if Adelie} \\ \beta_0 + \beta_1 + (\beta_3 + \beta_4) \text{body mass} + \varepsilon & \text{if Chinstrap} \\ \beta_0 + \beta_2 + (\beta_3 + \beta_5) \text{body mass} +  \varepsilon & \text{if Gentoo} \end{cases}$$
]

.large[
Written more concisely:

\begin{align}
\text{bill length} &= \beta_0 + \beta_1 \text{IsChinstrap} + \beta_2 \text{IsGentoo} \\
& \hspace{0.5cm} + \beta_3 \text{body mass} + \beta_4 \text{body mass} \cdot \text{IsChinstrap} \\
& \hspace{0.5cm} + \beta_5 \text{body mass} \cdot \text{IsGentoo} + \varepsilon
\end{align}
]

---

### Interactions

.large[
\begin{align}
\text{bill length} &= \beta_0 + \beta_1 \text{IsChinstrap} + \beta_2 \text{IsGentoo} \\
& \hspace{0.5cm} + \beta_3 \text{body mass} + \beta_4 \text{body mass} \cdot \text{IsChinstrap} \\
& \hspace{0.5cm} + \beta_5 \text{body mass} \cdot \text{IsGentoo} + \varepsilon
\end{align}
]

.large[
* The terms $\text{body mass} \cdot \text{IsChinstrap}$ and $\text{body mass} \cdot \text{IsGentoo}$ denote an *interaction* between body mass and species
* Interactions allow the slope for body mass to depend on species
* $\text{body mass} \cdot \text{IsChinstrap}$ literally means multiplication
]

---

### The fitted model

.large[
\begin{align}
\text{bill length} &= 27.11 + 5.06 \ \text{IsChinstrap} - 0.58 \ \text{IsGentoo} \\
& \hspace{0.5cm} + 0.0032 \ \text{body mass} \\
& \hspace{0.5cm} + 0.0013 \ \text{body mass} \cdot \text{IsChinstrap} \\
& \hspace{0.5cm} + 0.0010 \ \text{body mass} \cdot \text{IsGentoo}
\end{align}
]

.large[
.question[
What is the estimated bill length of a Gentoo penguin with a body mass of 5000g?
]

.abox[
(A) 43.11 mm
]

.bbox[
(B) 42.53 mm
]

.cbox[
(C) 47.53 mm
]
]

--

.large[
$27.11 − 0.58 + (0.0032 + 0.0010)∗5000 = 47.53$
]

---

### The fitted model

.large[
\begin{align}
\text{bill length} &= 27.11 + 5.06 \ \text{IsChinstrap} - 0.58 \ \text{IsGentoo} \\
& \hspace{0.5cm} + 0.0032 \ \text{body mass} \\
& \hspace{0.5cm} + 0.0013 \ \text{body mass} \cdot \text{IsChinstrap} \\
& \hspace{0.5cm} + 0.0010 \ \text{body mass} \cdot \text{IsGentoo}
\end{align}
]

.large[
.question[
What is the estimated slope on body mass, for Chinstrap penguins?
]

.abox[
(A) 0.0032
]

.bbox[
(B) 0.0013
]

.cbox[
(C) 0.0045
]
]

--

.large[
$0.0032 + 0.0013 = 0.0045$
]