---
title: Exam 1 Recap
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

### Agenda

.large[
* Exam 1 on Monday, in class
  * Closed notes
  * Bring a calculator (cannot use phone or laptop)
  * I won't ask you to write R code, but you may need to interpret R output
* Today:
  * Topics for Exam 1
  * Practice
* HW 3 released tonight, due next Friday
]

---

### Exam 1 Topics

.large[
* Data visualization
  * Appropriate plots for univariate and multivariate EDA
  * Appropriate summary statistics
* Linear regression models
  * Population model
  * Equation of the fitted line
  * Predictions with fitted line
  * Interpretation of slope and intercept
  * Least squares
]

---

### Exam 1 Topics (continued)

.large[
* Inference
  * Confidence intervals 
  * Hypothesis tests and p-values
  * Interpretation of confidence intervals and p-values
* Regression assumptions and diagnostic plots
  * Residual plots (shape and constant variance)
  * QQ plot (normality)
  * Independence and randomness
  * Transformations
* Unusual observations
  * Studentized residual plot
  * What to do with outliers?
]

---

### Class activity, Part I

.large[
Some practice questions with linear regression

[https://sta112-s22.github.io/class_activities/ca_lecture_14.html](https://sta112-s22.github.io/class_activities/ca_lecture_14.html)
]

---

### Class activity

.large[
Population regression model for the relationship between distance and price:
]

--

.large[
$\text{Price} = \beta_0 + \beta_1 \text{Distance} + \varepsilon$
]

---

### Class activity

.large[
Estimated regression line:
]

--

.large[
$\widehat{\text{Price}} = 352.16 - 53.01 \text{Distance}$
]

---

### Class activity

.large[
Predicted prices:
]

--

.large[
* 0.5 miles away: $352.16 - 53.01 \cdot 0.5 = 325.66$, so we predict a price of $325,660
* 2 miles away: $352.16 - 53.01 \cdot 2 = 246.14$, so we predict a price of $246,140
]

---

### Class activity

```{r, echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4.5}
library(tidyverse)
library(Stat2Data)
data("RailsTrails")

rt_lm <- RailsTrails %>%
  mutate(Price = Price2014) %>%
  lm(Price ~ Distance, data = .)

data.frame(x = predict(rt_lm),
           y = residuals(rt_lm)) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(size=2) +
  geom_abline(slope = 0, intercept = 0,
              lwd = 1.2, color="blue") +
  theme_bw() +
  labs(x = "Predicted price ($1000's)",
       y = "Residuals") +
  theme(text = element_text(size = 25))
```

.large[
.question[
Which of the two houses do you think will have a sale price closer its prediction?
]
]

--

.large[
The house 2 miles away
]

---

### Class activity

```{r, echo=F, fig.align='center', fig.width=7, fig.height=4.5}
rt_lm <- RailsTrails %>%
  mutate(Price = Price2014) %>%
  lm(Price ~ Distance, data = .)

data.frame(x = predict(rt_lm),
           y = residuals(rt_lm)) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(size=2) +
  geom_abline(slope = 0, intercept = 0,
              lwd = 1.2, color="blue") +
  theme_bw() +
  labs(x = "Predicted price ($1000's)",
       y = "Residuals") +
  theme(text = element_text(size = 25))
```

.large[
.question[
Do the shape and constant variance assumptions look reasonable?
]
]

--

.large[
Shape looks reasonable, constant variance doesn't.
]

---

### Class activity, Part II

.large[
Some practice questions with linear regression

[https://sta112-s22.github.io/class_activities/ca_lecture_14.html](https://sta112-s22.github.io/class_activities/ca_lecture_14.html)
]

---

### Class activity

.large[
Errors in interpretation:
]

--

.large[
* The p-value is *not* the probability that $H_0$ is true. It is the probability of the observed data, or something "more extreme", if $H_0$ is true
* The *slope*, not the intercept, is the change in GPA associated with a one hour increase in sleep
* The normality assumption refers to the residuals, not the predictor, and is checked with a QQ plot
* Can't talk about shape/direction of a relationship when the predictor is categorical with no inherent ordering
* We only have an observational study, so we can't conclude anything causal (we don't know whether changing sleep habits will change GPA)
]