---
title: Review
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

### Agenda

.large[
* Exam 2 on Wednesday
  * 50 minutes, in-class exam
  * Bring a calculator; no cell phones
  * Closed notes
* Project 1, Part 5 released
  * Due April 18
]

---

### Review activity

.large[
[https://sta112-s22.github.io/class_activities/ca_lecture_27.html](https://sta112-s22.github.io/class_activities/ca_lecture_27.html)
]

---

### Review activity

.large[

```{r, echo=F, message=F, warning=F, fig.align='center', fig.width=6, fig.height=4}
library(tidyverse)

set.seed(3)

x = rnorm(100, mean=50, sd=15)
crabs <- data.frame(Noise = rep(c("ocean", "ship"), each = 50),
                       Mass = x,
                       Oxygen = c(50 + 2*x[1:50] + rnorm(50, sd=15),
                                  300 - 2*x[51:100] + rnorm(50, sd=15)))

crabs %>%
  ggplot(aes(x = Mass, y = Oxygen, 
             color = Noise, shape = Noise)) +
  geom_point(size=2) +
  theme_bw() +
  theme(text = element_text(size=20))
```

.question[
Which model is the most appropriate, given this scatterplot?
]
]

--

.large[
$(3) \hspace{1cm} \text{oxygen} = \beta_0 + \beta_1 \text{IsShip} + \beta_2 \text{mass} + \beta_3 \text{IsShip} \cdot \text{mass} +  \varepsilon$
]

---

### Review activity

.large[
$(1) \hspace{1cm} \text{oxygen} = \beta_0 + \beta_1 \text{mass} + \varepsilon$

.question[
What would a scatterplot look like for this model?
]
]

--

```{r, echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=5}

set.seed(3)

x = rnorm(100, mean=50, sd=15)
sim1 <- data.frame(Noise = rep(c("ocean", "ship"), each = 50),
                       Mass = x,
                       Oxygen = c(2*x[1:50] + rnorm(50, sd=15),
                                  2*x[51:100] + rnorm(50, sd=15)))

sim1 %>%
  ggplot(aes(x = Mass, y = Oxygen, 
             color = Noise, shape = Noise)) +
  geom_point(size=3) +
  theme_bw() +
  theme(text = element_text(size=20))
```

---

### Review activity

.large[
$(2) \hspace{1cm} \text{oxygen} = \beta_0 + \beta_1 \text{IsShip} + \beta_2 \text{mass} + \varepsilon$

.question[
What would a scatterplot look like for this model?
]
]

--

```{r, echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=5}

set.seed(3)

x = rnorm(100, mean=50, sd=15)
sim2 <- data.frame(Noise = rep(c("ocean", "ship"), each = 50),
                       Mass = x,
                       Oxygen = c(50 + 2*x[1:50] + rnorm(50, sd=15),
                                  2*x[51:100] + rnorm(50, sd=15)))

sim2 %>%
  ggplot(aes(x = Mass, y = Oxygen, 
             color = Noise, shape = Noise)) +
  geom_point(size=3) +
  theme_bw() +
  theme(text = element_text(size=20))
```

---

### Review activity

.large[
$\widehat{\text{oxygen}} = 52.06 + 257.38 \ \text{IsShip} + 1.92 \ \text{mass} - 4.05 \ \text{IsShip} \cdot \text{mass}$

.question[
Using the estimated equation for the model you chose in question 1, what is the predicted rate of oxygen consumption for a 40g crab exposed to ocean noise? What about for a 40g crab exposed to ship noise?
]
]

--

.large[
**Exposed to ocean noise:**

$52.06 + 1.92 \cdot 40 = 128.86$

**Exposed to ship noise:**

$52.06 + 257.38 + (1.92 - 4.05) \cdot 40 = 224.24$
]

---

### Review activity

.large[
$\widehat{\text{oxygen}} = 52.06 + 257.38 \ \text{IsShip} + 1.92 \ \text{mass} - 4.05 \ \text{IsShip} \cdot \text{mass}$

.question[
What is the estimated change in oxygen consumption associated with a 1g increase in Mass, for crabs exposed to ship noise?
]
]

--

.large[
The slope on mass is $1.92 - 4.05 \cdot \text{IsShip}$. So for crabs exposed to ship noise, the slope is 

$1.92 - 4.05 = -2.13$
]

---

### Review activity

.large[
$\text{oxygen} = \beta_0 + \beta_1 \text{IsShip} + \beta_2 \text{mass} + \beta_3 \text{IsShip} \cdot \text{mass} +  \varepsilon$

.question[
The researchers want to test whether there is any relationship between ship noise and oxygen consumption, after accounting for mass. Write down a null and alternative hypothesis for this test, in terms of one or more parameters from the model.
]
]

--

.large[
$H_0: \beta_1 = \beta_3 = 0$

$H_A: \text{at least one of } \beta_1, \beta_3 \neq 0$
]

---

### Review activity

.large[
$\text{oxygen} = \beta_0 + \beta_1 \text{IsShip} + \beta_2 \text{mass} + \beta_3 \text{IsShip} \cdot \text{mass} +  \varepsilon$

$H_0: \beta_1 = \beta_3 = 0$

$H_A: \text{at least one of } \beta_1, \beta_3 \neq 0$

.question[
What test can the researchers use to test their hypotheses? Explain your reasoning, and provide the degrees of freedom.
]
]

--

.large[
* Test: a nested F test (because we are testing more than one parameter)
* Degrees of freedom: 2 and 96
]

---

### Review activity

.large[
```{r echo=F}
m1 <- lm(Oxygen ~ Mass, data = crabs)
anova(m1)
```

```{r echo=F}
m3 <- lm(Oxygen ~ Mass*Noise, data = crabs)
anova(m3)
```

]

---

### Review activity

.large[
**Test statistic:**

\begin{align}
F &= \dfrac{\frac{1}{\# \text{parameters tested}} (SSE_{reduced} - SSE_{full})}{\frac{1}{n-p} SSE_{full}} \\
&= \dfrac{\frac{1}{2}(163817 - 26192)}{\frac{1}{96} 26192} \\
&= 252.21
\end{align}
]