---
title: Hypothesis testing with multiple regression
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

### Agenda

.large[
* Reminder: HW 5 due Friday
* Recap: hypothesis testing with a categorical predictor
* Nested tests with multiple regression
]

---

### Hypothesis testing with a categorical predictor

.large[
$$\text{bill length} = \beta_0 + \beta_1 \text{IsChinstrap} + \beta_2 \text{IsGentoo} + \varepsilon$$

We want to know whether there is a relationship between species and bill length.

.question[
Can I translate this into null and alternative hypotheses about one or more model parameters?
]
]

--

.large[
$H_0: \beta_1 = \beta_2 = 0$

$H_A: \text{at least one of } \beta_1, \beta_2 \neq 0$
]

---

### Full and reduced models

.large[
$$\text{bill length} = \beta_0 + \beta_1 \text{IsChinstrap} + \beta_2 \text{IsGentoo} + \varepsilon$$
$H_0: \beta_1 = \beta_2 = 0$

$H_A: \text{at least one of } \beta_1, \beta_2 \neq 0$


**Reduced model ( $H_0$ ):**

$$\text{bill length} = \beta_0 + \varepsilon$$

**Full model ( $H_A$ ):**

$$\text{bill length} = \beta_0 + \beta_1 \text{IsChinstrap} + \beta_2 \text{IsGentoo} + \varepsilon$$

.question[
What test did we use to test these hypotheses?
]
]

---

### F test

.large[
**Variability:** $\hspace{3cm} \sum \limits_{i=1}^n (y_i - \overline{y})^2 = \sum \limits_{i=1}^n (\widehat{y}_i - \overline{y})^2 + \sum \limits_{i=1}^n (y_i - \widehat{y}_i)^2$

$\hspace{3cm} SSTotal \hspace{0.5cm} = \hspace{0.5cm} SSModel \hspace{0.5cm} + \hspace{0.5cm} SSE$

**Degrees of freedom:** 

$\hspace{4cm} n-1 \hspace{0.5cm} = \hspace{0.5cm} (p - 1)  \hspace{0.5cm} + \hspace{0.5cm} (n - p)$

( $n =$ number of observations, $p =$ number of parameters in full model)

**Test statistic:** $F = \dfrac{\text{between group variance}}{\text{within group variance}} = \dfrac{SSModel/(p-1)}{SSE/(n-p)}$

]

---

### F test

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=7, fig.height=4}
library(tidyverse)
library(palmerpenguins)

penguins_no_nas <- penguins %>%
  drop_na()

penguins_no_nas %>%
  ggplot(aes(x = species, y = bill_length_mm)) +
  geom_boxplot(lwd=1.5) +
  theme_classic() +
  labs(x = "Species", 
       y = "Bill length (mm)") +
  theme(text = element_text(size = 25))
```

.large[
**Test statistic:** $F = \dfrac{\text{between group variance}}{\text{within group variance}} = \dfrac{SSModel/(p-1)}{SSE/(n-p)} = 397.3$

.question[
How do I calculate a p-value?
]
]

---

### Calculating p-values

.large[

If $H_0$ is true, $F = \dfrac{SSModel/(p-1)}{SSE/(n-p)} \ \sim \ F_{p-1, \ n-p}$

**Degrees of freedom:** $p - 1 = 2$, $n - p = 330$

$F_{2, \ 330}$ distribution:

```{r echo=F, message=F, warning=F, fig.align='center', fig.width=6, fig.height=3.5}
data.frame(x = seq(0, 10, 0.1),
           y = df(seq(0, 10, 0.1), 2, 330)) %>%
  ggplot(aes(x = x, y = y)) +
  geom_line(lwd = 1.5) +
  labs(x = "", y = "density") +
  theme_bw() +
  theme(text = element_text(size = 25))
```

**p-value** =  $P(F_{2, \ 330} > 397.3) \approx 0$
]

---

### Summary so far

.large[
Our F test compares two *nested* models:

**Reduced model ( $H_0$ ):**

$$\text{bill length} = \beta_0 + \varepsilon$$

**Full model ( $H_A$ ):**

$$\text{bill length} = \beta_0 + \beta_1 \text{IsChinstrap} + \beta_2 \text{IsGentoo} + \varepsilon$$

**Test statistic:** $F = \dfrac{SSModel/(p-1)}{SSE/(n-p)} \ \sim \ F_{p-1, \ n-p}$

.question[
What if I want to test other hypotheses?
]
]

---

### Hypotheses for multiple regression

.large[
\begin{align}
\text{bill length} &= \beta_0 + \beta_1 \text{IsChinstrap} + \beta_2 \text{IsGentoo} \\ 
& \hspace{1cm} + \beta_3 \text{body mass} + \varepsilon
\end{align}

**Research question:** Is there any relationship between the predictors (species and body mass) and the response (bill length)?

.question[
How do I turn this question into hypotheses about one or more model parameters?
]
]

--

.large[
$H_0: \beta_1 = \beta_2 = \beta_3 = 0 \hspace{1cm} H_A: \text{at least one of } \beta_1, \beta_2, \beta_3 \neq 0$

.question[
What are the full and reduced models?
]
]

---

### Full and reduced models

.large[

$H_0: \beta_1 = \beta_2 = \beta_3 = 0 \hspace{1cm} H_A: \text{at least one of } \beta_1, \beta_2, \beta_3 \neq 0$

**Full model ( $H_A$ ):**

\begin{align}
\text{bill length} &= \beta_0 + \beta_1 \text{IsChinstrap} + \beta_2 \text{IsGentoo} \\ 
& \hspace{1cm} + \beta_3 \text{body mass} + \varepsilon
\end{align}

**Reduced model ( $H_0$ ):**

\begin{align}
\text{bill length} &= \beta_0 + \varepsilon
\end{align}

.question[
Note that the reduced model is *nested* inside the full model.
]
]

---

### Hypotheses for multiple regression

.large[
\begin{align}
\text{bill length} &= \beta_0 + \beta_1 \text{IsChinstrap} + \beta_2 \text{IsGentoo} \\ 
& \hspace{1cm} + \beta_3 \text{body mass} + \varepsilon
\end{align}

**Research question:** After accounting for species, is there a relationship between body mass and bill length?

.question[
How do I turn this question into hypotheses about one or more model parameters?
]
]

--

.large[
$H_0: \beta_3 = 0 \hspace{1cm} H_A: \beta_3 \neq 0$

.question[
What are the full and reduced models?
]
]

---

### Full and reduced models

.large[

$H_0: \beta_3 = 0 \hspace{1cm} H_A: \beta_3 \neq 0$

**Full model ( $H_A$ ):**

\begin{align}
\text{bill length} &= \beta_0 + \beta_1 \text{IsChinstrap} + \beta_2 \text{IsGentoo} \\ 
& \hspace{1cm} + \beta_3 \text{body mass} + \varepsilon
\end{align}

**Reduced model ( $H_0$ ):**

\begin{align}
\text{bill length} &= \beta_0 + \beta_1 \text{IsChinstrap} + \beta_2 \text{IsGentoo} + \varepsilon
\end{align}

.question[
Note that the reduced model is *nested* inside the full model.
]
]

---

### Hypotheses for multiple regression

.large[
\begin{align}
\text{bill length} &= \beta_0 + \beta_1 \text{IsChinstrap} + \beta_2 \text{IsGentoo} \\ 
& \hspace{1cm} + \beta_3 \text{body mass} + \varepsilon
\end{align}

**Research question:** After accounting for body mass, is there a relationship between species and bill length?

.question[
How do I turn this question into hypotheses about one or more model parameters?
]
]

--

.large[
$H_0: \beta_1, \beta_2 = 0 \hspace{1cm} H_A: \text{at least one of } \beta_1, \beta_2 \neq 0$

.question[
What are the full and reduced models?
]
]

---

### Full and reduced models

.large[

$H_0: \beta_1, \beta_2 = 0 \hspace{1cm} H_A: \text{at least one of } \beta_1, \beta_2 \neq 0$

**Full model ( $H_A$ ):**

\begin{align}
\text{bill length} &= \beta_0 + \beta_1 \text{IsChinstrap} + \beta_2 \text{IsGentoo} \\ 
& \hspace{1cm} + \beta_3 \text{body mass} + \varepsilon
\end{align}

**Reduced model ( $H_0$ ):**

\begin{align}
\text{bill length} &= \beta_0 + \beta_1 \text{body mass} + \varepsilon
\end{align}

.question[
Note that the reduced model is *nested* inside the full model.
]
]

---

### Hypotheses for multiple regression

.large[
\begin{align}
\text{bill length} &= \beta_0 + \beta_1 \text{IsChinstrap} + \beta_2 \text{IsGentoo} \\ 
& \hspace{1cm} + \beta_3 \text{body mass} + \varepsilon
\end{align}

**Research question:** Is the slope on body mass different for different species?

.question[
Can I test this research question with this model?
]
]

--

.large[
No -- the model assumes that the slope is the same for each species!

.question[
How can I change the model to test this research question?
]
]

---

### Adding interaction terms

.large[
\begin{align}
\text{bill length} &= \beta_0 + \beta_1 \text{IsChinstrap} + \beta_2 \text{IsGentoo} \\ 
& \hspace{1cm} + \beta_3 \text{body mass} + \beta_4 \text{body mass} \cdot \text{IsChinstrap} \\
& \hspace{1cm} + \beta_5 \text{body mass} \cdot \text{IsGentoo} + \varepsilon
\end{align}

**Research question:** Is the slope on body mass different for different species?

.question[
How do I turn this question into hypotheses about one or more model parameters?
]
]

--

.large[
$H_0: \beta_4, \beta_5 = 0 \hspace{1cm} H_A: \text{at least one of } \beta_4, \beta_5 \neq 0$

.question[
What are the full and reduced models?
]
]

---

### Testing interaction terms

.large[

$H_0: \beta_4, \beta_5 = 0 \hspace{1cm} H_A: \text{at least one of } \beta_4, \beta_5 \neq 0$

**Full model ( $H_A$ ):**

\begin{align}
\text{bill length} &= \beta_0 + \beta_1 \text{IsChinstrap} + \beta_2 \text{IsGentoo} \\ 
& \hspace{1cm} + \beta_3 \text{body mass} + \beta_4 \text{body mass} \cdot \text{IsChinstrap} \\
& \hspace{1cm} + \beta_5 \text{body mass} \cdot \text{IsGentoo} + \varepsilon
\end{align}

**Reduced model ( $H_0$ ):**

\begin{align}
\text{bill length} &= \beta_0 + \beta_1 \text{IsChinstrap} + \beta_2 \text{IsGentoo} \\ 
& \hspace{1cm} + \beta_3 \text{body mass} + \varepsilon
\end{align}

.question[
Note that the reduced model is *nested* inside the full model.
]
]

---

### Class activity

.large[
[https://sta112-s22.github.io/class_activities/ca_lecture_22.html](https://sta112-s22.github.io/class_activities/ca_lecture_22.html)
]

---

### Class activity

.large[
**Research question:** Is the slope on Age the same for both boys and girls?

.question[
What model should the researchers use to test this research question?
]
]

--

.large[
$$\text{Height} = \beta_0 + \beta_1 Age + \beta_2 \text{IsMale} + \beta_3 \text{Age} \cdot \text{IsMale} + \varepsilon$$

.question[
What are the full and reduced models that the researchers will compare?
]
]

---

### Class activity

.large[
**Full model:** 

$$\text{Height} = \beta_0 + \beta_1 Age + \beta_2 \text{IsMale} + \beta_3 \text{Age} \cdot \text{IsMale} + \varepsilon$$

**Reduced model:** 

$$\text{Height} = \beta_0 + \beta_1 Age + \beta_2 \text{IsMale} + \varepsilon$$
]

---

### Class activity

.large[
$$\text{Height} = \beta_0 + \beta_1 Age + \beta_2 \text{IsMale} + \beta_3 \text{Age} \cdot \text{IsMale} + \varepsilon$$

**New research question:** Is there *any* relationship between Age and Height, after accounting for gender?

.question[
What are the full and reduced models that the researchers will compare?
]
]

---

### Class activity

.large[
**New research question:** Is there *any* relationship between Age and Height, after accounting for gender?

**Full model:** 

$$\text{Height} = \beta_0 + \beta_1 Age + \beta_2 \text{IsMale} + \beta_3 \text{Age} \cdot \text{IsMale} + \varepsilon$$

**Reduced model:** 

$$\text{Height} = \beta_0 + \beta_1 \text{IsMale} + \varepsilon$$

]

---

### What we've done so far

.large[
* Specifying hypotheses for multiple regression models
* Hypotheses imply full and reduced models

.question[
How do we compare full and reduced models with a hypothesis test?
]
]

---

### When our reduced model is intercept-only

.large[
We've already done this for one categorical predictor!

**Reduced model ( $H_0$ ):**

$$\text{bill length} = \beta_0 + \varepsilon$$

**Full model ( $H_A$ ):**

$$\text{bill length} = \beta_0 + \beta_1 \text{IsChinstrap} + \beta_2 \text{IsGentoo} + \varepsilon$$

**Test statistic:** $F = \dfrac{SSModel/(p-1)}{SSE/(n-p)} \ \sim \ F_{p-1, \ n-p}$

.question[
What if I want to compare different nested models?
]
]

---

### Decomposing sums of squares

.large[

$F = \dfrac{SSModel/(p-1)}{SSE/(n-p)}$

* SSModel and SSE really refer to the *full* model ( $\widehat{y}_i$ are predictions from the *full* model, not the reduced model)
* So let's add subscripts to make that clear

$\hspace{3cm} \sum \limits_{i=1}^n (y_i - \overline{y})^2 = \sum \limits_{i=1}^n (\widehat{y}_{i, full} - \overline{y})^2 + \sum \limits_{i=1}^n (y_i - \widehat{y}_{i, full})^2$

$\hspace{3cm} SSTotal \hspace{0.5cm} = \hspace{0.5cm} SSModel_{full} \hspace{0.5cm} + \hspace{0.5cm} SSE_{full}$

$F = \dfrac{SSModel_{full}/(p-1)}{SSE_{full}/(n-p)} = \dfrac{(SSTotal - SSE_{full})/(p-1)}{SSE_{full}/(n-p)}$
]

---

### Decomposing sums of squares

.large[
$F = \dfrac{(SSTotal - SSE_{full})/(p-1)}{SSE_{full}/(n-p)}$

We can decompose sums of squares for the reduced model too:

$\hspace{3cm} \sum \limits_{i=1}^n (y_i - \overline{y})^2 = \sum \limits_{i=1}^n (\widehat{y}_{i, reduced} - \overline{y})^2 + \sum \limits_{i=1}^n (y_i - \widehat{y}_{i, reduced})^2$

$\hspace{3cm} SSTotal \hspace{0.5cm} = \hspace{0.5cm} SSModel_{reduced} \hspace{0.5cm} + \hspace{0.5cm} SSE_{reduced}$

.question[
For this test, the reduced model is 

$$\text{bill length} = \beta_0 + \varepsilon$$

What are the predictions $\widehat{y}_{i, reduced}$ for this intercept-only model?
]
]

---

### Decomposing sums of squares

.large[
$F = \dfrac{(SSTotal - SSE_{full})/(p-1)}{SSE_{full}/(n-p)}$

We can decompose sums of squares for the reduced model too:

$\hspace{3cm} \sum \limits_{i=1}^n (y_i - \overline{y})^2 = \sum \limits_{i=1}^n (\widehat{y}_{i, reduced} - \overline{y})^2 + \sum \limits_{i=1}^n (y_i - \widehat{y}_{i, reduced})^2$

$\hspace{3cm} SSTotal \hspace{0.5cm} = \hspace{0.5cm} SSModel_{reduced} \hspace{0.5cm} + \hspace{0.5cm} SSE_{reduced}$

**Reduced model:**  $\text{bill length} = \beta_0 + \varepsilon$

**Predictions:** $\widehat{y}_{i,reduced} = \overline{y}$

.question[
If $\widehat{y}_{i,reduced} = \overline{y}$, what are $SSModel_{reduced}$ and $SSE_{reduced}$ ?
]
]


---

### Decomposing sums of squares

.large[
$F = \dfrac{(SSTotal - SSE_{full})/(p-1)}{SSE_{full}/(n-p)}$

We can decompose sums of squares for the reduced model too:

$\hspace{3cm} \sum \limits_{i=1}^n (y_i - \overline{y})^2 = \sum \limits_{i=1}^n (\widehat{y}_{i, reduced} - \overline{y})^2 + \sum \limits_{i=1}^n (y_i - \widehat{y}_{i, reduced})^2$

$\hspace{3cm} SSTotal \hspace{0.5cm} = \hspace{0.5cm} SSModel_{reduced} \hspace{0.5cm} + \hspace{0.5cm} SSE_{reduced}$

**Reduced model:**  $\text{bill length} = \beta_0 + \varepsilon$

**Predictions:** $\widehat{y}_{i,reduced} = \overline{y}$

$SSModel_{reduced} = 0 \hspace{2cm} SSE_{reduced} = SSTotal$

]

---

### Decomposing sums of squares

.large[

$\hspace{3cm} \sum \limits_{i=1}^n (y_i - \overline{y})^2 = \sum \limits_{i=1}^n (\widehat{y}_{i, reduced} - \overline{y})^2 + \sum \limits_{i=1}^n (y_i - \widehat{y}_{i, reduced})^2$

$\hspace{3cm} SSTotal \hspace{0.5cm} = \hspace{0.5cm} SSModel_{reduced} \hspace{0.5cm} + \hspace{0.5cm} SSE_{reduced}$

**Reduced model:**  $\text{bill length} = \beta_0 + \varepsilon$

**Predictions:** $\widehat{y}_{i,reduced} = \overline{y}$

$SSModel_{reduced} = 0 \hspace{2cm} SSE_{reduced} = SSTotal$

\begin{align}
F &= \dfrac{(SSTotal - SSE_{full})/(p-1)}{SSE_{full}/(n-p)} \\ 
&= \dfrac{(SSE_{reduced} - SSE_{full})/(p-1)}{SSE_{full}/(n-p)}
\end{align}

]

---

### Re-writing our F test

.large[
Our F test compares two *nested* models:

**Reduced model ( $H_0$ ):**

$$\text{bill length} = \beta_0 + \varepsilon$$

**Full model ( $H_A$ ):**

$$\text{bill length} = \beta_0 + \beta_1 \text{IsChinstrap} + \beta_2 \text{IsGentoo} + \varepsilon$$

**Test statistic:** $F = \dfrac{(SSE_{reduced} - SSE_{full})/(p-1)}{SSE_{full}/(n-p)} \ \sim \ F_{p-1, \ n-p}$

* $p - 1 =$ difference in number of parameters between models
* $p =$ number of parameters in full model
]

---

### Nested F test

.large[
To compare *any* two **nested** models:

**Test statistic:** $F = \dfrac{\frac{1}{\# \ \text{parameters tested}}(SSE_{reduced} - SSE_{full})}{\frac{1}{n-p} SSE_{full}}$

**Distribution:** $F_{\# \ \text{parameters tested}, \ n-p}$

* $n =$ number of observations in data
* $p =$ number of parameters in full model
]