---
title: Hypothesis testing with multiple regression
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

### Agenda

.large[
* HW 6 and Project 1, Part 4 released tonight (both short)
* Today:
  * Introduce Project 1, Part 4
  * Hypothesis testing
  * Confidence intervals
]

---

### Project 1, Part 4

.large[
**Data analysis plan:** You will

* Describe the model(s) you plan to use to answer the research question(s)
  * *Note: It is ok if these models change a bit when you actually work with the data*
* Describe the assumptions made by your models
  * How do you plan to assess these assumptions?
  * What will you do if you the assumptions are violated? (e.g., transformations)
* Describe how you will use the model(s) to answer the research question(s)
  * E.g. confidence intervals, hypothesis tests, etc.
* You do not need to do any data analysis yet!
]

---

### Recap: hypothesis tests and nested models

.large[
**Data:** Height, age, and gender for 198 children between the ages of 8 and 18. Researchers think the relationship between age and height might depend on gender.

**Research question:** Is there *any* relationship between Age and Height, after accounting for gender?

**Full model:** 

$$\text{Height} = \beta_0 + \beta_1 Age + \beta_2 \text{IsMale} + \beta_3 \text{Age} \cdot \text{IsMale} + \varepsilon$$

**Reduced model:** 

$$\text{Height} = \beta_0 + \beta_1 \text{IsMale} + \varepsilon$$

]

---

### Recap: hypothesis tests and nested models

.large[
**Data:** Height, age, and gender for 198 children between the ages of 8 and 18. Researchers think the relationship between age and height might depend on gender.


**Research question:** Is the slope on Age the same for both boys and girls?

**Full model:** 

$$\text{Height} = \beta_0 + \beta_1 Age + \beta_2 \text{IsMale} + \beta_3 \text{Age} \cdot \text{IsMale} + \varepsilon$$

**Reduced model:** 

$$\text{Height} = \beta_0 + \beta_1 Age + \beta_2 \text{IsMale} + \varepsilon$$
]

---

### Nested F test

.large[
To compare *any* two **nested** models:

**Test statistic:** $F = \dfrac{\frac{1}{\# \ \text{parameters tested}}(SSE_{reduced} - SSE_{full})}{\frac{1}{n-p} SSE_{full}}$

**Distribution:** $F_{\# \ \text{parameters tested}, \ n-p}$

* $n =$ number of observations in data
* $p =$ number of parameters in full model
]

.large[
**Intuition:**  $SSE_{reduced} - SSE_{full}$ is large when the full model fits the data better than the reduced model
]

---

### Class activity, Part I

.large[
[https://sta112-s22.github.io/class_activities/ca_lecture_23.html](https://sta112-s22.github.io/class_activities/ca_lecture_23.html)
]

---

### Class activity

.large[
$$F = \dfrac{\frac{1}{\# \ \text{parameters tested}}(SSE_{reduced} - SSE_{full})}{\frac{1}{n-p} SSE_{full}}$$

.question[
Calculate the test statistic.
]
]

---

### Class activity

.large[
$$F = \dfrac{\frac{1}{\# \ \text{parameters tested}}(SSE_{reduced} - SSE_{full})}{\frac{1}{n-p} SSE_{full}}$$

```{r include=F}
library(tidyverse)
library(Stat2Data)
library(knitr)
data("Kids198")

Kids198 <- Kids198 %>%
  mutate(Sex = ifelse(Sex == 0, "male", "female"))

hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   more <- "..."
   if (length(lines)==1) {        # first n lines
     if (length(x) > lines) {
       # truncate the output, but add ....
       x <- c(head(x, lines), more)
     }
   } else {
     x <- c(more, x[lines], more)
   }
   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

m1 <- lm(Height ~ Age*Sex, data = Kids198)
m2 <- lm(Height ~ Sex, data = Kids198)
```

```{r output.lines=8:9}
anova(m1)
```

```{r output.lines=6:7}
anova(m2)
```
]

---

### Class activity

.large[
$$F = \dfrac{\frac{1}{\# \ \text{parameters tested}}(SSE_{reduced} - SSE_{full})}{\frac{1}{n-p} SSE_{full}}$$
```{r echo=F, output.lines=8:9}
anova(m1)
```

```{r echo=F, output.lines=6:7}
anova(m2)
```

$$F = \dfrac{\frac{1}{2}(6733.9 - 1686.3)}{\frac{1}{194}1686.3} = 290.35$$

]

---

### Class activity

.large[
.question[
What is the p-value for this test?
]

```{r}
pf(290.35, df1 = 2, df2 = 194, lower.tail=F)
```

The p-value is basically 0, so we have strong evidence that there is a relationship between Age and Height, after accounting for Sex.
]

---

### Testing in R

.large[
We can do comparisons between nested models a little more efficiently:

```{r}
m1 <- lm(Height ~ Age*Sex, data = Kids198)
m2 <- lm(Height ~ Sex, data = Kids198)

anova(m2, m1)
```
]

---

### Class activity, Part II

.large[
[https://sta112-s22.github.io/class_activities/ca_lecture_23.html](https://sta112-s22.github.io/class_activities/ca_lecture_23.html)
]

---

### Class activity

.large[
.question[
Do you have evidence that the slope is different for boys and girls?
]
]

--

.large[
```{r}
m1 <- lm(Height ~ Age*Sex, data = Kids198)
m2 <- lm(Height ~ Age + Sex, data = Kids198)

anova(m2, m1)
```
]

---

### Class activity


.large[
```{r}
m1 <- lm(Height ~ Age*Sex, data = Kids198)
m2 <- lm(Height ~ Age + Sex, data = Kids198)

anova(m2, m1)
```

Yes, we have strong evidence that the slope on Age is different for boys and girls.
]

---

### Testing a single parameter

.large[
$$\text{Height} = \beta_0 + \beta_1 Age + \beta_2 \text{IsMale} + \beta_3 \text{Age} \cdot \text{IsMale} + \varepsilon$$
$H_0: \beta_3 = 0 \hspace{1cm} H_A: \beta_3 \neq 0$

.question[
How did we test a single coefficient in simple linear regression?
]
]

--

.large[
With a t-test!
]

---

### Testing a single parameter

.large[
$$\text{Height} = \beta_0 + \beta_1 Age + \beta_2 \text{IsMale} + \beta_3 \text{Age} \cdot \text{IsMale} + \varepsilon$$

$H_0: \beta_3 = 0 \hspace{1cm} H_A: \beta_3 \neq 0$

$t = \dfrac{\widehat{\beta}_3 - 0}{SE_{\widehat{\beta}_3}}$

Under $H_0$, this test statistic follows a $t$ distribution.

.question[
What do you think should be the degrees of freedom of the $t$ distribution?
]
]

--

.large[
$n - p$
]

---

### Testing a single parameter

.large[
$$\text{Height} = \beta_0 + \beta_1 Age + \beta_2 \text{IsMale} + \beta_3 \text{Age} \cdot \text{IsMale} + \varepsilon$$

$H_0: \beta_3 = 0 \hspace{1cm} H_A: \beta_3 \neq 0$

$t = \dfrac{\widehat{\beta}_3 - 0}{SE_{\widehat{\beta}_3}}$

Under $H_0$, this test statistic follows a $t_{n-p}$ distribution.

]

---

### Testing a single parameter

.large[
$$\text{Height} = \beta_0 + \beta_1 Age + \beta_2 \text{IsMale} + \beta_3 \text{Age} \cdot \text{IsMale} + \varepsilon$$

```{r, output.lines = 10:14}
m1 <- lm(Height ~ Age*Sex, data = Kids198)
summary(m1)
```

$t = \dfrac{\widehat{\beta}_3 - 0}{SE_{\widehat{\beta}_3}} = \dfrac{0.0641}{0.0125} = 5.112$

p-value = $7.6 \times 10^{-7}$
]


---

### Testing a single parameter

.large[
$$\text{Height} = \beta_0 + \beta_1 Age + \beta_2 \text{IsMale} + \beta_3 \text{Age} \cdot \text{IsMale} + \varepsilon$$

**Research question:** Is the slope on Age *larger* for boys than for girls?

.question[
What hypotheses could I test to answer this research question?
]

]

--

.large[
$H_0: \beta_3 = 0 \hspace{1cm} H_A: \beta_3 > 0$

**Test statistic:** $t = \dfrac{\widehat{\beta}_3 - 0}{SE_{\widehat{\beta}_3}} = 5.112$

.question[
Which tail do I use to calculate the p-value?
]
]

---

### Testing a single parameter

.large[
$$\text{Height} = \beta_0 + \beta_1 Age + \beta_2 \text{IsMale} + \beta_3 \text{Age} \cdot \text{IsMale} + \varepsilon$$

**Research question:** Is the slope on Age *larger* for boys than for girls?

$H_0: \beta_3 = 0 \hspace{1cm} H_A: \beta_3 > 0$

**Test statistic:** $t = \dfrac{\widehat{\beta}_3 - 0}{SE_{\widehat{\beta}_3}} = 5.112$

```{r}
pt(5.112, df = 194, lower.tail=F)
```

p-value = $3.8 \times 10^{-7}$, so we have strong evidence that the slope on Age is larger for boys than for girls.
]

---

### Testing a single parameter

.large[
$$\text{Height} = \beta_0 + \beta_1 Age + \beta_2 \text{IsMale} + \beta_3 \text{Age} \cdot \text{IsMale} + \varepsilon$$

**Research question:** Is the slope on Age *larger* for boys than for girls?

$H_0: \beta_3 = 0 \hspace{1cm} H_A: \beta_3 > 0$

.question[
Can I use a nested F test here?
]

]

.large[
No -- a nested F test doesn't let me test a specific direction ( $\beta_3 > 0$ )
]

---

### t tests vs. nested F tests

.large[

.pull-left[
**t test:** test a single parameter

Some example hypotheses:
* $H_0: \beta_3 = 0 \hspace{0.5cm} H_A: \beta_3 \neq 0$
* $H_0: \beta_3 = 0 \hspace{0.5cm} H_A: \beta_3 > 0$
* $H_0: \beta_3 = 1 \hspace{0.5cm} H_A: \beta_3 < 1$
]

.pull-right[
**nested F test:** test one or more parameters

Some example hypotheses:

$H_0: \beta_1 = \beta_2 = 0$
$H_A: \text{at least one of } \beta_1, \beta_2 \neq 0$
]

]

---

### Confidence intervals

.large[
Just like with simple linear regression, we can construct confidence intervals for single parameters.

$$\text{Height} = \beta_0 + \beta_1 Age + \beta_2 \text{IsMale} + \beta_3 \text{Age} \cdot \text{IsMale} + \varepsilon$$

Confidence interval for $\beta_3$:

$$\widehat{\beta}_3 \pm t^* SE_{\widehat{\beta}_3}$$

where $t^*$ is the critical value for a $t_{n-p}$ distribution.
]

---

### Confidence intervals

.large[
$$\text{Height} = \beta_0 + \beta_1 Age + \beta_2 \text{IsMale} + \beta_3 \text{Age} \cdot \text{IsMale} + \varepsilon$$

**Confidence interval:** $\hspace{1cm} \widehat{\beta}_3 \pm t^* SE_{\widehat{\beta}_3}$

.question[
Suppose I want to calculate a 99% confidence interval. How do I find the critical value $t^*$?
]
]

--

.large[
```{r}
qt(0.005, df=194, lower.tail=F)
```

$t^* = 2.60$
]

---

### Confidence intervals

.large[
$$\text{Height} = \beta_0 + \beta_1 Age + \beta_2 \text{IsMale} + \beta_3 \text{Age} \cdot \text{IsMale} + \varepsilon$$

**Confidence interval:** $\hspace{1cm} \widehat{\beta}_3 \pm t^* SE_{\widehat{\beta}_3}$

$$0.0641 \pm 2.60 \cdot 0.0125 = (0.0316, 0.0966)$$

So we are 99% confident that a one-month increase in age is associated with an change in height that is between 0.0316 and 0.0966 inches greater for boys than for girls.
]