---
title: Transformations
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

### Agenda

.large[
* Project 1, Part 2 released
  * regression, diagnostics, and inference with the college scorecard data
  * we will talk about studentized residuals and outliers on Monday
  * due Friday, Feb 18, 12:00pm (noon) on Canvas
* Today:
  * Transformations
  * Lab 3
]

---

### Last time

.large[
Predicting the number of doctors from the number of hospitals in a county:

```{r, echo=F, message=F, fig.align='center', fig.width=7, fig.height=5}
library(Stat2Data)
library(tidyverse)
data("CountyHealth")
CountyHealth %>%
  ggplot(aes(x = Hospitals, 
             y = MDs)) +
  geom_point(size=2.5) +
  labs(x = "Number of hospitals",
       y = "Number of doctors") +
  theme_bw() +
  theme(text = element_text(size = 25))
```
]

---

### Regression diagnostics

.pull-left[
```{r, echo=F, message=F, fig.width=6, fig.height=5, fig.align='center'}
health_lm <- lm(MDs ~ Hospitals, data = CountyHealth)
CountyHealth %>%
  mutate(yhat = predict(health_lm),
         residuals = residuals(health_lm)) %>%
  ggplot(aes(x = yhat, y = residuals)) +
  geom_point(size=2) +
  geom_abline(slope = 0, intercept = 0, 
              color = "blue", lwd = 1.2) +
  labs(x = "Predicted number of doctors",
       y = "Residual") +
  theme_bw() +
  theme(text = element_text(size = 25))
```
]

.pull-right[
```{r, echo=F, message=F, fig.width=6, fig.height=5, fig.align='center'}
CountyHealth %>%
  mutate(residuals = residuals(health_lm)) %>%
  ggplot(aes(sample = residuals)) +
  geom_qq(size=2) +
  geom_qq_line() +
  labs(x = "Theoretical normal quantiles",
       y = "Observed residual quantiles") +
  theme_classic() +
  theme(text = element_text(size=25))
```
]

---

### Transformations

.large[
What if we look at the square root of the number of doctors instead?
]

.pull-left[
```{r, echo=F, fig.align='center', fig.width=6, fig.height=5}
CountyHealth %>%
  ggplot(aes(x = Hospitals, 
             y = MDs)) +
  geom_point(size=2.5) +
  labs(x = "Number of hospitals",
       y = "Number of doctors") +
  theme_bw() +
  theme(text = element_text(size = 25))
```
]

.pull-right[
```{r, echo=F, fig.align='center', fig.width=6, fig.height=5}
CountyHealth %>%
  ggplot(aes(x = Hospitals, 
             y = sqrt(MDs))) +
  geom_point(size=2.5) +
  labs(x = "Number of hospitals",
       y = "Sqrt(Number of doctors)") +
  theme_bw() +
  theme(text = element_text(size = 25))
```
]

---

### Diagnostic plots

.pull-left[
```{r, echo=F, message=F, fig.width=6, fig.height=5, fig.align='center'}
health_lm <- lm(sqrt(MDs) ~ Hospitals, data = CountyHealth)
CountyHealth %>%
  mutate(yhat = predict(health_lm),
         residuals = residuals(health_lm)) %>%
  ggplot(aes(x = yhat, y = residuals)) +
  geom_point(size=2) +
  geom_abline(slope = 0, intercept = 0, 
              color = "blue", lwd = 1.2) +
  labs(x = "Predicted sqrt(number of doctors)",
       y = "Residual") +
  theme_bw() +
  theme(text = element_text(size = 25))
```
]

.pull-right[
```{r, echo=F, message=F, fig.width=6, fig.height=5, fig.align='center'}
CountyHealth %>%
  mutate(residuals = residuals(health_lm)) %>%
  ggplot(aes(sample = residuals)) +
  geom_qq(size=2) +
  geom_qq_line() +
  labs(x = "Theoretical normal quantiles",
       y = "Observed residual quantiles") +
  theme_classic() +
  theme(text = element_text(size=25))
```
]

.large[
.question[
How do the shape, constant variance, and normality assumptions look now?
]
]

---

### Transformations

.large[
**Transformations** can sometimes be used to address violations of shape, constant variance, and normality assumptions.
]

.pull-left[
```{r, echo=F, fig.align='center', fig.width=6, fig.height=5}
data("SpeciesArea")
SpeciesArea %>%
  ggplot(aes(x = Area, 
             y = Species)) +
  geom_point(size=2.5) +
  labs(x = "Area of island (sq km)",
       y = "Number of species") +
  theme_bw() +
  theme(text = element_text(size = 25))
```
]

.pull-right[
```{r, echo=F, fig.align='center', fig.width=6, fig.height=5}
SpeciesArea %>%
  ggplot(aes(x = log(Area), 
             y = log(Species))) +
  geom_point(size=2.5) +
  labs(x = "log(Area)",
       y = "log(Number of species)") +
  theme_bw() +
  theme(text = element_text(size = 25))
```
]

---

### Using transformations

.large[
.center[
$\sqrt{\widehat{\text{MDs}}} = -2.75 + 6.88 \text{Hospitals}$
]
]

.large[
.question[
If a county has 8 hospitals, how many doctors do we predict?
]

.abox[
(A) 52.59
]

.bbox[
(B) 2734.24
]

.cbox[
(C) 7.23
]
]

--

.large[
**Answer:** $\widehat{\text{MDs}} = (-2.75 + 6.88 \cdot 8)^2 = 2734.24$
]

---

### Using transformations

.large[
.center[
$\sqrt{\widehat{\text{MDs}}} = -2.75 + 6.88 \text{Hospitals}$
]
]

.large[
.question[
How should I interpret the estimated slope, 6.88?
]

.abox[
(A) An increase of 1 hospital is associated with an increase of 6.88 doctors.
]

.bbox[
(B) An increase of 1 hospital is associated with an increase of $6.88^2$ doctors.
]

.cbox[
(C) An increase of 1 hospital is associated with an increase of 6.88 in the square root of the number of doctors.
]
]

--

.large[
**Answer:** An increase of 1 hospital is associated with an increase of 6.88 in the square root of the number of doctors.
]

---

### Choosing transformations

.large[
* Choose a transformation for $x$ that looks like the relationship
* Experiment with transforming $x$ and/or $y$
* See section 1.4 in book for intuition
]

```{r, echo=F, fig.align='center', fig.width=8, fig.height=5.5}
library(patchwork)
p1 <- data.frame(x = 1:100) %>%
  ggplot(aes(x = x, y = sqrt(x))) +
  geom_line(size = 1.2) +
  theme_bw() +
  theme(text = element_text(size = 20)) +
  labs(title = "Square root")

p2 <- data.frame(x = 1:100) %>%
  ggplot(aes(x = x, y = x^2)) +
  geom_line(size = 1.2) +
  theme_bw() +
  theme(text = element_text(size = 20)) +
  labs(title = "Square")

p3 <- data.frame(x = 1:100) %>%
  ggplot(aes(x = x, y = log(x))) +
  geom_line(size = 1.2) +
  theme_bw() +
  theme(text = element_text(size = 20)) +
  labs(title = "Log")

p4 <- data.frame(x = 1:100) %>%
  ggplot(aes(x = x, y = 1/x)) +
  geom_line(size = 1.2) +
  theme_bw() +
  theme(text = element_text(size = 20)) +
  labs(title = "Reciprocal")

(p1 + p2)/(p3 + p4)
```

---

### Lab 3

.large[
[https://sta112-s22.github.io/labs/lab_3.html](https://sta112-s22.github.io/labs/lab_3.html)
]