---
title: "Lab 4 Solutions"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(moderndive)
data("evals")
```

## Exercise 1

```{r}
evals %>%
  ggplot(aes(x = rank, y = score, color = gender)) +
  geom_boxplot() +
  theme_bw()
```

For male professors, it looks like teaching professors get higher scores than tenure-track and tenured professors. For female professors, on the other hand, it looks like teaching professors get lower scores than tenure-track and tenured professors. Female teaching and tenure-track professors appear to get lower scores than their male counterparts, whereas tenured professors from both genders seem to score the same.

## Exercise 2

```{r}
eval_lm1 <- lm(score ~ rank, data = evals)
summary(eval_lm1)
```

$$\widehat{\text{Score}} = 4.28 - 0.13 \text{IsTenureTrack} - 0.15 \text{IsTenured}$$

$R^2_{adj} = 0.0073$ (we explain very little variability in score by differences in rank)

## Exercise 3

Teaching professors: the average score is 4.28 (the intercept in our fitted model)

Tenure track professors: the average score is $4.28 - 0.13 = 4.15$

Tenured professors: the average score is $4.28 - 0.15 = 4.13$

## Exercise 4

```{r}
eval_lm2 <- lm(score ~ rank + gender, data = evals)
summary(eval_lm2)
```

$$\widehat{\text{Score}} = 4.20 - 0.10 \ \text{IsTenureTrack} - 0.18 \ \text{IsTenured} + 0.17 \ \text{IsMale}$$

$R^2_{adj} = 0.0266$ (we explain a little more variability in score by adding gender to the model)

## Exercise 5

The predicted average score for male teaching professors is $4.2 + 0.17 = 4.37$


## Exercise 6

On average, male professors score 0.17 points higher than female professors of the same rank.

## Exercise 7

No -- the plot from exercise 1 suggests that we need an interaction. There is a noticeable difference in scores for male and female teaching professors, but male and female tenured professors have very similar scores.

## Exercise 8

```{r}
eval_lm3 <- lm(score ~ rank*gender, data = evals)
summary(eval_lm3)
```

$$\widehat{\text{Score}} = 4.038 + 0.059 \ \text{IsTenureTrack} + 0.087 \ \text{IsTenured} + 0.483 \ \text{IsMale} - 0.324 \ \text{IsTenureTrack} \cdot \text{IsMale} - 0.463 \ \text{IsTenured} \cdot \text{IsMale}$$
$R^2_{adj} = 0.050$

## Exercise 9

The estimated average score for male teaching professors is $4.038 + 0.483 = 4.521$


## Exercise 10

On average, male teaching professors score 0.483 points higher than female teaching professors.


## Exercise 11

On average, male tenured professors score $0.483 - 0.463 = 0.02$ points higher than female tenured professors.

## Exercise 12

I would argue that it is important to include an interaction. While $R^2_{adj}$ is still low, adding the interaction improved $R^2_{adj}$ by as much as adding gender to the model. Furthermore, the plot for exercise 1 suggests an interaction is needed, because the difference between male and female professors depends on rank.
