---
title: "Class Activity, February 18"
output: 
  tufte::tufte_html:
    css: "lab.css"
    tufte_variant: "envisioned"
    highlight: pygments
link-citations: yes
---

# Part I

Is there a relationship between a house's distance from bike trails, and its selling price?  The `RailsTrails` data looks at a sample of 104 houses in Northampton, MA, and includes the following variables:

* `Price`: the sale price (in $1000's)
* `Distance`: distance (in feet) to the nearest bike trail

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(Stat2Data)
data("RailsTrails")
```

The scatterplot below plots Distance and Price for the sample of 104 homes.

```{r, echo=F, fig.align='center', fig.width=6, fig.height=4}
RailsTrails %>%
  ggplot(aes(x = Distance, y = Price2014)) +
  geom_point(size=2) +
  labs(x = "Distance (feet)", y = "Price ($1000's)") +
  theme_bw() +
  theme(text = element_text(size=20))
```

You want to model the relationship between Distance and Price in the population of all houses in Northampton, MA. You decide to use simple linear regression. You fit the regression model in R, producing the output below:

```{r, echo=F}
RailsTrails %>%
  mutate(Price = Price2014) %>%
  lm(Price ~ Distance, data = .)
```

To assess the shape and constant variance assumptions of your linear regression model, you also create a residual plot:

```{r, echo=F, fig.align='center', fig.width=6, fig.height=4}
rt_lm <- RailsTrails %>%
  mutate(Price = Price2014) %>%
  lm(Price ~ Distance, data = .)

data.frame(x = predict(rt_lm),
           y = residuals(rt_lm)) %>%
  ggplot(aes(x = x, y = y)) +
  geom_point(size=2) +
  geom_abline(slope = 0, intercept = 0,
              lwd = 1.2, color="blue") +
  theme_bw() +
  labs(x = "Predicted price ($1000's)",
       y = "Residuals") +
  theme(text = element_text(size = 20))
```

1. Write down the population linear regression model for the relationship between Distance and Price in the population.

2. Using the R output above, write the equation of the fitted regression line which predicts Price from Distance.

3. Use the estimated regression line to predict the price for two houses: one which is 0.5 miles from a bike trail, and one which is 2 miles from a bike trail.

4. The true selling prices for the two houses are unlikely to be exactly equal to our predictions. Based on the residual plot, which of the two houses do you think will have a sale price closer its prediction from Question 3? Explain your reasoning.

5. Use the residual plot to assess the shape and constant variance assumptions. Do these two assumptions look reasonable? Explain.


# Part II

Imagine you are a statistician studying factors associated with student success in college. One of your collaborators conducts an observational study in which they survey 100 Wake Forest students, and record their GPA, the average number of hours they sleep per night, and their favorite place to do homework. After analyzing the data, they make the following plots:

```{r echo=F, warning=F, message=F, fig.align='center', fig.width=10, fig.height=8}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(patchwork)

set.seed(3)
sleep = 3 + 40*rgamma(100, 2, 50)
gpa_sleep = 3.6 - 0.05*sleep + rnorm(100, sd=0.2)

p1 <- data.frame(sleep = sleep) %>%
  ggplot(aes(x = sleep)) +
  geom_histogram(bins = 10) +
  labs(x = "Hours of sleep per night") +
  theme_bw() +
  theme(text=element_text(size = 16))

p2 <- data.frame(sleep = sleep,
                 gpa_sleep = gpa_sleep) %>%
  ggplot(aes(x = sleep, y = gpa_sleep)) +
  geom_point() +
  geom_smooth(se=F, method="lm") +
  labs(x = "Hours of sleep per night",
       y = "GPA") +
  theme_bw() +
  theme(text=element_text(size = 16))

location <- rep(c("Dorm", "Food Court", "Library", "Office Hours"),
                each = 25)


gpa_location = rnorm(100, sd = 0.2) + rep(c(2.9, 3, 3.2, 3.4),
                                          each = 25)

p3 <- data.frame(x = location, 
           y = gpa_location) %>%
  ggplot(aes(x = x , y = y)) +
  geom_boxplot() +
  labs(x = "Favorite homework location",
       y = "GPA") +
  theme_bw() +
  theme(text=element_text(size = 16))

(p1 + p2)/(p3 + plot_spacer())
```

\newpage

Your collaborator then writes the following report on their analysis:

``To test whether there is a relationship between amount of sleep and GPA, we carried out a hypothesis test for the hypotheses $H_0: \beta_1 = 0$ vs. $H_A: \beta_1 \neq 0$, where $\beta_1$ is the true slope for the linear regression model between sleep and GPA in the population. 

The p-value for the hypothesis test is 0.000013. This means that there is only a 0.000013 probability that there is no relationship between sleep and GPA, so we have strong evidence for a relationship. The estimated slope is $-0.088$ and the estimated intercept is 3.79. This means that an additional hour of sleep is associated with a GPA of 3.79, on average. Unfortunately, the normality assumption for this linear model is not appropriate because the distribution of hours of sleep is right-skewed, as we can see in the histogram.

Finally, from the boxplots we can see that there is a positive, linear relationship between students' favorite place to do homework, and their GPA.

Based on these plots and the fitted model, we recommend that students who want to improve their GPA set their alarms for one hour earlier, and avoid doing any homework in their dorm rooms."

6. There are no errors in your collaborator's plots, and they correctly calculated the estimated slope (-0.088) and intercept (3.79) of the fitted regression line. The reported p-value (0.000013) is also correct. Unfortunately, they made several errors in their interpretation of the figures, estimated regression line, and/or p-value. List these errors below, and explain to your collaborator why each of their errors is wrong.