---
title: "Class Activity, February 23"
output: 
  tufte::tufte_html:
    css: "lab.css"
    tufte_variant: "envisioned"
    highlight: pygments
link-citations: yes
---

## Setup

In this activity, you will explore polynomial regression with data on CO2 emissions in Germany. 

The data is contained in the `CO2Germany` dataset in the `Stat2Data` package. You can load the data by running the following:

```r
library(Stat2Data)
data("CO2Germany")
```

The `CO2Germany` contains two variables: `CO2` (the CO2 level in ppm) and `Day` (the number of day in 2011).

Download the activity template: [activity template](https://sta112-s22.github.io/class_activities/ca_lecture_15_template.Rmd)

## Question

Here's a plot showing the relationship between Day and CO2:

```{r message=F, warning=F, echo=F, fig.align='center', fig.width=7, fig.height=5}
library(tidyverse)
library(Stat2Data)
data("CO2Germany")
CO2Germany %>%
  ggplot(aes(x = Day, y = CO2)) +
  geom_point(size = 2.5) +
  theme_bw() +
  theme(text = element_text(size = 25))
```

1. Fit a polynomial regression of order 2, with Day as the predictor and CO2 as the response.

2. What is the equation of the fitted model from exercise 1?

3. Based on the fitted model, is the relationship between Day and CO2 concave up or concave down?

4. What is the predicted CO2 level for day 200?

The figure below shows our fitted model (the blue curve) on the original scatterplot. We also try a polynomial of order 3 (the red dashed curve).

```{r message=F, warning=F, echo=F, fig.align='center', fig.width=7, fig.height=5}
library(tidyverse)
library(Stat2Data)
data("CO2Germany")
CO2Germany %>%
  ggplot(aes(x = Day, y = CO2)) +
  geom_point(size = 2.5) +
  geom_smooth(se=F, method="lm", formula = y ~ I(x) + I(x^2), lwd=1.2) +
  geom_smooth(se=F, method="lm", formula = y ~ I(x) + I(x^2) + I(x^3), lwd=1.2,
              color = "red", lty=2) +
  theme_bw() +
  theme(text = element_text(size = 25))
```

5. Does a polynomial of order 3 look like it fits better?