---
title: "Class Activity, March 18"
output: 
  tufte::tufte_html:
    css: "../labs/lab.css"
    tufte_variant: "envisioned"
    highlight: pygments
link-citations: yes
---

In this activity, you will explore an interaction between a quantitative predictor and a categorical predictor.

## Setup

Download the activity template: [activity template](https://sta112-s22.github.io/class_activities/ca_lecture_21_template.Rmd)

In this class activity, we will examine the relationship between Age and Height for children between the ages of 8 and 18. We are also interested in how that relationship differs for boys and girls. The `Kids198` dataset, part of the `Stat2Data` package, includes the following variables on 198 children:

* `Height`: Height (in inches)
* `Age`: Age (in months)
* `Sex`: 0=male, 1 = female

## Questions

The original `Kids198` data codes Sex as 0 or 1. Because these are numbers, R will treat Sex as a quantitative, rather than categorical, variable. We therefore need to modify Sex to be categorical before we fit the model.

1. Run the following code to re-code Sex as "male" or "female":

```{r eval=F}
Kids198 <- Kids198 %>%
  mutate(Sex = ifelse(Sex == 0, "male", "female"))
```

2. Fit a multiple linear regression model with Height as the response, Age and Sex as predictors, and an interaction between Age and Sex:

```{r eval=F}
m1 <- lm(Height ~ Sex*Age, data = Kids198)
summary(m1)
```

3. What is the estimated height for a girl aged 120 months (10 years)?

Now let's investigate whether an interaction seems important for this data. We will look at a plot, and also compare $R^2_{adj}$.

4. Run the following code to create a plot showing the relationship between Age and Height, with a separate line for each Sex:

```{r eval=F}
Kids198 %>%
  ggplot(aes(x = Age, y = Height, color = Sex)) +
  geom_point() +
  geom_smooth(se=F, method="lm", lwd=2) +
  theme_bw()
```

5. Fit a multiple linear regression model with Height as the response, and Age and Sex as predictors, but no interaction between Age and Sex. Compare the $R^2_{adj}$ from this new model to the $R^2_{adj}$ from your model in exercise 2.

6. Based on exercises 4 and 5, do you think an interaction is needed?
